### 2015-09-21 paper.tex
Initial Outline

\documentclass[11pt]{article}

\usepackage{deauthor,times,graphicx}

\begin{document}
\title{Apache Flink: Stream and Batch Processing at Scale}
\author{Authors}

\maketitle

\begin{abstract}
Abstract TBD.
\end{abstract}

\section{Introduction}
\subsection{State of the Art}

\section{System Architecture}
	\subsection{Overview}
	\subsection{Execution Model}
	\subsection{APIs and Examples}

\section{Data Processing with Flink}
	\subsection{Streaming Execution Engine}
		\subsubsection{Pipelining}
		\subsubsection{Windows}
		\subsubsection{Fault Tolerance}

	\subsection{Batch as a Special Case of Streaming}
		\subsubsection{Out-of-Core Operator Execution}
		\subsubsection{Program Optimization}
		\subsubsection{Native Iterations}

\section{Discussion and Outlook}
\section{Experiments}

\cite{DBLP:conf/hotcloud/ZahariaCFSS10}

\bibliographystyle{abbrv}
\bibliography{references}

\end{document}

### 2015-11-05 paper.tex
Added figures

@@
-\documentclass[11pt]{article}
+\documentclass{sig-alternate}
 
@@
 \begin{document}
+
 \title{Apache Flink: Stream and Batch Processing at Scale}
@@
 
-\section{Introduction}
-\subsection{State of the Art}
+Apache Flink™ : Streaming dataflows as a basis for universal data analytics
 
-\section{System Architecture}
-	\subsection{Overview}
-	\subsection{Execution Model}
-	\subsection{APIs and Examples}
+Abstract
 
-\section{Data Processing with Flink}
-	\subsection{Streaming Execution Engine}
-		\subsubsection{Pipelining}
-		\subsubsection{Windows}
-		\subsubsection{Fault Tolerance}
+Motivation
+Organizations rely more and more on the value of their data, which is directly correlated with the freshness of the data.
+Additionally, as the data infrastructure becomes more and more complex, development time increasingly goes into plumbing and integrating systems together rather than focusing on the business problem.
+These trends, together with the ever-growing need for lower latency (“real-time”), and scale create a pressing need for more real-time, asynchronous, push-based data infrastructure.
+This kind of infrastructure “turns the database around”, by allowing isolated applications to interact with their local datasets, rather than imposing the need to coordinate on a global data store.
+In conjunction with the growing trends of cloud-based data processing, containerization, Internet of Things, and microservices creates a new reality for the data infrastructure.
+Popularized by internet companies such as LinkedIn, this trend is quickly sweeping through the enterprise.
 
-	\subsection{Batch as a Special Case of Streaming}
-		\subsubsection{Out-of-Core Operator Execution}
-		\subsubsection{Program Optimization}
-		\subsubsection{Native Iterations}
+Collectively referred to as streaming, this new kind of infrastructure is based around the following notions.
+Data items (records) represent real-world events (e.g., a tap on a smartphone, a sensor reading, or a database transaction).
+An event stream is a time-ordered series of records together with the timestamp that they “happened”, measured, e.g., by the smartphone itself.
+An event stream can be thought of as a log, in which new events can only be added at the tail, and events can only be consumed from the head.
+To process historical data, the log needs to be replayed from a past logical time.
+Applications can consume event streams and create and maintain their own views.
+Contrast this paradigm with the traditional database-middleware paradigm.
+With streaming there is no global consistent view of the world that applications need to interact with.
+Rather, the history of the world is the ground truth, and isolated applications maintain their own views, often referred to as state.
 
-\section{Discussion and Outlook}
-\section{Experiments}
+Of course, none of this is really new.
+A large part of the enterprise infrastructure has always been based on continuously flowing event streams, typically served by tools such as enterprise message buses (e.g., TIBCO), CEP (Complex Event Processing) systems (e.g., IBM Infosphere Streams, Microsoft Streaminsight, Streambase), or roll-your-own internal systems.
+CEP systems in particular are based on years of research in the database community with projects such as Aurora, Telegraph, STREAMS, [[Microsoft project]], etc.
+While these systems had always served a niche market well, they never achieved widespread adoption in the industry for a variety of reasons, including limited scalability, cost, and lack of integration with other parts of the data infrastructure.
+Another reason for the limited adoption is, of course, timing.
+The “Big Data” movement has created an explosion of data, most of which is of a continuous, unbounded nature.
+
+Another way that industry is tackling this problem is by using a traditional batch processing paradigm to serve streaming use cases.
+This has led to approaches such as “micro-batching” [], where a batch processor is used to simulate continuous processing, and the “lambda architecture”, where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
+
+To summarize, there is a pressing need for streaming infrastructure that scales in order to tackle both integration and latency.
+The historical lack of good implementations of such systems has led to one-off solutions, or bolt-on solutions with questionable performance.
+What modern enterprises need is a streaming platform that is (1) consistent, (2) general, (3) efficient, and (4) scalable.
+
+Apache Flink is a platform that delivers on these requirements.
+Flink is an open source project licensed by the Apache Software Foundation (ASF), a non-profit organization that houses open source projects and their communities.
+The Flink community is counting more than 120 developers, and growing fast [].
+Historically, Flink stems from the database research community, and in particular the Stratosphere research project [].
+Flink entered the ASF in April 2014 as an “incubating” project, and “graduated” as a “top-level” project in December 2014.
+
+Flink incorporates several novel research contributions, some of which have been published in the past, and some of which have only been published in the form of blog posts or wiki pages.
+In this article, we attempt a complete description of all novel aspects of the system, including:
+A novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results.
+A novel mechanism to provide fault tolerance in a distributed data streaming system while keeping the latency low and the throughput high.
+How to overcome limitations of the JVM by careful engineering
+Efficiently embedding iterations within a DAG-based dataflow system [vldb12]
+Extending cost-based optimization to user-defined functions [socc10, vldb12]
+The value of this paper lies mostly into positioning all these within a common architecture and few conceptual building blocks.
+We believe that Flink is the artifact of “Big Data systems done right”, clearly superseding most other proposed architectures in the space in all fronts (including batch, stream, iterative processing).
+System architecture
+We begin by laying out the architecture of Flink (i) as a software stack, and (ii) as a distributed system.
+
+While Flink’s stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
+
+The core of Flink is the distributed dataflow engine, which executes dataflow programs called JobGraphs.
+A JobGraph is a DAG of operators and connections between operators, but it is oblivious to the semantics of these operators: for example, the engine does not distinguish between a join or a coGroup, and it similarly does not distinguish between a map or a stream window operator; semantics are left to the upper layers, which embed the specific operator logic withing the boxes provided by the runtime engine.
+
+There are two “core” APIs in Flink: the DataSet API for processing finite data sets (often referred to as “batch processing”), and the DataStream API for processing potentially unbounded data sets (often referred to as “stream processing”).
+We note that the the terms “batch” and “stream processing” here only cause confusion, as the carry the assumption that the underlying engine uses a particular style of physical data processing [Dataflow, Streaming 101].
+In Flink, the core runtime engine can be thought of as a streaming dataflow engine, and both APIs create programs (JobGraphs) accepted by this engine.
+The reason for the existence of the two APIs is largely historical.
+The DataSet API pre-existed to the DataStream API.
+In theory, one can treat finite data sets as a special case of unbounded data streams, and embed all functionality of the DataSet API within the DataStream API.
+On top of the core APIs, Flink bundles “domain-specific” libraries and APIs, currently:
+Gelly, an API and library for processing graphs
+FlinkML, an API and library for composing Machine Learning pipelines
+Table, an API similar in spirit to Microsoft’s LINQ
+All these APIs generate DataSet and DataStream API programs.
+In addition, Flink includes several compatibility packages, currently a Hadoop MapReduce and an Apache Storm compatibility package.
+Finally, third-party tools can also generate Flink programs.
+These are not shipped with the Flink codebase, but are part of other software projects.
+A non-exhaustive list of these are Cascading [], Google Cloud Dataflow [], Apache SAMOA [], Apache Mahout [], and Apache Zeppelin [].
+
+A Flink cluster involves three main types of processes: the client, the JobManager, and the TaskManager.
+When Flink is running in a YARN cluster, the JobManager and TaskManager processes are embedded within YARN containers.
+A typical cluster consists of one client (typically living in the machine that submits the program), one or more JobManagers (one leader and one or more replicas for high availability), and several TaskManagers, typically one per cluster node.
+
+[[Explain program compilation in detail, what JobManager does and main JM data structures, what TaskManagers do, and main TaM data structures, how messages are shipped, how data is transferred, and how heartbeats and high availability work]].
+The core: continuous dataflows
+
+Programmers interact with Flink using a combination of its APIs (currently DataSet, DataStream, Table, Graph, and Machine Learning).
+Every Flink program goes through a compilation phase (a combination of type inference, query optimization, and code generation, all described in the subsequent sections), and creates a dataflow graph (called JobGraph in Flink parlance) that serves as a unifying representation of all Flink programs.
+
+A JobGraph is a directed acyclic graph (DAG) that consists of nodes and edges.
+There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
+For example, the following graph consists of five operators (o1-o5), and three intermediate results (ir1-ir3).
+Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g., reading data from a file system, a socket, a message queue, etc), and data sinks.
+Operators produce intermediate results, as well as updates to state (which can be managed externally).
+An intermediate result is a logical handle (pointer) to the data that is produced by one operator.
+An intermediate result can be consumed by one or more operators.
+Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
+
+When the JobGraph is scheduled in a cluster for execution (Flink can run on standalone and YARN clusters), it is parallelized to form an ExecutionGraph.
+An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
+Let us take the above JobGraph and partition it using a parallelism of 2.
+Assume that o1 and o2 are data sources, o2 is a map operator, o4 is a join operator that needs both inputs partitioned by the same key, and o5 is a filter.
+Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
+The resulting ExecutionGraph is the following:
+
+The unit of data transfer in the Flink runtime is a buffer.
+Buffers contain one or more records, and a record can span multiple buffers.
+The default size of a buffer in Flink is 32 kilobytes.
+Buffers are requested and relinquished from local buffer pools, shared among operators that live in the same task manager.
+An IRP is simply a collection of buffers.
+Work in Flink progresses (i.e., records flow through the pipeline) as long as there are buffers available, essentially implementing distributed blocking queues (the logical streams) with bounded capacity (the amount of memory available to the buffer pools, which can be configured by the user).
+Flink  This mechanism, in addition to implementing record network transfers doubles down as a natural way to backpressure the flow in the case of slow operators (including external systems that consume data).
+See [http://data-artisans.com/how-flink-handles-backpressure/] for a detailed presentation.
+
+We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
+Internally, and intermediate result is an abstract class that has many implementations.
+Let us look at two concrete examples: implementing pipelined data exchange, and implementing blocking data exchange.
+
+Pipelined data exchange (also called intra-operator parallelism) means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
+Pipelining is present (and required) in streaming systems (such as Flink, Storm, CEP systems, etc), and is also used in MPP databases and DBMSs (Impala, Greenplum), but it is typically not used in batch processors (Spark, MapReduce, Tez).
+Pipelined data exchange is implemented in Flink by an intermediate result implementation, which activates network transfers as soon as the first buffer is available.
+In particular, as soon as the producing operator places the first buffer in the cache of the result partition, the IR signals to the JobManager its availability.
+The JobManager then signals to all consumers of this partition that the the partition is ready (scheduling the operators if they had not been previously scheduled), and initiates the data transfer.
+
+Pipelined data exchange is great for lower latency (time to first result) in batch jobs, and is required in streaming jobs.
+To better allocate resources in very large batch jobs, it is sometimes desirable to break up a job into stages, scheduling and executing each stage individually.
+To do that, a blocking data exchange that materializes intermediate results in a combination of memory and disk is desirable.
+Flink implements blocking data exchange via an intermediate result that signals its availability only when all the buffers from the producer have been cached.
+The cached buffers double down as a materialized reusable intermediate result, similar to Spark’s RDD data structure.
+Blocking data exchange is used in Flink to break deadlocks, or when specified by the user.
+All other data processing needs (including batch processing) currently use pipelined data exchange.
+
+All available implementations of intermediate results in Flink are shown in the following table:
+
+Pipelined
+Materialized
+Ephemeral
+
+Cached
+
+See [https://cwiki.apache.org/confluence/display/FLINK/Data+exchange+between+tasks] for a detailed presentation.
+
+Fault tolerance via asynchronous snapshots
+Backing up and restoring the state of a computation is vital in the presence of software or infrastructure failures.
+On the event of such a failure, the system has to recover the state of the failed operator, and continue the computation from where it left off.
+As in transaction execution, different applications require different consistency guarantees.
+For instance, if no consistency guarantee is required (e.g., a live monitoring dashboard that displays clicks in a website), a system might simply recover the operator state to any previous snapshot and continue ingesting the stream.
+However, a computation that requires correctness of results (e.g., user interactions with their shopping basket) requires stricter consistency guarantees.
+
+Such consistency guarantees include:
+at most once: no consistency guarantees, including possibility for data loss 
+at least once: the fault tolerance mechanism guarantees that an event is going to touch the state of an operator at least once i.e., in case of failure, some events of the stream might touch the state multiple times.
+exactly once: that the operator state will be touched exactly once per event, no matter how many failures may happen.
+
+In batch processing, guaranteeing exactly once consistency is simple: the system simply pauses the computation, re-deploys (part of) the dataflow and re-consumes the necessary parts of its input.
+In the streaming case, however, i) computations are continuous (e.g., months of processing) ii) a data stream does not have, in principle, a beginning or an end, nor can it be stored indefinitely and iii) the computation cannot simply be stopped mid-flight.
+Exactly-Once Fault Tolerance in Flink
+The problem of providing exactly once guarantees boils down to determining what state the streaming computation currently is in (including in-flight records, and operator state), drawing a consistent snapshot of that state, and storing that snapshot in durable storage.
+If one can do this frequently, recovery from a failure only means i) restoring the latest snapshot from durable storage, and ii) rewinding the stream source to the point when the snapshot was taken and continuing the computation.
+Flink’s algorithm is described in detail in []; in the sequel, we give a brief summary.
+
+Flink’s snapshot mechanism is based on an improvement of a technique introduced in [Chandy/Lamport], to draw consistent snapshots of the current state of a distributed system without missing information and without recording duplicates.
+In this technique, the snapshots are taken asynchronously: the computation does not need to stop in order to take a snapshot.
+Flink’s fault tolerance mechanism draws state snapshots of a running stream topology, and stores these snapshots to durable storage (e.g., to HDFS or an in-memory file system).
+The frequency of these checkpoints can vary according to the failure rates and the cost of taking a snapshot.
+
+The checkpointing mechanism is based on stream barriers (“markers” in []) that flow through the operators.
+Barriers are first injected at the sources (e.g., if using Apache Kafka as a source, barriers are aligned with offsets), and flow through the DAG as part of the data stream together with the data records (see figure below).
+A barrier separates records into two groups: those that are part of the current snapshot (a barrier signals the start of a checkpoint), and those that are part of the next snapshot.
+
+Barriers flow downstream and trigger state snapshots when they pass through operators.
+An operator first aligns its barriers from all incoming streams (if the operator has more than one input), buffering data from the faster ones.
+When an operator has received a barrier from all incoming streams, it checkpoints its state (if any) to durable storage.
+Once state checkpointing is done, the operator forwards the barrier downstream.
+Note that in this mechanism, state checkpointing can be both asynchronous (processing continues while the state is written out to durable storage) and incremental (only deltas are written to durable storage).
+Once all data sinks have released the barriers, the current checkpoint has finished.
+Recovery on a failure means simply restoring the latest checkpointed state, and restarting the sources from the last recorded barrier.
+
+The aforementioned mechanism guarantees “exactly-once” state updates, meaning that any state that takes part in the snapshotting mechanism will have the same value with or without failures.
+Flink also includes a mode for at-most-once processing (no snapshotting, no barriers), as well as a mode for at least once processing (operators simply forward the barriers, but sources snapshot their offsets).
+
+A common misinterpretation of the term “exactly-once” is that the output of a streaming job does not have any duplicates.
+That might still be the case, as data sinks are external systems, which Flink might not have access to.
+However, by building custom connectors to common systems that participate in Flink’s snapshotting  mechanism, one can ensure end-to-end exactly once guarantees.
+Flink bundles a number of these connectors.
+For example, Flink bundles a data sink to any HDFS-compatible file system that creates rolling files from a continuous streaming job.
+The user can specify when a new file will be created.
+In Flink data sinks are stateful operators just as any other operator.
+The state of the HDFS sink is the last offset written in the file.
+In the case of a failure, restoring the sink to its snapshotted state triggers a file truncation that deletes the data that was written between the last snapshot and the failure.
+
+Thus, Flink can be seen as a way to implement distributed transactions between systems.
+By creating sources and sinks that take part in the snapshotting, and relaying the data movement to Flink, one can move data consistently with transactional guarantees between systems.
+For example, the HDFS sink describes above implements “read uncommitted” semantics, as “garbage” data may be written and then deleted.
+We note that such systems implemented with Flink are CA systems in the CAP theorem characterization.
+Stream analytics on top of dataflows
+Window implementation
+Batch analytics on top of dataflows
+How to embed blocking operators in dataflow graph
+Why this is not enough
+Blocking data exchange (stages)
+Scheduling (WiP)
+Fault tolerance (WiP)
+Query optimization
+Memory management
+
+How Flink implements those
+Query optimization
+Motivation?
+
+Flink’s optimizer builds on technology from parallel database systems; namely, on the concepts of plan equivalences, cost models and interesting properties.
+Yet, certain issues which  stem from the UDF-heavy DAGs that represent Flink’s dataflow programs, do not allow Flink’s optimizer to employ those database techniques out of the box [black-boxes].
+In the sequel, we summarize these issues and present the solutions that Flink implements.
+
+The first issue is that queries in most database systems are represented and executed as trees of operators (i.e., each operator has exactly one output).
+whose semantics are well defined  and known to the optimizer.
+Based on those semantics, traditional optimizers enumerate alternative logical plans by e.g., employing equivalence rules for operator reordering [volcano, cascades].
+In contrast, Flink’s programs do not form trees of known operators but DAGs, which encapsulate UDFs with arbitrary user code.
+As a result, the operators hide their semantics from the optimizer and traditional techniques for plan enumerator are not directly applicable.
+Worse, traditional cardinality and cost estimation methods do not apply either, due to the presence of UDFs.
+Flink’s optimizer employs a number of novel methods for overcoming these issues [blackboxes, stratosphere-journal, iterations] for which we provide an short overview below.
+Plan Enumeration for UDF-heavy DAGs
+Plan enumeration (reordering) in the presence of UDFs. Semantics of operators. etc.
+
+Physical Optimization
+Flink’s runtime supports execution strategies known from parallel databases, such as repartition/broadcast data transfer strategies, as well as sort-based grouping and different join implementations (sort- and hash-based).
+Flink’s optimizer enumerates different physical plans based on the concept of interesting properties propagation [scope-optimizer].
+
+The optimizer uses a cost-based approach to choose among multiple equivalent physical plans, based on network/disk I/O, and CPU cost.
+For the optimizer to assign costs to plans, it needs to first estimate the size of intermediate results.
+Estimating the cardinality of UDFs’ outputs is very challenging, even in relational databases [], and physical properties can be destroyed by a UDF (e.g., by modifying the field on which the input of an operator is sorted).
+To overcome these issues,  Flink’s optimizer relies on hints that are provided by the programmer.
+Examples of hints can be to state which side of a join is very small (e.g., could lead to the execution of a broadcast join) or that a certain field of a record, entering a UDF, is forwarded intact (i.e., physical properties such as sorting on that field are preserved and can be propagated to further parts of the plan).
+These hints enrich the internal DAG representation of a Flink program with extra information and allow the optimizer to perform propagation of interesting physical properties in the DAG.
+
+Previous work on constructing optimal plans for DAG-structured queries have shown that it is a very challenging problem [neumann-moerkete].
+Flink performs the propagation of interesting properties (sorting, partitioning, etc.) from the sources to the sinks and then performs a pass from the sinks to the sources in order to prune the possible plans and choose the optimal one, similarly to [scope-optimizer].
+To correctly enumerate DAG-structured plans, Flink first analyses the DAG in order to find the operators that “branch” (i.e., have more than one outputs) and the operators that “join” those branches back together.
+The subplans rooted on the branching operator are treated like common subexpressions;  The plan candidates for that operator’s inputs must have the same subplan for the common subexpression.
+Finally, the optimizer adds artificial pipeline breakers to plans with deadlocks.
+A deadlock can happen when some of the paths between the branching and the joining operator are fully pipelined (i.e., pipelined operators, depend - possibly indirectly -  on each other).
+
+Iteration-specific optimizations here or in the iterations section?
+Memory Management
+In data analysis frameworks such as Flink, data needs to be stored in memory, for data caching and executing operations such as sorting and joining.
+A straightforward approach of processing in a JVM, is to store the data as Java objects, and operate on those, leaving the responsibility of garbage collection and spilling to disk to the JVM and the OS respectively.
+In that case, the overhead of garbage collection in a JVM heap filled with millions of objects, can easily exceed the actual computation time.
+Moreover, Java objects require a certain space overhead, depending on the JVM implementation and platform.
+Database operator implementations have been traditionally managing memory and employing smart memory spilling mechanisms [].
+
+Building on database technology, Flink, instead of storing objects in the heap, serializes objects into a fixed number of pre-allocated memory segments.
+Figure X, depicts the structure of a memory segment and the internal representation of a tuple of 3 fields: an integer, a double and a nested Pojo (a Person that has an ID and a name).
+
+In Flink, operations such as sorting, and joining operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum.
+If more data needs to be processed than can be kept in memory, Flink’s operators partially spill data to disk.
+Flink builds on, and generalizes database technology to handle arbitrary objects using type inference, and  custom serialization and comparison mechanisms.
+
+By employing such techniques for data processing off-heap and in binary representation, Flink manages to, not only reduce the garbage collection overhead, but also to perform memory-safe execution and implement cache-efficient and robust out-of-core algorithms.
+Iterative analytics on top of dataflows
+Supersteps or asynchronous
+Head and tail
+Delta iterations
+Backup of in-flight records in snapshotting
+
+Related work
+Fault Tolerance in Streaming Systems
+In stream processing, fault tolerance with exactly-once guarantees implemented in many different ways.
+For instance, Google Dataflow/Millwheel does it by storing, in a transactional store, the in-flight records (the ones that it pushes downstream) together with the current state, while in systems that implement micro-batching (e.g., Spark, Trident), the resulting state at each micro-batch step is checkpointed (as described above for the batch case) and can be easily recovered in case of failures.
+
+Conclusion
+
+References
 

### 2015-11-07 paper.tex
Text port from Google docs

--- 2015-09-21 paper.tex
@@
-\documentclass[11pt]{article}
+\documentclass{sig-alternate}
+
+% A hack to make deauthor not complain which redefines
+% the titlepage command
+\newcommand{\titlepage}{}
 
@@
 
+% trademark symbol
+\usepackage{textcomp}
+
 \begin{document}
-\title{Apache Flink: Stream and Batch Processing at Scale}
+
+\title{Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics}
 \author{Authors}
@@
 \section{Introduction}
-\subsection{State of the Art}
 
-\section{System Architecture}
-	\subsection{Overview}
-	\subsection{Execution Model}
-	\subsection{APIs and Examples}
+Organizations rely more and more on the value of their data, which is directly correlated with its freshness.
+In addition, as the data infrastructure becomes more and more complex, development time is focused on integrating different systems together, rather than focusing on business logic.
+The ever-growing need for lower latency analytics and scale, creates a pressing need for more real-time, asynchronous, data processing infrastructure.
 
-\section{Data Processing with Flink}
-	\subsection{Streaming Execution Engine}
-		\subsubsection{Pipelining}
-		\subsubsection{Windows}
-		\subsubsection{Fault Tolerance}
+Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc.
+These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc.
+However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc.
+These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.
+Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases.
+This has led to approaches such a ``micro-batching'' [], where a batch processor is used to simulate continuous processing, and the ``lambda architecture'', where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
 
-	\subsection{Batch as a Special Case of Streaming}
-		\subsubsection{Out-of-Core Operator Execution}
-		\subsubsection{Program Optimization}
-		\subsubsection{Native Iterations}
+We believe that there is still a need for a real-time analytics platform, that is not covered by the existing solutions described above.
+Our observation is that modern enterprises need a platform that integrates well in the Hadoop-related Big Data processing zoo, supports user-defined computations, while being able to perform efficient, scalable and consistent data analysis, on top of both streaming/unbounded and stored/batch data.
 
-\section{Discussion and Outlook}
-\section{Experiments}
+Apache Flink is a platform that was designed to deliver on these requirements.
+It is an open source project licensed by the Apache Software Foundation (ASF), with its community counting more than 120 developers, and growing fast [].
+Flink originates from the database research community, in particular the Stratosphere research project [], and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
+In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.
 
-\cite{DBLP:conf/hotcloud/ZahariaCFSS10}
+This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results.
+Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high.
+Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions [socc10, vldb12] and efficiently embed iterations within a DAG-based dataflow system [vldb12].
+
+\section{System architecture}
+
+In this section we lay out the architecture of Flink as a software stack, and as a distributed system.
+While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Stack figure comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% This is the sutext of the figure
+\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
+A Job Graph is a DAG of operators and connections between operators.
+There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing'').
+Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine.
+On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs.
+Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.
+
+A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager.
+The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager.
+The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
+DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).
+
+The JobManager is Flink's master node.
+It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers, receiving statistics, controls the tasks' lifecycle,  coordinates the fault tolerance machanisms.
+The actual data processing takes place in the TaskManagers, or worker processes that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
+TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
+
+\section{Execution model}
+
+\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
+There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
+For example, the graph below consists of five operators ($o_1$-$o_5$), and three intermediate results ($ir_1$-$ir_3$).
+Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
+reading data from a file system, a socket, a message queue, etc.), and data sinks.
+Operators produce intermediate results, as well as updates to state.
+An intermediate result is a logical handle (pointer) to the data that is produced by one operator.
+An intermediate result can be consumed by one or more operators.
+Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% JobGraph figure comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph.
+An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
+The above JobGraph, if it is partitioned using a parallelism of two, it results to the execution graph below.
+Assume that $o_1$ and $o_2$ are data sources, $o_2$ is a map operator, $o_4$ is a join operator that needs both inputs partitioned by the same key, and $o_5$ is a filter.
+Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
+The resulting ExecutionGraph is the following:
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% ExecutionGraph figure comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\textbf{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
+Buffers contain one or more records, and a record can span multiple buffers.
+Buffers are requested and relinquished from local buffer pools, shared among operators that live in the same task manager.
+An IRP is simply a collection of buffers.
+Work in Flink progresses (i.e., records flow through the pipeline) as long as there are buffers available, essentially implementing distributed blocking queues (the logical streams) with bounded capacity (the amount of memory available to the buffer pools, which can be configured by the user).
+This mechanism, in addition to implementing record network transfers doubles down as a natural way to backpressure the flow in the case of slow operators (including external systems that consume data).
+
+We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
+Internally, intermediate results are abstract classes with many implementations.
+These implementations can perform pipelined data exchange, and blocking data exchange.
+
+\textbf{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
+Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
+Flink implements pipelining by implementing intermediate results which activate network transfers between the producer task and consumer task,  as soon as their first buffer is available.
+Flink allows the configuration of its buffers' size and timeout.
+A buffer is sent to its consumer task as soon it is filled,  or as soon as a timeout is reached.
+Hence, by setting the buffer size and/or timeout to lower values, one can reduce latency, while achieving the opposite effect (i.e., high throughput) by using higher values.
+
+\textbf{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
+To do that, the system has to materialize intermediate results (in memory or disk).
+Flink implements blocking data exchange via an intermediate result that signals its availability only when all the buffers from the producer have been materialized.
+The cached buffers can double down as a materialized reusable intermediate result.
+
+\section{Fault Tolerance via \\ Asynchronous Snapshots}
+
+Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals.
+When failure occurs the state of the execution is restored from the latest snapshot.
+Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source.
+Normally this possible by using persistent message queueing systems, e.g.
+Kafka.
+
+Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS) [arXiv paper].
+It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
+It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
+Similar to Chandy-Lamport distributed snapshots [Chandy/Lamport], the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
+ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
+Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
+Furthermore, ABS [arXiv paper] supports snapshotting of cyclic graphs.
+The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
+The following figure illustrates the snapshot process.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% ABS figure comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Stream analytics on top of \\ dataflows}
+
+Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
+The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
+Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams.
+Each of the two supports specific operations since they exhibit different properties.
+For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations.
+Other transformations, such as (flat)map are  can be applied incrementally on data streams.
+
+% Give an example here instead of simplicity
+The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty.
+Here is, for example, a word count program on a simple  time-based window:
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% WindowWC snippet comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Flink distinguishes between two notions of time: 
+\begin{description}
+\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
+\item[Processing time] is the wall-clock time of the machine that is processing the data.
+\end{description}
+
+In distributed systems, there is an arbitrary lag between event-time and processing-time [dataflow].
+This may mean arbitrary delays for getting an answer based on event-time semantics.
+To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
+The watermarks aid the execution engine to process events in the correct event order.
+
+Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency.
+Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
+Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system.
+This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.
+
+\textbf{Streaming Windows.} A window defines a logical group of records that are processed together.
+Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google.
+A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
+The assigner predefines the logical groups to which each record belongs, e.g.
+count-based or time-based periodic windows.
+The trigger defines when the window operation is performed on each group.
+The evictor defines which records to keep in each group.
+The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner).
+The window results are computed once the watermark passes the end of the window (Trigger).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Trigger snippet comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+A short hand notation for the example above is:
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Shorthand snippet comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+A global window creates a single logical group.
+The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Global window snippet comes here
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
+This mechanism can be used to implement a wide variety of windowing functionality [Dataflow], and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
+
+Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
+We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.
+
+\section{Batch analytics on top of \\ dataflows}
+
+A bounded data set is a special case of an unbounded data stream.
+Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above.
+However, i) the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops) and ii) we can simplify the fault tolerance mechanisms and iii) we can apply query optimization borrowing ideas from MPP database systems.
+
+For these reasons, Flink treats specially batch computations, by implementing the above optimizations.
+For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime.
+These operators simply block until they have received all of their input.
+Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad [Dryad].
+
+With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
+
+\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
+However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
+For the same reason, cardinality and cost estimation methods are equally difficult to employ.
+Flink's optimizer employs a number of novel methods for overcoming these issues [blackboxes, stratosphere-journal, iterations] for which we provide a short overview below.
+Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
+Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation [scope-optimizer], using a cost-based approach to choose among multiple physical plans.
+The cost includes network/disk I/O and CPU cost.
+To overcome the cardinality estimation issues that were mentioned earlier,  Flink's optimizer uses hints that are provided by the programmer.
+
+\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
+These memory segments resemble database blocks into which, java objects representing the tuples that go through the runtime are serialized.
+Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
+To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.
+By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.
+
+\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
+Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG [Hadoop iterative works, Spark], hiding from  the engine that it is executing an iterative program.
+The approach, implemented in Naiad [] adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
+A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration [].
+
+Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
+This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines [], and outperforms the driver-based approach [].
+Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
+Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
+
+\section{Related work}
+
+There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
+
+\textbf{Batch Processing.} Apache Hadoop [] is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm [].
+Dryad [], introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE [] which added a language and an SQL optimizer on top of it.
+Apache Tez [] can be seen as an open source implementation of the ideas proposed in Dryad.
+MPP databases [], and recent open-source implementations [Apache Drill, Impala] restrict their API to SQL variants.
+Very similar to Flink, Apache Spark [] is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+In contrast, Flink is the only system that i) sports an optimizer that can optimize DAG programs beyond SQL queries, ii) is able to perform very efficient iterative processing natively, iii)  includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
+
+\textbf{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
+Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis [], Stanford STREAMS [], Trill [], and IBM System S [].
+Most of the above systems are either i) academic prototypes, ii) closed-source commercial products, or iii) do not scale the computation horizontally on clusters of commodity servers.
+Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide very low level APIs and offer only at-least-once and at-most-once guarantees.
+MillWheel [], a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow [] to implement  out-of-order stream analytics.
+To the best of our knowledge, Flink is the only open-source project that i) offers high level APIs and processing libraries, ii) provides state management with exactly-once guarantees and iii) achieves high throughput and low latency, serving both batch and true streaming computations equally well.
+
+\section{Conclusion}
+
+In this paper we presented Apache Flink, a platform that implements one universal dataflow engine designed to perform both stream and batch analytics.
+Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by the batch and a data stream APIs.
+The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state, and to partition, transform, discretize and aggregate a data stream.
+While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
+
+\section{References}
 

### 2015-11-07 paper.tex
Added some figs and listings

@@
 
+% JobGraph and Execution figs
+\usepackage{tikz}
+\usetikzlibrary{arrows,automata,calc,positioning,shapes.geometric}
+
 % trademark symbol
 \usepackage{textcomp}
+
+% code listings
+\usepackage{listings}
+
+\definecolor{dkgreen}{rgb}{0,0.6,0}
+\definecolor{gray}{rgb}{0.5,0.5,0.5}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
+
+% there is no built in support or Scala yet, good enough
+\lstset{frame=tb,
+  language=Java,
+  aboveskip=3mm,
+  belowskip=3mm,
+  showstringspaces=false,
+  columns=flexible,
+  basicstyle={\scriptsize\ttfamily},
+  numbers=none,
+  numberstyle=\tiny\color{black},
+  keywordstyle=\color{blue},
+  commentstyle=\color{dkgreen},
+  stringstyle=\color{mauve},
+  breaklines=true,
+  breakatwhitespace=true,
+  tabsize=4,
+  %for scala
+  emph={%  
+    object, def, val, zip%
+    },emphstyle={\color{blue}}%
+}%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Stack figure comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h!]
+	\centering  	
+  	\includegraphics[width=.45\textwidth]{figs/flink_stack.png}
+	\caption{The Flink software stack.}
+	\label{fig:FlinkStack}
+\end{figure}
 
-% This is the sutext of the figure
 \textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
@@
 There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
-For example, the graph below consists of five operators ($o_1$-$o_5$), and three intermediate results ($ir_1$-$ir_3$).
+For example, the graph below consists of five operators ($op_1$-$op_5$), and three intermediate results ($ir_1$-$ir_3$).
 Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% JobGraph figure comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
+\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
+\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
+
+\node[op] (op_1) {$op_1$};
+\node[ir] (ir_1) [right of = op_1] {$ir_1$};
+\node[op] (op_2) [right of = ir_1] {$op_2$};
+\node[ir] (ir_2) [right of = op_2] {$ir_2$};
+\node[op] (op_3) [below of = op_2] {$op_3$};
+\node[ir] (ir_3) [right of = op_3] {$ir_3$};
+\node[op] (op_4) [right of = ir_2] {$op_4$};
+\node[op] (op_5) [right of = ir_3] {$op_5$};
+
+\path[->] 	
+			(op_1) 	edge node [above] {} (ir_1)
+			(ir_1) 	edge node [above] {} (op_2)
+			(op_2) 	edge node [above] {} (ir_2)
+			(ir_2) 	edge node [above] {} (op_4)
+			(op_3) 	edge node [above] {} (ir_3)
+			(ir_3) 	edge node [above] {} (op_4)
+			(ir_3) 	edge node [above] {} (op_5)
+			;
+\end{tikzpicture}
+\label{fig:JobGraph}
+\caption{The JobGraph is the logical view of a Flink job.}
+\end{figure}
 
@@
 An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
-The above JobGraph, if it is partitioned using a parallelism of two, it results to the execution graph below.
-Assume that $o_1$ and $o_2$ are data sources, $o_2$ is a map operator, $o_4$ is a join operator that needs both inputs partitioned by the same key, and $o_5$ is a filter.
+The JobGraph in fig.
+\ref{fig:JobGraph}, if it is partitioned using a parallelism of two, it results to the ExecutionGraph shown in fig.
+\ref{fig:ExecutionGraph}.
+Assume that $op_1$ and $op_2$ are data sources, $op_2$ is a map operator, $op_4$ is a join operator that needs both inputs partitioned by the same key, and $op_5$ is a filter.
 Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ExecutionGraph figure comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+[shorten >=1pt,node distance=1.6cm,>=stealth',initial text=, font=\tiny]
+\tikzstyle{every state}=[circle, fill,draw=none,white,text=black]
+\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
+\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.2, draw, minimum size=0.8cm, text width=0.8cm, align=center]
+
+\node[op] (op_11) {$op_{11}$};
+\node[ir] (isp_111) [right of = op_11] {$isp_{111}$};
+\node[op] (op_21) [right of = isp_111] {$op_{21}$};
+\node[state] (ir_21) [right of = op_21] {};
+\node[ir] (isp_212) [below of = ir_21, node distance = 0.35cm] {$isp_{212}$};
+\node[ir] (isp_211) [above of = ir_21, node distance = 0.35cm] {$isp_{211}$};
+\node[op] (op_31) [below of = op_21] {$op_{31}$};
+\node[state] (ir_31) [right of = op_31] {};
+\node[ir] (isp_312) [below of = ir_31, node distance = 0.35cm] {$isp_{312}$};
+\node[ir] (isp_311) [above of = ir_31, node distance = 0.35cm] {$isp_{311}$};
+\node[op] (op_41) [right of = ir_21, node distance = 2.25cm] {$op_{41}$};
+\node[op] (op_51) [right of = ir_31, node distance = 2.25cm] {$op_{51}$};
+
+\node[op] (op_22) [below of = op_31] {$op_{22}$};
+\node[ir] (isp_121) [left of = op_22] {$isp_{121}$};
+\node[op] (op_12) [left of = isp_121] {$op_{12}$};
+\node[state] (ir_22) [right of = op_22] {};
+\node[ir] (isp_222) [below of = ir_22, node distance = 0.35cm] {$isp_{222}$};
+\node[ir] (isp_221) [above of = ir_22, node distance = 0.35cm] {$isp_{221}$};
+\node[op] (op_32) [below of = op_22] {$op_{32}$};
+\node[state] (ir_32) [right of = op_32] {};
+\node[ir] (isp_322) [below of = ir_32, node distance = 0.35cm] {$isp_{322}$};
+\node[ir] (isp_321) [above of = ir_32, node distance = 0.35cm] {$isp_{321}$};
+\node[op] (op_42) [right of = ir_22, node distance = 2.25cm] {$op_{42}$};
+\node[op] (op_52) [right of = ir_32, node distance = 2.25cm] {$op_{52}$};
+
+\path[->] 	
+			(op_11) 	edge node [above] {} (isp_111)
+			(isp_111) 	edge node [above] {} (op_21)
+			(op_21) 	edge node [above] {} (isp_211)
+			(op_21) 	edge node [above] {} (isp_212)
+			(isp_211) 	edge node [above] {} (op_41)
+			(isp_212) 	edge [bend left = 12.5] node [above] {} (op_42)
+			(op_31) 	edge node [above] {} (isp_311)
+			(op_31) 	edge node [above] {} (isp_312)
+			(isp_311) 	edge node [above] {} (op_41)			
+			(isp_311) 	edge node [above] {} (op_51)
+			(isp_312) 	edge node [above] {} (op_51)
+			(isp_312) 	edge [bend left = 7.5] node [above] {} (op_42)
+
+			(op_12) 	edge node [above] {} (isp_121)
+			(isp_121) 	edge node [above] {} (op_22)
+			(op_22) 	edge node [above] {} (isp_221)
+			(op_22) 	edge node [above] {} (isp_222)
+			(isp_221) 	edge [bend right = 7.5] node [above] {} (op_41)			
+			(isp_222) 	edge node [above] {} (op_42)
+			(op_32) 	edge node [above] {} (isp_321)
+			(op_32) 	edge node [above] {} (isp_322)
+			(isp_321) 	edge node [above] {} (op_52)
+			(isp_321) 	edge [bend right = 12.5] node [above] {} (op_41)
+			(isp_322) 	edge node [above] {} (op_52)
+			(isp_322) 	edge [bend right = 7.5] node [above] {} (op_42)
+			;
+\end{tikzpicture}
+\label{fig:ExecutionGraph}
+\caption{The ExecutionGraph is the physical view of a Flink job.}
+\end{figure}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% WindowWC snippet comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}
 
@@
 The window results are computed once the watermark passes the end of the window (Trigger).
+Flink also offer shart hands for commonly used functionalities.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Trigger snippet comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-A short hand notation for the example above is:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Shorthand snippet comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Global window snippet comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
 

### 2015-11-07 src/genwin.scala

@@ -1,0 +1,7 @@
+//general window API
+stream
+  .window(SlidingTimeWindows.of(Time.seconds(6), Time.seconds(2))
+  .trigger(TimeTrigger.create())
+
+//equivalent short hand for common use case
+stream.timeWindow(Time.seconds(6), Time.seconds(2))

### 2015-11-07 src/glowin.scala

@@ -1,0 +1,4 @@
+stream
+  .window(GlobalWindow.create)
+  .trigger(Count.of(1000))
+  .evict(Count.of(100))

### 2015-11-07 src/wwc.scala

@@ -1,0 +1,20 @@
+object WindowWordCount {
+  def main(args: Array[String]) {
+
+    val env = StreamExecutionEnvironment.getExecutionEnvironment
+    val text = env.socketTextStream("localhost", 9999)
+
+    val counts = text
+	.flatMap {line =>
+		    line.toLowerCase.split("\\W+")
+				.filter { w => w.nonEmpty }}
+    .map {w =>  (w, 1) }
+    .keyBy(0)
+    .timeWindow(Time.of(5, TimeUnit.SECONDS))
+    .sum(1)
+
+    counts.print
+    env.execute("Window Stream WordCount")
+  }
+}
+

### 2015-11-10 paper.tex
First half of references

@@
 
+% url references
+\usepackage{hyperref}
+
+% in paragraph enums
+\usepackage{paralist}
+
 \begin{document}
@@
 Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases.
-This has led to approaches such a ``micro-batching'' [], where a batch processor is used to simulate continuous processing, and the ``lambda architecture'', where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
+This has led to approaches such a ``micro-batching''~\cite{Zaharia:2013:DSF:2517349.2522737}, where a batch processor is used to simulate continuous processing, and the ``lambda architecture''~\cite{marz2015big}, where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
 
@@
 Apache Flink is a platform that was designed to deliver on these requirements.
-It is an open source project licensed by the Apache Software Foundation (ASF), with its community counting more than 120 developers, and growing fast [].
-Flink originates from the database research community, in particular the Stratosphere research project [], and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
+It is an open source project licensed by the Apache Software Foundation (ASF)~\cite{CUSTOM:web/Apache}, with its community counting more than 120 developers, and growing fast.~\cite{CUSTOM:web/FlinkContrib} Flink originates from the database research community, in particular the Stratosphere research project~\cite{stratosphere}, and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
 In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.
@@
 Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high.
-Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions [socc10, vldb12] and efficiently embed iterations within a DAG-based dataflow system [vldb12].
+Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions~\cite{battre2010nephele, DBLP:journals/pvldb/EwenTKM12} and efficiently embed iterations within a DAG-based dataflow system~\cite{DBLP:journals/pvldb/EwenTKM12}.
 
@@
 On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs.
-Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.
-
+Currently these are the following:
+\begin{inparaenum}[i)]
+  \item Gelly, an API and library for processing graphs,
+  \item FlinkML, an API and library for composing Machine Learning pipelines and
+  \item Table, an API similar in spirit to Microsoft's LINQ.
+\end{inparaenum}
 A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager.
@@
 Normally this possible by using persistent message queueing systems, e.g.
-Kafka.
+Apache Kafka.~\cite{CUSTOM:web/Kafka}
 
-Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS) [arXiv paper].
-It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
+Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
 It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
-Similar to Chandy-Lamport distributed snapshots [Chandy/Lamport], the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
+Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
 ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
 Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
-Furthermore, ABS [arXiv paper] supports snapshotting of cyclic graphs.
+Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.
 The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ABS figure comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h!]
+	\centering  	
+  	\includegraphics[width=.45\textwidth]{figs/snaps.pdf}
+	\caption{Asynchronous Barrier Snapshotting.}
+	\label{fig:FlinkStack}
+\end{figure}
 
@@
 
-In distributed systems, there is an arbitrary lag between event-time and processing-time [dataflow].
-This may mean arbitrary delays for getting an answer based on event-time semantics.
+In distributed systems, there is an arbitrary lag between event-time and processing-time.~\cite{akidau2015dataflow} This may mean arbitrary delays for getting an answer based on event-time semantics.
 To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
@@
 \textbf{Streaming Windows.} A window defines a logical group of records that are processed together.
-Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google.
+Flink's windowing system largely follows the Dataflow model~\cite{akidau2015dataflow} proposed by Google.
 A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
@@
 The evictor defines which records to keep in each group.
-The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner).
+The following example defines windows that operate in event time of 6 seconds that slide every 2 seconds (Assigner).
 The window results are computed once the watermark passes the end of the window (Trigger).
-Flink also offer shart hands for commonly used functionalities.
+Flink also offer short hands for commonly used functionalities.
 
@@
 Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality [Dataflow], and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
+This mechanism can be used to implement a wide variety of windowing functionality~\cite{akidau2015dataflow}, and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
 
@@
 Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above.
-However, i) the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops) and ii) we can simplify the fault tolerance mechanisms and iii) we can apply query optimization borrowing ideas from MPP database systems.
-
+However, \begin{inparaenum}[i)]
+  \item the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops),
+  \item we can simplify the fault tolerance mechanisms and
+  \item we can apply query optimization borrowing ideas from MPP database systems.
+\end{inparaenum}
 For these reasons, Flink treats specially batch computations, by implementing the above optimizations.
@@
 These operators simply block until they have received all of their input.
-Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad [Dryad].
+Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad.~\cite{isard2007dryad}
 
@@
 For the same reason, cardinality and cost estimation methods are equally difficult to employ.
-Flink's optimizer employs a number of novel methods for overcoming these issues [blackboxes, stratosphere-journal, iterations] for which we provide a short overview below.
+Flink's optimizer employs a number of novel methods for overcoming these issues~\cite{blackBoxes, stratosphere, DBLP:journals/pvldb/EwenTKM12} for which we provide a short overview below.
 Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
-Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation [scope-optimizer], using a cost-based approach to choose among multiple physical plans.
+Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation~\cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans.
 The cost includes network/disk I/O and CPU cost.
@@
 \textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
-These memory segments resemble database blocks into which, java objects representing the tuples that go through the runtime are serialized.
+These memory segments resemble database blocks into which, Java objects representing the tuples that go through the runtime are serialized.
 Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
@@
 \textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
-Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG [Hadoop iterative works, Spark], hiding from  the engine that it is executing an iterative program.
-The approach, implemented in Naiad [] adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration [].
+Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
+The approach, implemented in Naiad~\cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
+A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration.~\cite{low2012distributed}
 
 Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines [], and outperforms the driver-based approach [].
+This approach simulates a cyclic dataflow within a DAG engine making Flink i) competitive with specialized graph engines [], and outperforms the driver-based approach [].
 Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
@@
 
-\section{References}
-
 \bibliographystyle{abbrv}

### 2015-11-10 paper.tex
Most of the references in place

@@
 Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i) competitive with specialized graph engines [], and outperforms the driver-based approach [].
+This approach simulates a cyclic dataflow within a DAG engine making Flink competitive with specialized graph engines [], while outperforming the driver-based approach [].
 Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
@@
 
-\textbf{Batch Processing.} Apache Hadoop [] is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm [].
-Dryad [], introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE [] which added a language and an SQL optimizer on top of it.
-Apache Tez [] can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases [], and recent open-source implementations [Apache Drill, Impala] restrict their API to SQL variants.
-Very similar to Flink, Apache Spark [] is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
-In contrast, Flink is the only system that i) sports an optimizer that can optimize DAG programs beyond SQL queries, ii) is able to perform very efficient iterative processing natively, iii)  includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
+\textbf{Batch Processing.} Apache Hadoop~\cite{CUSTOM:web/Hadoop} is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
+Dryad~\cite{dryad}, introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer} which added a language and an SQL optimizer on top of it.
+Apache Tez~\cite{CUSTOM:web/Tez} can be seen as an open source implementation of the ideas proposed in Dryad.
+MPP databases [], and recent open-source implementations~\cite{CUSTOM:web/Drill} [Impala] restrict their API to SQL variants.
+Very similar to Flink, Apache Spark~\cite{CUSTOM:web/Spark} is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+In contrast, Flink is the only system that
+\begin{inparaenum}[i)]
+  \item supports an optimizer that can optimize DAG programs beyond SQL queries,
+  \item is able to perform very efficient iterative processing natively,
+  \item includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
+\end{inparaenum}
 
 \textbf{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis [], Stanford STREAMS [], Trill [], and IBM System S [].
-Most of the above systems are either i) academic prototypes, ii) closed-source commercial products, or iii) do not scale the computation horizontally on clusters of commodity servers.
-Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide very low level APIs and offer only at-least-once and at-most-once guarantees.
-MillWheel [], a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow [] to implement  out-of-order stream analytics.
-To the best of our knowledge, Flink is the only open-source project that i) offers high level APIs and processing libraries, ii) provides state management with exactly-once guarantees and iii) achieves high throughput and low latency, serving both batch and true streaming computations equally well.
+Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{chandramouli2014trill}.
+Most of the above systems are either
+\begin{inparaenum}[i)]
+  \item academic prototypes,
+  \item closed-source commercial products,
+  \item do not scale the computation horizontally on clusters of commodity servers.
+\end{inparaenum}
+Newer open source streaming systems that scale horizontally, such as Apache Storm~\cite{CUSTOM:web/Storm} and Apache Samza~\cite{CUSTOM:web/Samza}, provide low level APIs and offer only at-least-once and at-most-once guarantees.
+MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics.
+To the best of our knowledge, Flink is the only open-source project that
+\begin{inparaenum}[i)]
+  \item offers high level APIs and processing libraries,
+  \item provides state management with exactly-once guarantees and
+  \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.
+\end{inparaenum}
 

### 2015-11-10 architecture.tex
Split and jerk

--- 2015-11-10 paper.tex
@@
-\documentclass{sig-alternate}
+%!TEX root = paper.tex
 
-% A hack to make deauthor not complain which redefines
-% the titlepage command
-\newcommand{\titlepage}{}
-
-\usepackage{deauthor,times,graphicx}
-
-% JobGraph and Execution figs
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata,calc,positioning,shapes.geometric}
-
-% trademark symbol
-\usepackage{textcomp}
-
-% code listings
-\usepackage{listings}
-
-\definecolor{dkgreen}{rgb}{0,0.6,0}
-\definecolor{gray}{rgb}{0.5,0.5,0.5}
-\definecolor{mauve}{rgb}{0.58,0,0.82}
-
-% there is no built in support or Scala yet, good enough
-\lstset{frame=tb,
-  language=Java,
-  aboveskip=3mm,
-  belowskip=3mm,
-  showstringspaces=false,
-  columns=flexible,
-  basicstyle={\scriptsize\ttfamily},
-  numbers=none,
-  numberstyle=\tiny\color{black},
-  keywordstyle=\color{blue},
-  commentstyle=\color{dkgreen},
-  stringstyle=\color{mauve},
-  breaklines=true,
-  breakatwhitespace=true,
-  tabsize=4,
-  %for scala
-  emph={%  
-    object, def, val, zip%
-    },emphstyle={\color{blue}}%
-}%
-
-% url references
-\usepackage{hyperref}
-
-% in paragraph enums
-\usepackage{paralist}
-
-\begin{document}
-
-\title{Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics}
-\author{Authors}
-
-\maketitle
-
-\begin{abstract}
-Abstract TBD.
-\end{abstract}
-
-\section{Introduction}
-
-Organizations rely more and more on the value of their data, which is directly correlated with its freshness.
-In addition, as the data infrastructure becomes more and more complex, development time is focused on integrating different systems together, rather than focusing on business logic.
-The ever-growing need for lower latency analytics and scale, creates a pressing need for more real-time, asynchronous, data processing infrastructure.
-
-Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc.
-These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc.
-However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc.
-These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.
-Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases.
-This has led to approaches such a ``micro-batching''~\cite{Zaharia:2013:DSF:2517349.2522737}, where a batch processor is used to simulate continuous processing, and the ``lambda architecture''~\cite{marz2015big}, where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
-
-We believe that there is still a need for a real-time analytics platform, that is not covered by the existing solutions described above.
-Our observation is that modern enterprises need a platform that integrates well in the Hadoop-related Big Data processing zoo, supports user-defined computations, while being able to perform efficient, scalable and consistent data analysis, on top of both streaming/unbounded and stored/batch data.
-
-Apache Flink is a platform that was designed to deliver on these requirements.
-It is an open source project licensed by the Apache Software Foundation (ASF)~\cite{CUSTOM:web/Apache}, with its community counting more than 120 developers, and growing fast.~\cite{CUSTOM:web/FlinkContrib} Flink originates from the database research community, in particular the Stratosphere research project~\cite{stratosphere}, and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
-In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.
-
-This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results.
-Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high.
-Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions~\cite{battre2010nephele, DBLP:journals/pvldb/EwenTKM12} and efficiently embed iterations within a DAG-based dataflow system~\cite{DBLP:journals/pvldb/EwenTKM12}.
-
-\section{System architecture}
+\section{System Architecture}
 
@@
 	\centering  	
-  	\includegraphics[width=.45\textwidth]{figs/flink_stack.png}
+  	\includegraphics[width=.55\textwidth]{figs/stack.pdf}
+	\vspace{-2mm}
 	\caption{The Flink software stack.}
+	\vspace{-3mm}
 	\label{fig:FlinkStack}
@@
 
-\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
+\para{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
 A Job Graph is a DAG of operators and connections between operators.
@@
 TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
-
-\section{Execution model}
-
-\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
-There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
-For example, the graph below consists of five operators ($op_1$-$op_5$), and three intermediate results ($ir_1$-$ir_3$).
-Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
-reading data from a file system, a socket, a message queue, etc.), and data sinks.
-Operators produce intermediate results, as well as updates to state.
-An intermediate result is a logical handle (pointer) to the data that is produced by one operator.
-An intermediate result can be consumed by one or more operators.
-Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
-
-\node[op] (op_1) {$op_1$};
-\node[ir] (ir_1) [right of = op_1] {$ir_1$};
-\node[op] (op_2) [right of = ir_1] {$op_2$};
-\node[ir] (ir_2) [right of = op_2] {$ir_2$};
-\node[op] (op_3) [below of = op_2] {$op_3$};
-\node[ir] (ir_3) [right of = op_3] {$ir_3$};
-\node[op] (op_4) [right of = ir_2] {$op_4$};
-\node[op] (op_5) [right of = ir_3] {$op_5$};
-
-\path[->] 	
-			(op_1) 	edge node [above] {} (ir_1)
-			(ir_1) 	edge node [above] {} (op_2)
-			(op_2) 	edge node [above] {} (ir_2)
-			(ir_2) 	edge node [above] {} (op_4)
-			(op_3) 	edge node [above] {} (ir_3)
-			(ir_3) 	edge node [above] {} (op_4)
-			(ir_3) 	edge node [above] {} (op_5)
-			;
-\end{tikzpicture}
-\label{fig:JobGraph}
-\caption{The JobGraph is the logical view of a Flink job.}
-\end{figure}
-
-\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph.
-An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
-The JobGraph in fig.
-\ref{fig:JobGraph}, if it is partitioned using a parallelism of two, it results to the ExecutionGraph shown in fig.
-\ref{fig:ExecutionGraph}.
-Assume that $op_1$ and $op_2$ are data sources, $op_2$ is a map operator, $op_4$ is a join operator that needs both inputs partitioned by the same key, and $op_5$ is a filter.
-Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
-The resulting ExecutionGraph is the following:
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.6cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{every state}=[circle, fill,draw=none,white,text=black]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.2, draw, minimum size=0.8cm, text width=0.8cm, align=center]
-
-\node[op] (op_11) {$op_{11}$};
-\node[ir] (isp_111) [right of = op_11] {$isp_{111}$};
-\node[op] (op_21) [right of = isp_111] {$op_{21}$};
-\node[state] (ir_21) [right of = op_21] {};
-\node[ir] (isp_212) [below of = ir_21, node distance = 0.35cm] {$isp_{212}$};
-\node[ir] (isp_211) [above of = ir_21, node distance = 0.35cm] {$isp_{211}$};
-\node[op] (op_31) [below of = op_21] {$op_{31}$};
-\node[state] (ir_31) [right of = op_31] {};
-\node[ir] (isp_312) [below of = ir_31, node distance = 0.35cm] {$isp_{312}$};
-\node[ir] (isp_311) [above of = ir_31, node distance = 0.35cm] {$isp_{311}$};
-\node[op] (op_41) [right of = ir_21, node distance = 2.25cm] {$op_{41}$};
-\node[op] (op_51) [right of = ir_31, node distance = 2.25cm] {$op_{51}$};
-
-\node[op] (op_22) [below of = op_31] {$op_{22}$};
-\node[ir] (isp_121) [left of = op_22] {$isp_{121}$};
-\node[op] (op_12) [left of = isp_121] {$op_{12}$};
-\node[state] (ir_22) [right of = op_22] {};
-\node[ir] (isp_222) [below of = ir_22, node distance = 0.35cm] {$isp_{222}$};
-\node[ir] (isp_221) [above of = ir_22, node distance = 0.35cm] {$isp_{221}$};
-\node[op] (op_32) [below of = op_22] {$op_{32}$};
-\node[state] (ir_32) [right of = op_32] {};
-\node[ir] (isp_322) [below of = ir_32, node distance = 0.35cm] {$isp_{322}$};
-\node[ir] (isp_321) [above of = ir_32, node distance = 0.35cm] {$isp_{321}$};
-\node[op] (op_42) [right of = ir_22, node distance = 2.25cm] {$op_{42}$};
-\node[op] (op_52) [right of = ir_32, node distance = 2.25cm] {$op_{52}$};
-
-\path[->] 	
-			(op_11) 	edge node [above] {} (isp_111)
-			(isp_111) 	edge node [above] {} (op_21)
-			(op_21) 	edge node [above] {} (isp_211)
-			(op_21) 	edge node [above] {} (isp_212)
-			(isp_211) 	edge node [above] {} (op_41)
-			(isp_212) 	edge [bend left = 12.5] node [above] {} (op_42)
-			(op_31) 	edge node [above] {} (isp_311)
-			(op_31) 	edge node [above] {} (isp_312)
-			(isp_311) 	edge node [above] {} (op_41)			
-			(isp_311) 	edge node [above] {} (op_51)
-			(isp_312) 	edge node [above] {} (op_51)
-			(isp_312) 	edge [bend left = 7.5] node [above] {} (op_42)
-
-			(op_12) 	edge node [above] {} (isp_121)
-			(isp_121) 	edge node [above] {} (op_22)
-			(op_22) 	edge node [above] {} (isp_221)
-			(op_22) 	edge node [above] {} (isp_222)
-			(isp_221) 	edge [bend right = 7.5] node [above] {} (op_41)			
-			(isp_222) 	edge node [above] {} (op_42)
-			(op_32) 	edge node [above] {} (isp_321)
-			(op_32) 	edge node [above] {} (isp_322)
-			(isp_321) 	edge node [above] {} (op_52)
-			(isp_321) 	edge [bend right = 12.5] node [above] {} (op_41)
-			(isp_322) 	edge node [above] {} (op_52)
-			(isp_322) 	edge [bend right = 7.5] node [above] {} (op_42)
-			;
-\end{tikzpicture}
-\label{fig:ExecutionGraph}
-\caption{The ExecutionGraph is the physical view of a Flink job.}
-\end{figure}
-
-\textbf{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
-Buffers contain one or more records, and a record can span multiple buffers.
-Buffers are requested and relinquished from local buffer pools, shared among operators that live in the same task manager.
-An IRP is simply a collection of buffers.
-Work in Flink progresses (i.e., records flow through the pipeline) as long as there are buffers available, essentially implementing distributed blocking queues (the logical streams) with bounded capacity (the amount of memory available to the buffer pools, which can be configured by the user).
-This mechanism, in addition to implementing record network transfers doubles down as a natural way to backpressure the flow in the case of slow operators (including external systems that consume data).
-
-We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
-Internally, intermediate results are abstract classes with many implementations.
-These implementations can perform pipelined data exchange, and blocking data exchange.
-
-\textbf{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
-Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
-Flink implements pipelining by implementing intermediate results which activate network transfers between the producer task and consumer task,  as soon as their first buffer is available.
-Flink allows the configuration of its buffers' size and timeout.
-A buffer is sent to its consumer task as soon it is filled,  or as soon as a timeout is reached.
-Hence, by setting the buffer size and/or timeout to lower values, one can reduce latency, while achieving the opposite effect (i.e., high throughput) by using higher values.
-
-\textbf{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
-To do that, the system has to materialize intermediate results (in memory or disk).
-Flink implements blocking data exchange via an intermediate result that signals its availability only when all the buffers from the producer have been materialized.
-The cached buffers can double down as a materialized reusable intermediate result.
-
-\section{Fault Tolerance via \\ Asynchronous Snapshots}
-
-Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals.
-When failure occurs the state of the execution is restored from the latest snapshot.
-Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source.
-Normally this possible by using persistent message queueing systems, e.g.
-Apache Kafka.~\cite{CUSTOM:web/Kafka}
-
-Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
-It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
-Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
-ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
-Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
-Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.
-The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
-The following figure illustrates the snapshot process.
-
-\begin{figure}[h!]
-	\centering  	
-  	\includegraphics[width=.45\textwidth]{figs/snaps.pdf}
-	\caption{Asynchronous Barrier Snapshotting.}
-	\label{fig:FlinkStack}
-\end{figure}
-
-\section{Stream analytics on top of \\ dataflows}
-
-Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
-The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
-Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams.
-Each of the two supports specific operations since they exhibit different properties.
-For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations.
-Other transformations, such as (flat)map are  can be applied incrementally on data streams.
-
-% Give an example here instead of simplicity
-The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty.
-Here is, for example, a word count program on a simple  time-based window:
-
-\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}
-
-Flink distinguishes between two notions of time: 
-\begin{description}
-\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
-\item[Processing time] is the wall-clock time of the machine that is processing the data.
-\end{description}
-
-In distributed systems, there is an arbitrary lag between event-time and processing-time.~\cite{akidau2015dataflow} This may mean arbitrary delays for getting an answer based on event-time semantics.
-To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
-The watermarks aid the execution engine to process events in the correct event order.
-
-Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency.
-Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
-Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system.
-This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.
-
-\textbf{Streaming Windows.} A window defines a logical group of records that are processed together.
-Flink's windowing system largely follows the Dataflow model~\cite{akidau2015dataflow} proposed by Google.
-A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
-The assigner predefines the logical groups to which each record belongs, e.g.
-count-based or time-based periodic windows.
-The trigger defines when the window operation is performed on each group.
-The evictor defines which records to keep in each group.
-The following example defines windows that operate in event time of 6 seconds that slide every 2 seconds (Assigner).
-The window results are computed once the watermark passes the end of the window (Trigger).
-Flink also offer short hands for commonly used functionalities.
-
-\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}
-
-A global window creates a single logical group.
-The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
-
-\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
-
-Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality~\cite{akidau2015dataflow}, and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
-
-Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.
-
-\section{Batch analytics on top of \\ dataflows}
-
-A bounded data set is a special case of an unbounded data stream.
-Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above.
-However, \begin{inparaenum}[i)]
-  \item the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops),
-  \item we can simplify the fault tolerance mechanisms and
-  \item we can apply query optimization borrowing ideas from MPP database systems.
-\end{inparaenum}
-For these reasons, Flink treats specially batch computations, by implementing the above optimizations.
-For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime.
-These operators simply block until they have received all of their input.
-Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad.~\cite{isard2007dryad}
-
-With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
-
-\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
-However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
-For the same reason, cardinality and cost estimation methods are equally difficult to employ.
-Flink's optimizer employs a number of novel methods for overcoming these issues~\cite{blackBoxes, stratosphere, DBLP:journals/pvldb/EwenTKM12} for which we provide a short overview below.
-Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
-Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation~\cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans.
-The cost includes network/disk I/O and CPU cost.
-To overcome the cardinality estimation issues that were mentioned earlier, Flink's optimizer uses hints that are provided by the programmer.
-
-\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
-These memory segments resemble database blocks into which, Java objects representing the tuples that go through the runtime are serialized.
-Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
-To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.
-By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.
-
-\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
-Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
-The approach, implemented in Naiad~\cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration.~\cite{low2012distributed}
-
-Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i) competitive with specialized graph engines [], and outperforms the driver-based approach [].
-Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
-Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
-
-\section{Related work}
-
-There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
-
-\textbf{Batch Processing.} Apache Hadoop [] is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm [].
-Dryad [], introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE [] which added a language and an SQL optimizer on top of it.
-Apache Tez [] can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases [], and recent open-source implementations [Apache Drill, Impala] restrict their API to SQL variants.
-Very similar to Flink, Apache Spark [] is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
-In contrast, Flink is the only system that i) sports an optimizer that can optimize DAG programs beyond SQL queries, ii) is able to perform very efficient iterative processing natively, iii)  includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
-
-\textbf{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis [], Stanford STREAMS [], Trill [], and IBM System S [].
-Most of the above systems are either i) academic prototypes, ii) closed-source commercial products, or iii) do not scale the computation horizontally on clusters of commodity servers.
-Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide very low level APIs and offer only at-least-once and at-most-once guarantees.
-MillWheel [], a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow [] to implement  out-of-order stream analytics.
-To the best of our knowledge, Flink is the only open-source project that i) offers high level APIs and processing libraries, ii) provides state management with exactly-once guarantees and iii) achieves high throughput and low latency, serving both batch and true streaming computations equally well.
-
-\section{Conclusion}
-
-In this paper we presented Apache Flink, a platform that implements one universal dataflow engine designed to perform both stream and batch analytics.
-Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by the batch and a data stream APIs.
-The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state, and to partition, transform, discretize and aggregate a data stream.
-While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
-
-\bibliographystyle{abbrv}
-\bibliography{references}
-
-\end{document}

### 2015-11-10 batch.tex
Split and jerk

--- 2015-11-10 paper.tex
@@
-\documentclass{sig-alternate}
+%!TEX root = paper.tex
 
-% A hack to make deauthor not complain which redefines
-% the titlepage command
-\newcommand{\titlepage}{}
-
-\usepackage{deauthor,times,graphicx}
-
-% JobGraph and Execution figs
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata,calc,positioning,shapes.geometric}
-
-% trademark symbol
-\usepackage{textcomp}
-
-% code listings
-\usepackage{listings}
-
-\definecolor{dkgreen}{rgb}{0,0.6,0}
-\definecolor{gray}{rgb}{0.5,0.5,0.5}
-\definecolor{mauve}{rgb}{0.58,0,0.82}
-
-% there is no built in support or Scala yet, good enough
-\lstset{frame=tb,
-  language=Java,
-  aboveskip=3mm,
-  belowskip=3mm,
-  showstringspaces=false,
-  columns=flexible,
-  basicstyle={\scriptsize\ttfamily},
-  numbers=none,
-  numberstyle=\tiny\color{black},
-  keywordstyle=\color{blue},
-  commentstyle=\color{dkgreen},
-  stringstyle=\color{mauve},
-  breaklines=true,
-  breakatwhitespace=true,
-  tabsize=4,
-  %for scala
-  emph={%  
-    object, def, val, zip%
-    },emphstyle={\color{blue}}%
-}%
-
-% url references
-\usepackage{hyperref}
-
-% in paragraph enums
-\usepackage{paralist}
-
-\begin{document}
-
-\title{Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics}
-\author{Authors}
-
-\maketitle
-
-\begin{abstract}
-Abstract TBD.
-\end{abstract}
-
-\section{Introduction}
-
-Organizations rely more and more on the value of their data, which is directly correlated with its freshness.
-In addition, as the data infrastructure becomes more and more complex, development time is focused on integrating different systems together, rather than focusing on business logic.
-The ever-growing need for lower latency analytics and scale, creates a pressing need for more real-time, asynchronous, data processing infrastructure.
-
-Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc.
-These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc.
-However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc.
-These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.
-Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases.
-This has led to approaches such a ``micro-batching''~\cite{Zaharia:2013:DSF:2517349.2522737}, where a batch processor is used to simulate continuous processing, and the ``lambda architecture''~\cite{marz2015big}, where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
-
-We believe that there is still a need for a real-time analytics platform, that is not covered by the existing solutions described above.
-Our observation is that modern enterprises need a platform that integrates well in the Hadoop-related Big Data processing zoo, supports user-defined computations, while being able to perform efficient, scalable and consistent data analysis, on top of both streaming/unbounded and stored/batch data.
-
-Apache Flink is a platform that was designed to deliver on these requirements.
-It is an open source project licensed by the Apache Software Foundation (ASF)~\cite{CUSTOM:web/Apache}, with its community counting more than 120 developers, and growing fast.~\cite{CUSTOM:web/FlinkContrib} Flink originates from the database research community, in particular the Stratosphere research project~\cite{stratosphere}, and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
-In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.
-
-This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results.
-Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high.
-Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions~\cite{battre2010nephele, DBLP:journals/pvldb/EwenTKM12} and efficiently embed iterations within a DAG-based dataflow system~\cite{DBLP:journals/pvldb/EwenTKM12}.
-
-\section{System architecture}
-
-In this section we lay out the architecture of Flink as a software stack, and as a distributed system.
-While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
-
-\begin{figure}[h!]
-	\centering  	
-  	\includegraphics[width=.45\textwidth]{figs/flink_stack.png}
-	\caption{The Flink software stack.}
-	\label{fig:FlinkStack}
-\end{figure}
-
-\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
-A Job Graph is a DAG of operators and connections between operators.
-There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing'').
-Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine.
-On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs.
-Currently these are the following:
-\begin{inparaenum}[i)]
-  \item Gelly, an API and library for processing graphs,
-  \item FlinkML, an API and library for composing Machine Learning pipelines and
-  \item Table, an API similar in spirit to Microsoft's LINQ.
-\end{inparaenum}
-A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager.
-The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager.
-The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
-DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).
-
-The JobManager is Flink's master node.
-It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers, receiving statistics, controls the tasks' lifecycle,  coordinates the fault tolerance machanisms.
-The actual data processing takes place in the TaskManagers, or worker processes that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
-TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
-
-\section{Execution model}
-
-\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
-There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
-For example, the graph below consists of five operators ($op_1$-$op_5$), and three intermediate results ($ir_1$-$ir_3$).
-Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
-reading data from a file system, a socket, a message queue, etc.), and data sinks.
-Operators produce intermediate results, as well as updates to state.
-An intermediate result is a logical handle (pointer) to the data that is produced by one operator.
-An intermediate result can be consumed by one or more operators.
-Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
-
-\node[op] (op_1) {$op_1$};
-\node[ir] (ir_1) [right of = op_1] {$ir_1$};
-\node[op] (op_2) [right of = ir_1] {$op_2$};
-\node[ir] (ir_2) [right of = op_2] {$ir_2$};
-\node[op] (op_3) [below of = op_2] {$op_3$};
-\node[ir] (ir_3) [right of = op_3] {$ir_3$};
-\node[op] (op_4) [right of = ir_2] {$op_4$};
-\node[op] (op_5) [right of = ir_3] {$op_5$};
-
-\path[->] 	
-			(op_1) 	edge node [above] {} (ir_1)
-			(ir_1) 	edge node [above] {} (op_2)
-			(op_2) 	edge node [above] {} (ir_2)
-			(ir_2) 	edge node [above] {} (op_4)
-			(op_3) 	edge node [above] {} (ir_3)
-			(ir_3) 	edge node [above] {} (op_4)
-			(ir_3) 	edge node [above] {} (op_5)
-			;
-\end{tikzpicture}
-\label{fig:JobGraph}
-\caption{The JobGraph is the logical view of a Flink job.}
-\end{figure}
-
-\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph.
-An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
-The JobGraph in fig.
-\ref{fig:JobGraph}, if it is partitioned using a parallelism of two, it results to the ExecutionGraph shown in fig.
-\ref{fig:ExecutionGraph}.
-Assume that $op_1$ and $op_2$ are data sources, $op_2$ is a map operator, $op_4$ is a join operator that needs both inputs partitioned by the same key, and $op_5$ is a filter.
-Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
-The resulting ExecutionGraph is the following:
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.6cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{every state}=[circle, fill,draw=none,white,text=black]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.2, draw, minimum size=0.8cm, text width=0.8cm, align=center]
-
-\node[op] (op_11) {$op_{11}$};
-\node[ir] (isp_111) [right of = op_11] {$isp_{111}$};
-\node[op] (op_21) [right of = isp_111] {$op_{21}$};
-\node[state] (ir_21) [right of = op_21] {};
-\node[ir] (isp_212) [below of = ir_21, node distance = 0.35cm] {$isp_{212}$};
-\node[ir] (isp_211) [above of = ir_21, node distance = 0.35cm] {$isp_{211}$};
-\node[op] (op_31) [below of = op_21] {$op_{31}$};
-\node[state] (ir_31) [right of = op_31] {};
-\node[ir] (isp_312) [below of = ir_31, node distance = 0.35cm] {$isp_{312}$};
-\node[ir] (isp_311) [above of = ir_31, node distance = 0.35cm] {$isp_{311}$};
-\node[op] (op_41) [right of = ir_21, node distance = 2.25cm] {$op_{41}$};
-\node[op] (op_51) [right of = ir_31, node distance = 2.25cm] {$op_{51}$};
-
-\node[op] (op_22) [below of = op_31] {$op_{22}$};
-\node[ir] (isp_121) [left of = op_22] {$isp_{121}$};
-\node[op] (op_12) [left of = isp_121] {$op_{12}$};
-\node[state] (ir_22) [right of = op_22] {};
-\node[ir] (isp_222) [below of = ir_22, node distance = 0.35cm] {$isp_{222}$};
-\node[ir] (isp_221) [above of = ir_22, node distance = 0.35cm] {$isp_{221}$};
-\node[op] (op_32) [below of = op_22] {$op_{32}$};
-\node[state] (ir_32) [right of = op_32] {};
-\node[ir] (isp_322) [below of = ir_32, node distance = 0.35cm] {$isp_{322}$};
-\node[ir] (isp_321) [above of = ir_32, node distance = 0.35cm] {$isp_{321}$};
-\node[op] (op_42) [right of = ir_22, node distance = 2.25cm] {$op_{42}$};
-\node[op] (op_52) [right of = ir_32, node distance = 2.25cm] {$op_{52}$};
-
-\path[->] 	
-			(op_11) 	edge node [above] {} (isp_111)
-			(isp_111) 	edge node [above] {} (op_21)
-			(op_21) 	edge node [above] {} (isp_211)
-			(op_21) 	edge node [above] {} (isp_212)
-			(isp_211) 	edge node [above] {} (op_41)
-			(isp_212) 	edge [bend left = 12.5] node [above] {} (op_42)
-			(op_31) 	edge node [above] {} (isp_311)
-			(op_31) 	edge node [above] {} (isp_312)
-			(isp_311) 	edge node [above] {} (op_41)			
-			(isp_311) 	edge node [above] {} (op_51)
-			(isp_312) 	edge node [above] {} (op_51)
-			(isp_312) 	edge [bend left = 7.5] node [above] {} (op_42)
-
-			(op_12) 	edge node [above] {} (isp_121)
-			(isp_121) 	edge node [above] {} (op_22)
-			(op_22) 	edge node [above] {} (isp_221)
-			(op_22) 	edge node [above] {} (isp_222)
-			(isp_221) 	edge [bend right = 7.5] node [above] {} (op_41)			
-			(isp_222) 	edge node [above] {} (op_42)
-			(op_32) 	edge node [above] {} (isp_321)
-			(op_32) 	edge node [above] {} (isp_322)
-			(isp_321) 	edge node [above] {} (op_52)
-			(isp_321) 	edge [bend right = 12.5] node [above] {} (op_41)
-			(isp_322) 	edge node [above] {} (op_52)
-			(isp_322) 	edge [bend right = 7.5] node [above] {} (op_42)
-			;
-\end{tikzpicture}
-\label{fig:ExecutionGraph}
-\caption{The ExecutionGraph is the physical view of a Flink job.}
-\end{figure}
-
-\textbf{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
-Buffers contain one or more records, and a record can span multiple buffers.
-Buffers are requested and relinquished from local buffer pools, shared among operators that live in the same task manager.
-An IRP is simply a collection of buffers.
-Work in Flink progresses (i.e., records flow through the pipeline) as long as there are buffers available, essentially implementing distributed blocking queues (the logical streams) with bounded capacity (the amount of memory available to the buffer pools, which can be configured by the user).
-This mechanism, in addition to implementing record network transfers doubles down as a natural way to backpressure the flow in the case of slow operators (including external systems that consume data).
-
-We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
-Internally, intermediate results are abstract classes with many implementations.
-These implementations can perform pipelined data exchange, and blocking data exchange.
-
-\textbf{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
-Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
-Flink implements pipelining by implementing intermediate results which activate network transfers between the producer task and consumer task,  as soon as their first buffer is available.
-Flink allows the configuration of its buffers' size and timeout.
-A buffer is sent to its consumer task as soon it is filled,  or as soon as a timeout is reached.
-Hence, by setting the buffer size and/or timeout to lower values, one can reduce latency, while achieving the opposite effect (i.e., high throughput) by using higher values.
-
-\textbf{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
-To do that, the system has to materialize intermediate results (in memory or disk).
-Flink implements blocking data exchange via an intermediate result that signals its availability only when all the buffers from the producer have been materialized.
-The cached buffers can double down as a materialized reusable intermediate result.
-
-\section{Fault Tolerance via \\ Asynchronous Snapshots}
-
-Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals.
-When failure occurs the state of the execution is restored from the latest snapshot.
-Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source.
-Normally this possible by using persistent message queueing systems, e.g.
-Apache Kafka.~\cite{CUSTOM:web/Kafka}
-
-Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
-It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
-Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
-ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
-Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
-Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.
-The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
-The following figure illustrates the snapshot process.
-
-\begin{figure}[h!]
-	\centering  	
-  	\includegraphics[width=.45\textwidth]{figs/snaps.pdf}
-	\caption{Asynchronous Barrier Snapshotting.}
-	\label{fig:FlinkStack}
-\end{figure}
-
-\section{Stream analytics on top of \\ dataflows}
-
-Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
-The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
-Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams.
-Each of the two supports specific operations since they exhibit different properties.
-For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations.
-Other transformations, such as (flat)map are  can be applied incrementally on data streams.
-
-% Give an example here instead of simplicity
-The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty.
-Here is, for example, a word count program on a simple  time-based window:
-
-\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}
-
-Flink distinguishes between two notions of time: 
-\begin{description}
-\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
-\item[Processing time] is the wall-clock time of the machine that is processing the data.
-\end{description}
-
-In distributed systems, there is an arbitrary lag between event-time and processing-time.~\cite{akidau2015dataflow} This may mean arbitrary delays for getting an answer based on event-time semantics.
-To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
-The watermarks aid the execution engine to process events in the correct event order.
-
-Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency.
-Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
-Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system.
-This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.
-
-\textbf{Streaming Windows.} A window defines a logical group of records that are processed together.
-Flink's windowing system largely follows the Dataflow model~\cite{akidau2015dataflow} proposed by Google.
-A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
-The assigner predefines the logical groups to which each record belongs, e.g.
-count-based or time-based periodic windows.
-The trigger defines when the window operation is performed on each group.
-The evictor defines which records to keep in each group.
-The following example defines windows that operate in event time of 6 seconds that slide every 2 seconds (Assigner).
-The window results are computed once the watermark passes the end of the window (Trigger).
-Flink also offer short hands for commonly used functionalities.
-
-\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}
-
-A global window creates a single logical group.
-The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
-
-\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
-
-Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality~\cite{akidau2015dataflow}, and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
-
-Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.
-
-\section{Batch analytics on top of \\ dataflows}
-
+\section{Batch analytics on top of dataflows}
+\label{sec:batch}
 A bounded data set is a special case of an unbounded data stream.
@@
 
-\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
+\para{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
 However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
@@
 
-\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
+\para{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
 These memory segments resemble database blocks into which, Java objects representing the tuples that go through the runtime are serialized.
@@
 
-\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
+\para{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
 Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
@@
 Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i) competitive with specialized graph engines [], and outperforms the driver-based approach [].
+This approach simulates a cyclic dataflow within a DAG engine making Flink competitive with specialized graph engines [], while outperforming the driver-based approach [].
 Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
 Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
-
-\section{Related work}
-
-There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
-
-\textbf{Batch Processing.} Apache Hadoop [] is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm [].
-Dryad [], introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE [] which added a language and an SQL optimizer on top of it.
-Apache Tez [] can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases [], and recent open-source implementations [Apache Drill, Impala] restrict their API to SQL variants.
-Very similar to Flink, Apache Spark [] is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
-In contrast, Flink is the only system that i) sports an optimizer that can optimize DAG programs beyond SQL queries, ii) is able to perform very efficient iterative processing natively, iii)  includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
-
-\textbf{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis [], Stanford STREAMS [], Trill [], and IBM System S [].
-Most of the above systems are either i) academic prototypes, ii) closed-source commercial products, or iii) do not scale the computation horizontally on clusters of commodity servers.
-Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide very low level APIs and offer only at-least-once and at-most-once guarantees.
-MillWheel [], a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow [] to implement  out-of-order stream analytics.
-To the best of our knowledge, Flink is the only open-source project that i) offers high level APIs and processing libraries, ii) provides state management with exactly-once guarantees and iii) achieves high throughput and low latency, serving both batch and true streaming computations equally well.
-
-\section{Conclusion}
-
-In this paper we presented Apache Flink, a platform that implements one universal dataflow engine designed to perform both stream and batch analytics.
-Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by the batch and a data stream APIs.
-The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state, and to partition, transform, discretize and aggregate a data stream.
-While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
-
-\bibliographystyle{abbrv}
-\bibliography{references}
-
-\end{document}

### 2015-11-10 exec-model.tex
Split and jerk

--- 2015-11-07 paper.tex
@@
-\documentclass{sig-alternate}
-
-% A hack to make deauthor not complain which redefines
-% the titlepage command
-\newcommand{\titlepage}{}
-
-\usepackage{deauthor,times,graphicx}
-
-% JobGraph and Execution figs
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata,calc,positioning,shapes.geometric}
-
-% trademark symbol
-\usepackage{textcomp}
-
-% code listings
-\usepackage{listings}
-
-\definecolor{dkgreen}{rgb}{0,0.6,0}
-\definecolor{gray}{rgb}{0.5,0.5,0.5}
-\definecolor{mauve}{rgb}{0.58,0,0.82}
-
-% there is no built in support or Scala yet, good enough
-\lstset{frame=tb,
-  language=Java,
-  aboveskip=3mm,
-  belowskip=3mm,
-  showstringspaces=false,
-  columns=flexible,
-  basicstyle={\scriptsize\ttfamily},
-  numbers=none,
-  numberstyle=\tiny\color{black},
-  keywordstyle=\color{blue},
-  commentstyle=\color{dkgreen},
-  stringstyle=\color{mauve},
-  breaklines=true,
-  breakatwhitespace=true,
-  tabsize=4,
-  %for scala
-  emph={%  
-    object, def, val, zip%
-    },emphstyle={\color{blue}}%
-}%
-
-\begin{document}
-
-\title{Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics}
-\author{Authors}
-
-\maketitle
-
-\begin{abstract}
-Abstract TBD.
-\end{abstract}
-
-\section{Introduction}
-
-Organizations rely more and more on the value of their data, which is directly correlated with its freshness.
-In addition, as the data infrastructure becomes more and more complex, development time is focused on integrating different systems together, rather than focusing on business logic.
-The ever-growing need for lower latency analytics and scale, creates a pressing need for more real-time, asynchronous, data processing infrastructure.
-
-Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc.
-These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc.
-However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc.
-These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.
-Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases.
-This has led to approaches such a ``micro-batching'' [], where a batch processor is used to simulate continuous processing, and the ``lambda architecture'', where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
-
-We believe that there is still a need for a real-time analytics platform, that is not covered by the existing solutions described above.
-Our observation is that modern enterprises need a platform that integrates well in the Hadoop-related Big Data processing zoo, supports user-defined computations, while being able to perform efficient, scalable and consistent data analysis, on top of both streaming/unbounded and stored/batch data.
-
-Apache Flink is a platform that was designed to deliver on these requirements.
-It is an open source project licensed by the Apache Software Foundation (ASF), with its community counting more than 120 developers, and growing fast [].
-Flink originates from the database research community, in particular the Stratosphere research project [], and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
-In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.
-
-This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results.
-Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high.
-Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions [socc10, vldb12] and efficiently embed iterations within a DAG-based dataflow system [vldb12].
-
-\section{System architecture}
-
-In this section we lay out the architecture of Flink as a software stack, and as a distributed system.
-While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
-
-\begin{figure}[h!]
-	\centering  	
-  	\includegraphics[width=.45\textwidth]{figs/flink_stack.png}
-	\caption{The Flink software stack.}
-	\label{fig:FlinkStack}
-\end{figure}
-
-\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
-A Job Graph is a DAG of operators and connections between operators.
-There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing'').
-Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine.
-On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs.
-Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.
-
-A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager.
-The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager.
-The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
-DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).
-
-The JobManager is Flink's master node.
-It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers, receiving statistics, controls the tasks' lifecycle,  coordinates the fault tolerance machanisms.
-The actual data processing takes place in the TaskManagers, or worker processes that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
-TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
+%!TEX root = paper.tex
 
 \section{Execution model}
-
-\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
+\label{sec:execution}
+\para{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
 There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
-For example, the graph below consists of five operators ($op_1$-$op_5$), and three intermediate results ($ir_1$-$ir_3$).
+For example, the graph below consists of five operators (circles), and three intermediate results.
 Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
@@
 Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
+When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph that consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs).
 
@@
 
-\node[op] (op_1) {$op_1$};
-\node[ir] (ir_1) [right of = op_1] {$ir_1$};
-\node[op] (op_2) [right of = ir_1] {$op_2$};
-\node[ir] (ir_2) [right of = op_2] {$ir_2$};
-\node[op] (op_3) [below of = op_2] {$op_3$};
-\node[ir] (ir_3) [right of = op_3] {$ir_3$};
-\node[op] (op_4) [right of = ir_2] {$op_4$};
-\node[op] (op_5) [right of = ir_3] {$op_5$};
+\node[op] (op_1) {~~~~~~~};
+\node[ir] (ir_1) [right of = op_1] {~~~~~~~};
+\node[op] (op_2) [right of = ir_1] {~~~~~~~};
+\node[ir] (ir_2) [right of = op_2] {~~~~~~~};
+\node[op] (op_3) [below of = op_2] {~~~~~~~};
+\node[ir] (ir_3) [right of = op_3] {~~~~~~~};
+\node[op] (op_4) [right of = ir_2] {~~~~~~~};
+\node[op] (op_5) [right of = ir_3] {~~~~~~~};
 
@@
 
-\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph.
-An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
-The JobGraph in fig.
-\ref{fig:JobGraph}, if it is partitioned using a parallelism of two, it results to the ExecutionGraph shown in fig.
-\ref{fig:ExecutionGraph}.
-Assume that $op_1$ and $op_2$ are data sources, $op_2$ is a map operator, $op_4$ is a join operator that needs both inputs partitioned by the same key, and $op_5$ is a filter.
-Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
-The resulting ExecutionGraph is the following:
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.6cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{every state}=[circle, fill,draw=none,white,text=black]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.2, draw, minimum size=0.8cm, text width=0.8cm, align=center]
-
-\node[op] (op_11) {$op_{11}$};
-\node[ir] (isp_111) [right of = op_11] {$isp_{111}$};
-\node[op] (op_21) [right of = isp_111] {$op_{21}$};
-\node[state] (ir_21) [right of = op_21] {};
-\node[ir] (isp_212) [below of = ir_21, node distance = 0.35cm] {$isp_{212}$};
-\node[ir] (isp_211) [above of = ir_21, node distance = 0.35cm] {$isp_{211}$};
-\node[op] (op_31) [below of = op_21] {$op_{31}$};
-\node[state] (ir_31) [right of = op_31] {};
-\node[ir] (isp_312) [below of = ir_31, node distance = 0.35cm] {$isp_{312}$};
-\node[ir] (isp_311) [above of = ir_31, node distance = 0.35cm] {$isp_{311}$};
-\node[op] (op_41) [right of = ir_21, node distance = 2.25cm] {$op_{41}$};
-\node[op] (op_51) [right of = ir_31, node distance = 2.25cm] {$op_{51}$};
-
-\node[op] (op_22) [below of = op_31] {$op_{22}$};
-\node[ir] (isp_121) [left of = op_22] {$isp_{121}$};
-\node[op] (op_12) [left of = isp_121] {$op_{12}$};
-\node[state] (ir_22) [right of = op_22] {};
-\node[ir] (isp_222) [below of = ir_22, node distance = 0.35cm] {$isp_{222}$};
-\node[ir] (isp_221) [above of = ir_22, node distance = 0.35cm] {$isp_{221}$};
-\node[op] (op_32) [below of = op_22] {$op_{32}$};
-\node[state] (ir_32) [right of = op_32] {};
-\node[ir] (isp_322) [below of = ir_32, node distance = 0.35cm] {$isp_{322}$};
-\node[ir] (isp_321) [above of = ir_32, node distance = 0.35cm] {$isp_{321}$};
-\node[op] (op_42) [right of = ir_22, node distance = 2.25cm] {$op_{42}$};
-\node[op] (op_52) [right of = ir_32, node distance = 2.25cm] {$op_{52}$};
-
-\path[->] 	
-			(op_11) 	edge node [above] {} (isp_111)
-			(isp_111) 	edge node [above] {} (op_21)
-			(op_21) 	edge node [above] {} (isp_211)
-			(op_21) 	edge node [above] {} (isp_212)
-			(isp_211) 	edge node [above] {} (op_41)
-			(isp_212) 	edge [bend left = 12.5] node [above] {} (op_42)
-			(op_31) 	edge node [above] {} (isp_311)
-			(op_31) 	edge node [above] {} (isp_312)
-			(isp_311) 	edge node [above] {} (op_41)			
-			(isp_311) 	edge node [above] {} (op_51)
-			(isp_312) 	edge node [above] {} (op_51)
-			(isp_312) 	edge [bend left = 7.5] node [above] {} (op_42)
-
-			(op_12) 	edge node [above] {} (isp_121)
-			(isp_121) 	edge node [above] {} (op_22)
-			(op_22) 	edge node [above] {} (isp_221)
-			(op_22) 	edge node [above] {} (isp_222)
-			(isp_221) 	edge [bend right = 7.5] node [above] {} (op_41)			
-			(isp_222) 	edge node [above] {} (op_42)
-			(op_32) 	edge node [above] {} (isp_321)
-			(op_32) 	edge node [above] {} (isp_322)
-			(isp_321) 	edge node [above] {} (op_52)
-			(isp_321) 	edge [bend right = 12.5] node [above] {} (op_41)
-			(isp_322) 	edge node [above] {} (op_52)
-			(isp_322) 	edge [bend right = 7.5] node [above] {} (op_42)
-			;
-\end{tikzpicture}
-\label{fig:ExecutionGraph}
-\caption{The ExecutionGraph is the physical view of a Flink job.}
-\end{figure}
-
-\textbf{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
+\para{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
 Buffers contain one or more records, and a record can span multiple buffers.
@@
 
-\textbf{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
+\para{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
 Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
@@
 
-\textbf{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
+\para{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
 To do that, the system has to materialize intermediate results (in memory or disk).
@@
 The cached buffers can double down as a materialized reusable intermediate result.
-
-\section{Fault Tolerance via \\ Asynchronous Snapshots}
-
-Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals.
-When failure occurs the state of the execution is restored from the latest snapshot.
-Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source.
-Normally this possible by using persistent message queueing systems, e.g.
-Kafka.
-
-Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS) [arXiv paper].
-It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
-It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
-Similar to Chandy-Lamport distributed snapshots [Chandy/Lamport], the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
-ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
-Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
-Furthermore, ABS [arXiv paper] supports snapshotting of cyclic graphs.
-The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
-The following figure illustrates the snapshot process.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ABS figure comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Stream analytics on top of \\ dataflows}
-
-Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
-The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
-Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams.
-Each of the two supports specific operations since they exhibit different properties.
-For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations.
-Other transformations, such as (flat)map are  can be applied incrementally on data streams.
-
-% Give an example here instead of simplicity
-The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty.
-Here is, for example, a word count program on a simple  time-based window:
-
-\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}
-
-Flink distinguishes between two notions of time: 
-\begin{description}
-\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
-\item[Processing time] is the wall-clock time of the machine that is processing the data.
-\end{description}
-
-In distributed systems, there is an arbitrary lag between event-time and processing-time [dataflow].
-This may mean arbitrary delays for getting an answer based on event-time semantics.
-To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
-The watermarks aid the execution engine to process events in the correct event order.
-
-Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency.
-Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
-Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system.
-This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.
-
-\textbf{Streaming Windows.} A window defines a logical group of records that are processed together.
-Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google.
-A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
-The assigner predefines the logical groups to which each record belongs, e.g.
-count-based or time-based periodic windows.
-The trigger defines when the window operation is performed on each group.
-The evictor defines which records to keep in each group.
-The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner).
-The window results are computed once the watermark passes the end of the window (Trigger).
-Flink also offer shart hands for commonly used functionalities.
-
-\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}
-
-A global window creates a single logical group.
-The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
-
-\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
-
-Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality [Dataflow], and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
-
-Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.
-
-\section{Batch analytics on top of \\ dataflows}
-
-A bounded data set is a special case of an unbounded data stream.
-Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above.
-However, i) the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops) and ii) we can simplify the fault tolerance mechanisms and iii) we can apply query optimization borrowing ideas from MPP database systems.
-
-For these reasons, Flink treats specially batch computations, by implementing the above optimizations.
-For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime.
-These operators simply block until they have received all of their input.
-Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad [Dryad].
-
-With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
-
-\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
-However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
-For the same reason, cardinality and cost estimation methods are equally difficult to employ.
-Flink's optimizer employs a number of novel methods for overcoming these issues [blackboxes, stratosphere-journal, iterations] for which we provide a short overview below.
-Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
-Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation [scope-optimizer], using a cost-based approach to choose among multiple physical plans.
-The cost includes network/disk I/O and CPU cost.
-To overcome the cardinality estimation issues that were mentioned earlier,  Flink's optimizer uses hints that are provided by the programmer.
-
-\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
-These memory segments resemble database blocks into which, java objects representing the tuples that go through the runtime are serialized.
-Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
-To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.
-By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.
-
-\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
-Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG [Hadoop iterative works, Spark], hiding from  the engine that it is executing an iterative program.
-The approach, implemented in Naiad [] adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration [].
-
-Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines [], and outperforms the driver-based approach [].
-Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
-Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
-
-\section{Related work}
-
-There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
-
-\textbf{Batch Processing.} Apache Hadoop [] is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm [].
-Dryad [], introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE [] which added a language and an SQL optimizer on top of it.
-Apache Tez [] can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases [], and recent open-source implementations [Apache Drill, Impala] restrict their API to SQL variants.
-Very similar to Flink, Apache Spark [] is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
-In contrast, Flink is the only system that i) sports an optimizer that can optimize DAG programs beyond SQL queries, ii) is able to perform very efficient iterative processing natively, iii)  includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
-
-\textbf{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis [], Stanford STREAMS [], Trill [], and IBM System S [].
-Most of the above systems are either i) academic prototypes, ii) closed-source commercial products, or iii) do not scale the computation horizontally on clusters of commodity servers.
-Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide very low level APIs and offer only at-least-once and at-most-once guarantees.
-MillWheel [], a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow [] to implement  out-of-order stream analytics.
-To the best of our knowledge, Flink is the only open-source project that i) offers high level APIs and processing libraries, ii) provides state management with exactly-once guarantees and iii) achieves high throughput and low latency, serving both batch and true streaming computations equally well.
-
-\section{Conclusion}
-
-In this paper we presented Apache Flink, a platform that implements one universal dataflow engine designed to perform both stream and batch analytics.
-Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by the batch and a data stream APIs.
-The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state, and to partition, transform, discretize and aggregate a data stream.
-While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
-
-\section{References}
-
-\bibliographystyle{abbrv}
-\bibliography{references}
-
-\end{document}

### 2015-11-10 intro.tex
Split and jerk

\section{Introduction}
\label{sec:intro}
Data stream processing (complex event processing / continuous queries) and static data processing (batch processing, OLAP) were traditionally considered as two very different types of applications. They were programmed using different programming models and APIs, and were executed by different systems (for example, dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft Streaminsight, or Streambase versus relational databases or Apache Hadoop execution engines). Traditionally, batch data analysis made up for the lion share of the use cases, data sizes, and market, while streaming data analysis existed mostly in specialized applications.

It is becoming more and more apparent, however, that a huge number of today's large scale data processing use cases handle data that is, in reality, continuously produced over time. These continuous streams of data come for example from web logs, application logs, sensors, or as changes to application state in databases (transaction log records). Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of the data production. Instead, data records are (often artificially) batched into static data sets (for example hourly/daily/monthly chunks) and then processed in a time-agnostic fashion. Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline. Architectural patterns like the "lambda architecture" [] combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results.

Apache Flink follows a paradigm that embraces data stream processing as the unifying model for real-time analysis, continuous streams, and batch processing, both in the programming model and in the execution engine. In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka [] or Amazon Kinesis []), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation. Through a highly flexible windowing mechanism, Flink programs can compute both early and approximate, as well as delayed and accurate results in the same operation, obviating the need to combine different systems for the two use cases. Flink supports different notions of time (event time, ingestion time, processing time) in order to give programmer high flexibility in defining how events should be correlated.

At the same time, Flink acknowledges that there is, and will be, a need for dedicated batch processing (dealing with static data sets). Complex queries over static data are still a good match for a batch processing abstraction. Furthermore, batch processing is still needed both for legacy implementations of streaming use cases, and for analysis applications where no efficient algorithms are yet known that perform this kind of processing on streaming data. Batch programs are special cases of streaming programs, where the stream is finite, and order/time of records does not matter (all records implicitly belong to one all-encompassing window). However, to support batch use cases with a competitive ease and performance, Flink has a specialized API for processing static data sets, uses specialized data structures and algorithms for the batch versions of operators like join or grouping, and uses dedicated scheduling strategies. The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning. 
Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing []), and is used in production in several companies.

\noindent The contributions of this paper are the following:
\begin{itemize}
	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
	\vspace{-3mm}
	\item We show how to make these dataflows fault tolerant via an asynchronous snapshotting mechanism (\autoref{sec:snapshots}).
	\vspace{-3mm}
	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:windows}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
\end{itemize}

### 2015-11-10 paper.tex
fix format
Split and jerk

--- 2015-11-07 paper.tex
@@
-\documentclass{sig-alternate}
+\documentclass[11pt]{article}
 
@@
 % the titlepage command
-\newcommand{\titlepage}{}
+% \newcommand{\titlepage}{}
 
@@
 \usepackage{listings}
+
+\newcommand{\sectionautorefname}{Section}%
+
+\newcommand{\para}[1]{\vspace{2mm}\noindent\textbf{#1}}
 
@@
 
+% url references
+\usepackage{hyperref}
+
+% in paragraph enums
+\usepackage{paralist}
+
 \begin{document}
 
-\title{Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics}
+\title{Apache Flink\texttrademark : Streaming Dataflows as a Basis \\ for Universal Data Analytics}
 \author{Authors}
@@
 \begin{abstract}
-Abstract TBD.
+Apache Flink is an open source system for processing streaming and batch data.
+Flink is built on the philosophy that many classes of data processing applications, including real-time analytics, continuous data pipelines, historic data processing (batch), and iterative algorithms (machine learning, graph analysis) can be expressed and executed as pipelined fault tolerant dataflows.
+In this paper, we present Flink’s architecture and expand on how a (seemingly diverse) set of use cases can be unified under the same execution model.
 \end{abstract}
 
-\section{Introduction}
-
-Organizations rely more and more on the value of their data, which is directly correlated with its freshness.
-In addition, as the data infrastructure becomes more and more complex, development time is focused on integrating different systems together, rather than focusing on business logic.
-The ever-growing need for lower latency analytics and scale, creates a pressing need for more real-time, asynchronous, data processing infrastructure.
-
-Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc.
-These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc.
-However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc.
-These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.
-Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases.
-This has led to approaches such a ``micro-batching'' [], where a batch processor is used to simulate continuous processing, and the ``lambda architecture'', where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.
-
-We believe that there is still a need for a real-time analytics platform, that is not covered by the existing solutions described above.
-Our observation is that modern enterprises need a platform that integrates well in the Hadoop-related Big Data processing zoo, supports user-defined computations, while being able to perform efficient, scalable and consistent data analysis, on top of both streaming/unbounded and stored/batch data.
-
-Apache Flink is a platform that was designed to deliver on these requirements.
-It is an open source project licensed by the Apache Software Foundation (ASF), with its community counting more than 120 developers, and growing fast [].
-Flink originates from the database research community, in particular the Stratosphere research project [], and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages.
-In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.
-
-This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results.
-Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high.
-Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions [socc10, vldb12] and efficiently embed iterations within a DAG-based dataflow system [vldb12].
-
-\section{System architecture}
-
-In this section we lay out the architecture of Flink as a software stack, and as a distributed system.
-While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
-
-\begin{figure}[h!]
-	\centering  	
-  	\includegraphics[width=.45\textwidth]{figs/flink_stack.png}
-	\caption{The Flink software stack.}
-	\label{fig:FlinkStack}
-\end{figure}
-
-\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
-A Job Graph is a DAG of operators and connections between operators.
-There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing'').
-Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine.
-On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs.
-Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.
-
-A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager.
-The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager.
-The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
-DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).
-
-The JobManager is Flink's master node.
-It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers, receiving statistics, controls the tasks' lifecycle,  coordinates the fault tolerance machanisms.
-The actual data processing takes place in the TaskManagers, or worker processes that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
-TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
-
-\section{Execution model}
-
-\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
-There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
-For example, the graph below consists of five operators ($op_1$-$op_5$), and three intermediate results ($ir_1$-$ir_3$).
-Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
-reading data from a file system, a socket, a message queue, etc.), and data sinks.
-Operators produce intermediate results, as well as updates to state.
-An intermediate result is a logical handle (pointer) to the data that is produced by one operator.
-An intermediate result can be consumed by one or more operators.
-Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
-
-\node[op] (op_1) {$op_1$};
-\node[ir] (ir_1) [right of = op_1] {$ir_1$};
-\node[op] (op_2) [right of = ir_1] {$op_2$};
-\node[ir] (ir_2) [right of = op_2] {$ir_2$};
-\node[op] (op_3) [below of = op_2] {$op_3$};
-\node[ir] (ir_3) [right of = op_3] {$ir_3$};
-\node[op] (op_4) [right of = ir_2] {$op_4$};
-\node[op] (op_5) [right of = ir_3] {$op_5$};
-
-\path[->] 	
-			(op_1) 	edge node [above] {} (ir_1)
-			(ir_1) 	edge node [above] {} (op_2)
-			(op_2) 	edge node [above] {} (ir_2)
-			(ir_2) 	edge node [above] {} (op_4)
-			(op_3) 	edge node [above] {} (ir_3)
-			(ir_3) 	edge node [above] {} (op_4)
-			(ir_3) 	edge node [above] {} (op_5)
-			;
-\end{tikzpicture}
-\label{fig:JobGraph}
-\caption{The JobGraph is the logical view of a Flink job.}
-\end{figure}
-
-\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph.
-An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys.
-The JobGraph in fig.
-\ref{fig:JobGraph}, if it is partitioned using a parallelism of two, it results to the ExecutionGraph shown in fig.
-\ref{fig:ExecutionGraph}.
-Assume that $op_1$ and $op_2$ are data sources, $op_2$ is a map operator, $op_4$ is a join operator that needs both inputs partitioned by the same key, and $op_5$ is a filter.
-Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present).
-The resulting ExecutionGraph is the following:
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.6cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{every state}=[circle, fill,draw=none,white,text=black]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.2, draw, minimum size=0.8cm, text width=0.8cm, align=center]
-
-\node[op] (op_11) {$op_{11}$};
-\node[ir] (isp_111) [right of = op_11] {$isp_{111}$};
-\node[op] (op_21) [right of = isp_111] {$op_{21}$};
-\node[state] (ir_21) [right of = op_21] {};
-\node[ir] (isp_212) [below of = ir_21, node distance = 0.35cm] {$isp_{212}$};
-\node[ir] (isp_211) [above of = ir_21, node distance = 0.35cm] {$isp_{211}$};
-\node[op] (op_31) [below of = op_21] {$op_{31}$};
-\node[state] (ir_31) [right of = op_31] {};
-\node[ir] (isp_312) [below of = ir_31, node distance = 0.35cm] {$isp_{312}$};
-\node[ir] (isp_311) [above of = ir_31, node distance = 0.35cm] {$isp_{311}$};
-\node[op] (op_41) [right of = ir_21, node distance = 2.25cm] {$op_{41}$};
-\node[op] (op_51) [right of = ir_31, node distance = 2.25cm] {$op_{51}$};
-
-\node[op] (op_22) [below of = op_31] {$op_{22}$};
-\node[ir] (isp_121) [left of = op_22] {$isp_{121}$};
-\node[op] (op_12) [left of = isp_121] {$op_{12}$};
-\node[state] (ir_22) [right of = op_22] {};
-\node[ir] (isp_222) [below of = ir_22, node distance = 0.35cm] {$isp_{222}$};
-\node[ir] (isp_221) [above of = ir_22, node distance = 0.35cm] {$isp_{221}$};
-\node[op] (op_32) [below of = op_22] {$op_{32}$};
-\node[state] (ir_32) [right of = op_32] {};
-\node[ir] (isp_322) [below of = ir_32, node distance = 0.35cm] {$isp_{322}$};
-\node[ir] (isp_321) [above of = ir_32, node distance = 0.35cm] {$isp_{321}$};
-\node[op] (op_42) [right of = ir_22, node distance = 2.25cm] {$op_{42}$};
-\node[op] (op_52) [right of = ir_32, node distance = 2.25cm] {$op_{52}$};
-
-\path[->] 	
-			(op_11) 	edge node [above] {} (isp_111)
-			(isp_111) 	edge node [above] {} (op_21)
-			(op_21) 	edge node [above] {} (isp_211)
-			(op_21) 	edge node [above] {} (isp_212)
-			(isp_211) 	edge node [above] {} (op_41)
-			(isp_212) 	edge [bend left = 12.5] node [above] {} (op_42)
-			(op_31) 	edge node [above] {} (isp_311)
-			(op_31) 	edge node [above] {} (isp_312)
-			(isp_311) 	edge node [above] {} (op_41)			
-			(isp_311) 	edge node [above] {} (op_51)
-			(isp_312) 	edge node [above] {} (op_51)
-			(isp_312) 	edge [bend left = 7.5] node [above] {} (op_42)
-
-			(op_12) 	edge node [above] {} (isp_121)
-			(isp_121) 	edge node [above] {} (op_22)
-			(op_22) 	edge node [above] {} (isp_221)
-			(op_22) 	edge node [above] {} (isp_222)
-			(isp_221) 	edge [bend right = 7.5] node [above] {} (op_41)			
-			(isp_222) 	edge node [above] {} (op_42)
-			(op_32) 	edge node [above] {} (isp_321)
-			(op_32) 	edge node [above] {} (isp_322)
-			(isp_321) 	edge node [above] {} (op_52)
-			(isp_321) 	edge [bend right = 12.5] node [above] {} (op_41)
-			(isp_322) 	edge node [above] {} (op_52)
-			(isp_322) 	edge [bend right = 7.5] node [above] {} (op_42)
-			;
-\end{tikzpicture}
-\label{fig:ExecutionGraph}
-\caption{The ExecutionGraph is the physical view of a Flink job.}
-\end{figure}
-
-\textbf{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
-Buffers contain one or more records, and a record can span multiple buffers.
-Buffers are requested and relinquished from local buffer pools, shared among operators that live in the same task manager.
-An IRP is simply a collection of buffers.
-Work in Flink progresses (i.e., records flow through the pipeline) as long as there are buffers available, essentially implementing distributed blocking queues (the logical streams) with bounded capacity (the amount of memory available to the buffer pools, which can be configured by the user).
-This mechanism, in addition to implementing record network transfers doubles down as a natural way to backpressure the flow in the case of slow operators (including external systems that consume data).
-
-We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
-Internally, intermediate results are abstract classes with many implementations.
-These implementations can perform pipelined data exchange, and blocking data exchange.
-
-\textbf{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
-Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
-Flink implements pipelining by implementing intermediate results which activate network transfers between the producer task and consumer task,  as soon as their first buffer is available.
-Flink allows the configuration of its buffers' size and timeout.
-A buffer is sent to its consumer task as soon it is filled,  or as soon as a timeout is reached.
-Hence, by setting the buffer size and/or timeout to lower values, one can reduce latency, while achieving the opposite effect (i.e., high throughput) by using higher values.
-
-\textbf{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
-To do that, the system has to materialize intermediate results (in memory or disk).
-Flink implements blocking data exchange via an intermediate result that signals its availability only when all the buffers from the producer have been materialized.
-The cached buffers can double down as a materialized reusable intermediate result.
-
-\section{Fault Tolerance via \\ Asynchronous Snapshots}
-
-Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals.
-When failure occurs the state of the execution is restored from the latest snapshot.
-Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source.
-Normally this possible by using persistent message queueing systems, e.g.
-Kafka.
-
-Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS) [arXiv paper].
-It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
-It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
-Similar to Chandy-Lamport distributed snapshots [Chandy/Lamport], the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
-ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
-Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
-Furthermore, ABS [arXiv paper] supports snapshotting of cyclic graphs.
-The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
-The following figure illustrates the snapshot process.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ABS figure comes here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Stream analytics on top of \\ dataflows}
-
-Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
-The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
-Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams.
-Each of the two supports specific operations since they exhibit different properties.
-For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations.
-Other transformations, such as (flat)map are  can be applied incrementally on data streams.
-
-% Give an example here instead of simplicity
-The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty.
-Here is, for example, a word count program on a simple  time-based window:
-
-\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}
-
-Flink distinguishes between two notions of time: 
-\begin{description}
-\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
-\item[Processing time] is the wall-clock time of the machine that is processing the data.
-\end{description}
-
-In distributed systems, there is an arbitrary lag between event-time and processing-time [dataflow].
-This may mean arbitrary delays for getting an answer based on event-time semantics.
-To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
-The watermarks aid the execution engine to process events in the correct event order.
-
-Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency.
-Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
-Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system.
-This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.
-
-\textbf{Streaming Windows.} A window defines a logical group of records that are processed together.
-Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google.
-A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
-The assigner predefines the logical groups to which each record belongs, e.g.
-count-based or time-based periodic windows.
-The trigger defines when the window operation is performed on each group.
-The evictor defines which records to keep in each group.
-The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner).
-The window results are computed once the watermark passes the end of the window (Trigger).
-Flink also offer shart hands for commonly used functionalities.
-
-\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}
-
-A global window creates a single logical group.
-The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
-
-\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
-
-Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality [Dataflow], and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
-
-Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.
-
-\section{Batch analytics on top of \\ dataflows}
-
-A bounded data set is a special case of an unbounded data stream.
-Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above.
-However, i) the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops) and ii) we can simplify the fault tolerance mechanisms and iii) we can apply query optimization borrowing ideas from MPP database systems.
-
-For these reasons, Flink treats specially batch computations, by implementing the above optimizations.
-For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime.
-These operators simply block until they have received all of their input.
-Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad [Dryad].
-
-With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
-
-\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
-However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
-For the same reason, cardinality and cost estimation methods are equally difficult to employ.
-Flink's optimizer employs a number of novel methods for overcoming these issues [blackboxes, stratosphere-journal, iterations] for which we provide a short overview below.
-Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
-Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation [scope-optimizer], using a cost-based approach to choose among multiple physical plans.
-The cost includes network/disk I/O and CPU cost.
-To overcome the cardinality estimation issues that were mentioned earlier,  Flink's optimizer uses hints that are provided by the programmer.
-
-\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
-These memory segments resemble database blocks into which, java objects representing the tuples that go through the runtime are serialized.
-Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
-To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.
-By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.
-
-\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
-Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG [Hadoop iterative works, Spark], hiding from  the engine that it is executing an iterative program.
-The approach, implemented in Naiad [] adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration [].
-
-Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines [], and outperforms the driver-based approach [].
-Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
-Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
-
-\section{Related work}
-
-There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
-
-\textbf{Batch Processing.} Apache Hadoop [] is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm [].
-Dryad [], introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE [] which added a language and an SQL optimizer on top of it.
-Apache Tez [] can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases [], and recent open-source implementations [Apache Drill, Impala] restrict their API to SQL variants.
-Very similar to Flink, Apache Spark [] is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
-In contrast, Flink is the only system that i) sports an optimizer that can optimize DAG programs beyond SQL queries, ii) is able to perform very efficient iterative processing natively, iii)  includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
-
-\textbf{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis [], Stanford STREAMS [], Trill [], and IBM System S [].
-Most of the above systems are either i) academic prototypes, ii) closed-source commercial products, or iii) do not scale the computation horizontally on clusters of commodity servers.
-Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide very low level APIs and offer only at-least-once and at-most-once guarantees.
-MillWheel [], a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow [] to implement  out-of-order stream analytics.
-To the best of our knowledge, Flink is the only open-source project that i) offers high level APIs and processing libraries, ii) provides state management with exactly-once guarantees and iii) achieves high throughput and low latency, serving both batch and true streaming computations equally well.
+\input{intro}
+\input{architecture}
+\input{exec-model}
+\input{snapshots}
+\input{streaming}
+\input{batch}
+\input{related}
 
 \section{Conclusion}
-
+\label{sec:conclusions}
 In this paper we presented Apache Flink, a platform that implements one universal dataflow engine designed to perform both stream and batch analytics.

### 2015-11-10 related.tex
Split and jerk

%!TEX root = paper.tex

\section{Related work}
\label{sec:related}
There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.

\para{Batch Processing.} Apache Hadoop~\cite{CUSTOM:web/Hadoop} is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}. Dryad~\cite{isard2007dryad}, introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer} which added a language and an SQL optimizer on top of it. Apache Tez~\cite{CUSTOM:web/Tez} can be seen as an open source implementation of the ideas proposed in Dryad. MPP databases [], and recent open-source implementations~\cite{CUSTOM:web/Drill} [Impala] restrict their API to SQL variants. Very similar to Flink, Apache Spark~\cite{CUSTOM:web/Spark} is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches. In contrast, Flink is the only system that
\begin{inparaenum}[i)]
  \item supports an optimizer that can optimize DAG programs beyond SQL queries,
  \item is able to perform very efficient iterative processing natively,
  \item includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
\end{inparaenum}

\para{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams. Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{chandramouli2014trill}. Most of the above systems are either
\begin{inparaenum}[i)]
  \item academic prototypes,
  \item closed-source commercial products,
  \item do not scale the computation horizontally on clusters of commodity servers.
\end{inparaenum}
Newer open source streaming systems that scale horizontally, such as Apache Storm~\cite{CUSTOM:web/Storm} and Apache Samza~\cite{CUSTOM:web/Samza}, provide low level APIs and offer only at-least-once and at-most-once guarantees. MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics. To the best of our knowledge, Flink is the only open-source project that
\begin{inparaenum}[i)]
  \item offers high level APIs and processing libraries,
  \item provides state management with exactly-once guarantees and
  \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.
\end{inparaenum}

### 2015-11-10 snapshots.tex
Split and jerk

%!TEX root = paper.tex

\section{Fault Tolerance via Asynchronous Snapshots}
\label{sec:snapshots}
Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals. When failure occurs the state of the execution is restored from the latest snapshot. Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the dryad snapshot are reprocessed in the same order at each respective source. Normally this possible by using persistent message queueing systems, e.g. Apache Kafka.~\cite{CUSTOM:web/Kafka}

Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner. It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream. Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum. ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further. Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records. Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.  The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated. The following figure illustrates the snapshot process.

\begin{figure}[h!]
	\centering  	
  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
  	\vspace{-4mm}
	\caption{Asynchronous Barrier Snapshotting.}
	\vspace{-2mm}
	\label{fig:FlinkStack}
\end{figure}

### 2015-11-10 streaming.tex
Split and jerk

%!TEX root = paper.tex

\section{Stream Analytics on Top of Dataflows}
\label{sec:windows}

Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state. The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type. Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams. Each of the two supports specific operations since they exhibit different properties. For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations. Other transformations, such as (flat)map are  can be applied incrementally on data streams.

% Give an example here instead of simplicity
The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty. Here is, for example, a word count program on a simple  time-based window:

\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}

Flink distinguishes between two notions of time: 
\begin{description}
\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
\item[Processing time] is the wall-clock time of the machine that is processing the data.
\end{description}

In distributed systems, there is an arbitrary lag between event-time and processing-time \cite{akidau2015dataflow}. This may mean arbitrary delays for getting an answer based on event-time semantics. To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system. The watermarks aid the execution engine to process events in the correct event order.

Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency. Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag. Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system. This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.

\para{Streaming Windows.} A window defines a logical group of records that are processed together. Flink's windowing system largely follows the Dataflow model~\cite{akidau2015dataflow} proposed by Google. A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor. 
The assigner predefines the logical groups to which each record belongs, e.g. count-based or time-based periodic windows. The trigger defines when the window operation is performed on each group. The evictor defines which records to keep in each group. The following example defines windows that operate in event time of 6 seconds that slide every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger). Flink also offer short hands for commonly used functionalities.

\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}

A global window creates a single logical group. The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).

\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}

Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality~\cite{akidau2015dataflow}, and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).

Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections. We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.

### 2015-11-10 intro.tex
Added a couple of citations

@@
+%!TEX root = paper.tex
+
 \section{Introduction}
@@
 Apache Flink follows a paradigm that embraces data stream processing as the unifying model for real-time analysis, continuous streams, and batch processing, both in the programming model and in the execution engine.
-In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka [] or Amazon Kinesis []), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
+In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka \cite{CUSTOM:web/Kafka} or Amazon Kinesis\footnote{https://aws.amazon.com/kinesis/}), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
 Through a highly flexible windowing mechanism, Flink programs can compute both early and approximate, as well as delayed and accurate results in the same operation, obviating the need to combine different systems for the two use cases.

### 2015-11-10 related.tex
Added a couple of citations

@@
 Apache Tez~\cite{CUSTOM:web/Tez} can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases [], and recent open-source implementations~\cite{CUSTOM:web/Drill} [Impala] restrict their API to SQL variants.
-Very similar to Flink, Apache Spark~\cite{CUSTOM:web/Spark} is a very popular data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+MPP databases \cite{dewitt1990gamma} , and recent open-source implementations~\cite{CUSTOM:web/Drill,kornacker2015impala} restrict their API to SQL variants.
+Similar to Flink, Apache Spark~\cite{CUSTOM:web/Spark} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
 In contrast, Flink is the only system that
@@
 \para{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as Telegraph [], Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{chandramouli2014trill}.
+Many of these systems are based on research in the database community in projects such as PSoup/Telegraph \cite{chandrasekaran2003psoup}, Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{gedik2008spade}.
 Most of the above systems are either
@@
   \item academic prototypes,
-  \item closed-source commercial products,
+  \item closed-source commercial products, or 
   \item do not scale the computation horizontally on clusters of commodity servers.

### 2015-11-10 batch.tex
pass over the batch processing section

@@
 These operators simply block until they have received all of their input.
-Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad.~\cite{isard2007dryad}
-
-With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
+Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad \cite{isard2007dryad}.
+In the sequel, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
 
 \para{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
-However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
+However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer.
 For the same reason, cardinality and cost estimation methods are equally difficult to employ.
@@
 \para{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
-Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
+Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
 The approach, implemented in Naiad~\cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration.~\cite{low2012distributed}
+A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration \cite{low2012distributed}.
 
 Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink competitive with specialized graph engines [], while outperforming the driver-based approach [].
-Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
+This approach simulates a cyclic dataflow within a DAG engine making Flink competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}, while outperforming the driver-based approach.
+To do that, Flink supports delta iterations, which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
 Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.

### 2015-11-10 architecture.tex
Figures and listings update

@@
 While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
-
-\begin{figure}[h!]
-	\centering  	
-  	\includegraphics[width=.55\textwidth]{figs/stack.pdf}
-	\vspace{-2mm}
-	\caption{The Flink software stack.}
-	\vspace{-3mm}
-	\label{fig:FlinkStack}
-\end{figure}
 
@@
 The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
-DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).
+DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in~\autoref{sec:batch}).
 
@@
 TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
+
+\begin{figure}[ht]
+\centering     %%% not \center
+\subfigure[The Flink software stack.]{\label{fig:FlinkStack}\includegraphics[width=.5\textwidth]{figs/stack.pdf}}
+\subfigure[The JobGraph is the logical view of a Flink job.]{\label{fig:JobGraph}\begin{tikzpicture}
+[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
+\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20, text width=0.5cm]
+\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
+
+\node[op] (op_1) {~~~~~~~};
+\node[ir] (ir_1) [right of = op_1] {~~~~~~~};
+\node[op] (op_2) [right of = ir_1] {~~~~~~~};
+\node[ir] (ir_2) [right of = op_2] {~~~~~~~};
+\node[op] (op_3) [below of = op_2] {~~~~~~~};
+\node[ir] (ir_3) [right of = op_3] {~~~~~~~};
+\node[op] (op_4) [right of = ir_2] {~~~~~~~};
+\node[op] (op_5) [right of = ir_3] {~~~~~~~};
+
+\path[->] 	
+			(op_1) 	edge node [above] {} (ir_1)
+			(ir_1) 	edge node [above] {} (op_2)
+			(op_2) 	edge node [above] {} (ir_2)
+			(ir_2) 	edge node [above] {} (op_4)
+			(op_3) 	edge node [above] {} (ir_3)
+			(ir_3) 	edge node [above] {} (op_4)
+			(ir_3) 	edge node [above] {} (op_5)
+			;
+\end{tikzpicture}}
+\caption{Apache Flink architecture.}
+\end{figure}

### 2015-11-10 exec-model.tex
Figures and listings update

@@
 \para{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
-There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs).
-For example, the graph below consists of five operators (circles), and three intermediate results.
+There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs) as presented on Figure \autoref{fig:JobGraph}.
+For example, the aforementioned graph consists of five operators (circles), and three intermediate results (cylinders).
 Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
@@
 When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph that consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs).
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
-
-\node[op] (op_1) {~~~~~~~};
-\node[ir] (ir_1) [right of = op_1] {~~~~~~~};
-\node[op] (op_2) [right of = ir_1] {~~~~~~~};
-\node[ir] (ir_2) [right of = op_2] {~~~~~~~};
-\node[op] (op_3) [below of = op_2] {~~~~~~~};
-\node[ir] (ir_3) [right of = op_3] {~~~~~~~};
-\node[op] (op_4) [right of = ir_2] {~~~~~~~};
-\node[op] (op_5) [right of = ir_3] {~~~~~~~};
-
-\path[->] 	
-			(op_1) 	edge node [above] {} (ir_1)
-			(ir_1) 	edge node [above] {} (op_2)
-			(op_2) 	edge node [above] {} (ir_2)
-			(ir_2) 	edge node [above] {} (op_4)
-			(op_3) 	edge node [above] {} (ir_3)
-			(ir_3) 	edge node [above] {} (op_4)
-			(ir_3) 	edge node [above] {} (op_5)
-			;
-\end{tikzpicture}
-\label{fig:JobGraph}
-\caption{The JobGraph is the logical view of a Flink job.}
-\end{figure}
 

### 2015-11-10 src/genwin.scala

@@ -1,7 +1,11 @@
 //general window API
-stream
-  .window(SlidingTimeWindows.of(Time.seconds(6), Time.seconds(2))
+stream.window(SlidingTimeWindows.of(Time.seconds(6), Time.seconds(2))
   .trigger(TimeTrigger.create())
 
 //equivalent short hand for common use case
 stream.timeWindow(Time.seconds(6), Time.seconds(2))
+
+//global window
+stream.window(GlobalWindow.create)
+  .trigger(Count.of(1000))
+  .evict(Count.of(100))

### 2015-11-10 src/wwc.scala

@@ -5,8 +5,7 @@ object WindowWordCount {
     val text = env.socketTextStream("localhost", 9999)
 
     val counts = text
-	.flatMap {line =>
-		    line.toLowerCase.split("\\W+")
+	.flatMap {line => line.toLowerCase.split("\\W+")
 				.filter { w => w.nonEmpty }}
     .map {w =>  (w, 1) }
     .keyBy(0)

### 2015-11-10 intro.tex
Links moved to footnote

--- 2015-11-10 intro.tex
@@
 Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline.
-Architectural patterns like the "lambda architecture" [] combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results.
+Architectural patterns like the "lambda architecture"~\cite{marz2015big} combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results.
 
 Apache Flink follows a paradigm that embraces data stream processing as the unifying model for real-time analysis, continuous streams, and batch processing, both in the programming model and in the execution engine.
-In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka [] or Amazon Kinesis []), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
+In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka\footnote{http://kafka.apache.org/} or Amazon Kinesis\footnote{https://aws.amazon.com/kinesis/}), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
 Through a highly flexible windowing mechanism, Flink programs can compute both early and approximate, as well as delayed and accurate results in the same operation, obviating the need to combine different systems for the two use cases.
@@
 The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning.
-Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing []), and is used in production in several companies.
+Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing~\cite{CUSTOM:web/FlinkContrib}), and is used in production in several companies.
 

### 2015-11-10 related.tex
Links moved to footnote

@@
 
-\para{Batch Processing.} Apache Hadoop~\cite{CUSTOM:web/Hadoop} is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
+\para{Batch Processing.} Apache Hadoop\footnote{https://hadoop.apache.org/} is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
 Dryad~\cite{isard2007dryad}, introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer} which added a language and an SQL optimizer on top of it.
-Apache Tez~\cite{CUSTOM:web/Tez} can be seen as an open source implementation of the ideas proposed in Dryad.
-MPP databases \cite{dewitt1990gamma} , and recent open-source implementations~\cite{CUSTOM:web/Drill,kornacker2015impala} restrict their API to SQL variants.
-Similar to Flink, Apache Spark~\cite{CUSTOM:web/Spark} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+Apache Tez\footnote{https://tez.apache.org/} can be seen as an open source implementation of the ideas proposed in Dryad.
+MPP databases \cite{dewitt1990gamma}, and recent open-source implementations~\cite{CUSTOM:web/Drill,kornacker2015impala} restrict their API to SQL variants.
+Similar to Flink, Apache Spark\footnote{spark.apache.org} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
 In contrast, Flink is the only system that
@@
 \end{inparaenum}
-Newer open source streaming systems that scale horizontally, such as Apache Storm~\cite{CUSTOM:web/Storm} and Apache Samza~\cite{CUSTOM:web/Samza}, provide low level APIs and offer only at-least-once and at-most-once guarantees.
+Newer open source streaming systems that scale horizontally, such as Apache Storm\footnote{http://storm.apache.org/} and Apache Samza\footnote{http://samza.apache.org/}, provide low level APIs and offer only at-least-once and at-most-once guarantees.
 MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics.

### 2015-11-13 architecture.tex
Incorporated latest suggestions from doc

@@
 
-\para{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs.
-A Job Graph is a DAG of operators and connections between operators.
-There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing'').
-Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine.
-On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs.
-Currently these are the following:
-\begin{inparaenum}[i)]
-  \item Gelly, an API and library for processing graphs,
-  \item FlinkML, an API and library for composing Machine Learning pipelines and
-  \item Table, an API similar in spirit to Microsoft's LINQ.
-\end{inparaenum}
-A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager.
-The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager.
+\para{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs.
+A Flink runtime program is a DAG of stateful operators connected with data streams.
+There are two core APIs in Flink: the DataSet API for processing finite data sets (often referred to as “batch processing”), and the DataStream API for processing potentially unbounded data streams (often referred to as “stream processing”).
+Flink’s core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create runtime programs executable by the engine.
+As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing.
+On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML, Gelly, and Table).
+
+A Flink cluster comprises of three types of processes: the client, the JobManager (JM - master), and one or more TaskManagers (RM - workers).
+The client takes the program code, transforms it to a runtime executable, and submits it to the JobManager.
 The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
-DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in~\autoref{sec:batch}).
+DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in ???).
 
-The JobManager is Flink's master node.
-It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers, receiving statistics, controls the tasks' lifecycle,  coordinates the fault tolerance machanisms.
-The actual data processing takes place in the TaskManagers, or worker processes that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
+The JobManager is Flink’s master node.
+It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers and receiving statistics, controls the tasks’ lifecycle, and coordinates checkpointing used for recovery.
+The actual data processing takes place in the TaskManagers that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
 TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
+
+\begin{figure}
+\centering
+	\includegraphics[width=.5\textwidth]{figs/architecture}
+\caption{Flink's Process Model.}
+\end{figure}
 

### 2015-11-13 batch.tex
Incorporated latest suggestions from doc

--- 2015-11-10 batch.tex
@@
 \label{sec:batch}
+
 A bounded data set is a special case of an unbounded data stream.
-Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above.
-However, \begin{inparaenum}[i)]
-  \item the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops),
-  \item we can simplify the fault tolerance mechanisms and
-  \item we can apply query optimization borrowing ideas from MPP database systems.
-\end{inparaenum}
-For these reasons, Flink treats specially batch computations, by implementing the above optimizations.
-For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime.
-These operators simply block until they have received all of their input.
-Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad.~\cite{isard2007dryad}
+Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by Flink’s features that we presented above.
+However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and iii) programs that process bounded data sets are amenable to additional optimizations, more efficient book keeping for fault tolerance, and staged scheduling.
+.
 
-With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.
+\vspace{2mm}
+\noindent Flink approaches batch processing as follows:
+\begin{itemize}
+	\item Batch computations are executed by the same runtime as streaming computations.
+The runtime executable may be parameterized with blocked data streams to break up large computations in isolated stages that are scheduled successively.
+	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).
+	\item Blocking operators (e.g., joins and sorts) are simply operator implementations that happen to block until they have consumed their entire input.
+The runtime is not aware of whether an operator is blocking or not.
+These operators use managed memory provided by Flink (either on- or off- the JVM heap), and can spill to disk if their inputs exceed their memory bounds.
+	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc)
+	\item A query optimization layer transforms a DataSet program to an efficient executable.
+\end{itemize}
 
-\para{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties.
-However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer.
+\para{Query Optimization.} Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation.
+However, the arbitrary UDF-heavy DAGs that make up Flink’s dataflow programs do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer.
 For the same reason, cardinality and cost estimation methods are equally difficult to employ.
-Flink's optimizer employs a number of novel methods for overcoming these issues~\cite{blackBoxes, stratosphere, DBLP:journals/pvldb/EwenTKM12} for which we provide a short overview below.
-Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
-Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation~\cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans.
+Flink’s runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
+Flink’s optimizer enumerates different physical plans based on the concept of interesting properties propagation \cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans.
 The cost includes network/disk I/O and CPU cost.
-To overcome the cardinality estimation issues that were mentioned earlier, Flink's optimizer uses hints that are provided by the programmer.
+To overcome the cardinality estimation issues that were mentioned earlier,  Flink’s optimizer uses hints that are provided by the programmer.
 
-\para{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments.
-These memory segments resemble database blocks into which, Java objects representing the tuples that go through the runtime are serialized.
-Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
+\para{Memory Management.} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into a  memory segments.
+Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
 To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.
-By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.
+By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and use cache-efficient and robust algorithms [Graefe join paper] that scale gracefully under memory pressure.
 
-\para{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine.
-Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
-The approach, implemented in Naiad~\cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration.~\cite{low2012distributed}
+\para{Native Iterations on top of Dataflows.}  Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
+The approach, implemented in Naiad \cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
+A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration \cite{low2012distributed}.
 
-Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink competitive with specialized graph engines [], while outperforming the driver-based approach [].
-Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API.
+Flink follows an approach that retains the DAG-based runtime, but allows for special “head” and “tail” tasks to signify the beginning and end of iteration, that exchange data via shared memory.
+This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}, and outperforms the driver-based approach.
+Flink supports delta iterations, which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink’s Graph API.
 Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.

### 2015-11-13 intro.tex
Incorporated latest suggestions from doc

@@
 \label{sec:intro}
-Data stream processing (complex event processing / continuous queries) and static data processing (batch processing, OLAP) were traditionally considered as two very different types of applications.
-They were programmed using different programming models and APIs, and were executed by different systems (for example, dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft Streaminsight, or Streambase versus relational databases or Apache Hadoop execution engines).
-Traditionally, batch data analysis made up for the lion share of the use cases, data sizes, and market, while streaming data analysis existed mostly in specialized applications.
+Data stream processing (e.g, as exemplified by Complex Event Processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop ) were traditionally considered as two very different types of applications.
+They were programmed using different programming models and APIs, and were executed by different systems (for example, dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft Streaminsight, or Streambase versus relational databases or execution engines for Hadoop, including Apache Spark, Apache Drill, etc).
+Traditionally, batch data analysis made up for the lion share of the use cases, data sizes, and market, while streaming data analysis mostly served specialized applications.
 
@@
 These continuous streams of data come for example from web logs, application logs, sensors, or as changes to application state in databases (transaction log records).
-Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of the data production.
+Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of data production.
 Instead, data records are (often artificially) batched into static data sets (for example hourly/daily/monthly chunks) and then processed in a time-agnostic fashion.
 Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline.
-Architectural patterns like the "lambda architecture"~\cite{marz2015big} combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results.
+Architectural patterns like the "lambda architecture" \cite{marz2015big} combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results.
+All these approaches suffer from high latency (imposed by batches), high complexity (connecting and orchestrating several systems, and implementing business logic twice), as well as arbitrary inaccuracy, as the time dimension is not explicitly handled by the application code.
 
@@
 	\vspace{-3mm}
-	\item We show how to make these dataflows fault tolerant via an asynchronous snapshotting mechanism (\autoref{sec:snapshots}).
+	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault tolerant streaming dataflows (\autoref{sec:snapshots}).
 	\vspace{-3mm}

### 2015-11-13 snapshots.tex
Links moved to footnote
Incorporated latest suggestions from doc

@@
 \label{sec:snapshots}
-Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals.
-When failure occurs the state of the execution is restored from the latest snapshot.
-Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the dryad snapshot are reprocessed in the same order at each respective source.
-Normally this possible by using persistent message queueing systems, e.g.
-Apache Kafka.~\cite{CUSTOM:web/Kafka}
+Before jumping to the higher layers of the system, we detail how the common fabric (Flink’s runtime) guarantees consistent results in the case of failures.
+Since Flink data streams are replayable, discarding partial results and re-producing the lost partitions of the resulting streams (which possibly may trigger partial re-computation down to the sources in the case of non-buffered streams) is enough to guarantee fault tolerance when the computation is finite \cite{DBLP:conf/hotcloud/ZahariaCFSS10}.
+This is the mechanism used for Flink DataSet (batch) API programs.
 
-Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner.
-It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream.
-Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum.
-ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further.
-Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records.
-Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.
-The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated.
-The following figure illustrates the snapshot process.
+In stream processing (Flink’s DataStream API), the computation is possibly infinite, which makes the above approach not practical, as possibly months of computation will need to be replayed.
+To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.
 
-\begin{figure}[h!]
+The core challenge lies in taking a consistent snapshot of all parallel operators without stopping the topology.
+In essence, the snapshot of all operators should refer to the same logical time in the computation.
+The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
+Barriers are messages injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
+
+\begin{figure}[ht]
 	\centering  	
@@
 \end{figure}
+
+An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received.
+Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS).
+Once the state has been backed up, the operator forwards the barrier downstream.
+ABS bears resemblances to the seminal Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
+However, because of the DAG structure of a Flink program and the alignment phase, ABS does not need to checkpoint in-flight records, guaranteeing that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).
+
+Recovery from failures reverts the operator state to the most recently snapshotted one, and restarts the input streams from the latest barrier.
+This guarantees that the replayed work is limited to the processing between two checkpoints.
+We note that recovery need not start from the sources if intermediate streams have been partially buffered.
+
+\vspace{2mm}
+\noindent ABS provides several benefits:
+
+\begin{itemize}
+\item It guarantees exactly-once state updates without ever pausing the computation (e.g., as compared to discretized streams \cite{DBLP:conf/hotcloud/ZahariaCFSS10} which is essentially a synchronous implementation of snapshotting).
+\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval (again compare with \cite{DBLP:conf/hotcloud/ZahariaCFSS10})
+\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
+\end{itemize}

### 2015-11-25 architecture.tex
Updated buil for proposal
Updates from ShareLaTeX

@@
 \centering
-	\includegraphics[width=.5\textwidth]{figs/architecture}
+	\includegraphics[width=.75\textwidth]{figs/process_model_updated}
 \caption{Flink's Process Model.}
@@
 \centering     %%% not \center
-\subfigure[The Flink software stack.]{\label{fig:FlinkStack}\includegraphics[width=.5\textwidth]{figs/stack.pdf}}
-\subfigure[The JobGraph is the logical view of a Flink job.]{\label{fig:JobGraph}\begin{tikzpicture}
-[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
-\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20, text width=0.5cm]
-\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
+\label{fig:FlinkStack}
+\includegraphics[width=.5\textwidth]{figs/stack.pdf}
+% \subfigure[The JobGraph is the logical view of a Flink job.]{\label{fig:JobGraph}\begin{tikzpicture}
+% [shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
+% \tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20, text width=0.5cm]
+% \tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
 
-\node[op] (op_1) {~~~~~~~};
-\node[ir] (ir_1) [right of = op_1] {~~~~~~~};
-\node[op] (op_2) [right of = ir_1] {~~~~~~~};
-\node[ir] (ir_2) [right of = op_2] {~~~~~~~};
-\node[op] (op_3) [below of = op_2] {~~~~~~~};
-\node[ir] (ir_3) [right of = op_3] {~~~~~~~};
-\node[op] (op_4) [right of = ir_2] {~~~~~~~};
-\node[op] (op_5) [right of = ir_3] {~~~~~~~};
+% \node[op] (op_1) {~~~~~~~};
+% \node[ir] (ir_1) [right of = op_1] {~~~~~~~};
+% \node[op] (op_2) [right of = ir_1] {~~~~~~~};
+% \node[ir] (ir_2) [right of = op_2] {~~~~~~~};
+% \node[op] (op_3) [below of = op_2] {~~~~~~~};
+% \node[ir] (ir_3) [right of = op_3] {~~~~~~~};
+% \node[op] (op_4) [right of = ir_2] {~~~~~~~};
+% \node[op] (op_5) [right of = ir_3] {~~~~~~~};
 
-\path[->] 	
-			(op_1) 	edge node [above] {} (ir_1)
-			(ir_1) 	edge node [above] {} (op_2)
-			(op_2) 	edge node [above] {} (ir_2)
-			(ir_2) 	edge node [above] {} (op_4)
-			(op_3) 	edge node [above] {} (ir_3)
-			(ir_3) 	edge node [above] {} (op_4)
-			(ir_3) 	edge node [above] {} (op_5)
-			;
-\end{tikzpicture}}
-\caption{Apache Flink architecture.}
+% \path[->] 	
+% 			(op_1) 	edge node [above] {} (ir_1)
+% 			(ir_1) 	edge node [above] {} (op_2)
+% 			(op_2) 	edge node [above] {} (ir_2)
+% 			(ir_2) 	edge node [above] {} (op_4)
+% 			(op_3) 	edge node [above] {} (ir_3)
+% 			(ir_3) 	edge node [above] {} (op_4)
+% 			(ir_3) 	edge node [above] {} (op_5)
+% 			;
+% \end{tikzpicture}}
+\caption{Flink's software stack.}
 \end{figure}

### 2015-11-25 architecture.tex
Figures.
More copy-pasting.

@@
 
-\para{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs.
+\para{Flink's Runtime and APIS.} Flink's software stack can be seen in \autoref{fig:FlinkStack}.
+The core of Flink is the distributed dataflow engine, which executes dataflow programs.
 A Flink runtime program is a DAG of stateful operators connected with data streams.
@@
 
-A Flink cluster comprises of three types of processes: the client, the JobManager (JM - master), and one or more TaskManagers (RM - workers).
-The client takes the program code, transforms it to a runtime executable, and submits it to the JobManager.
-The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types.
-DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in ???).
+As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the JobManager (JM - master), and one or more TaskManagers (TM - workers).
+The client takes the program code, transforms it to a dataflow graph, and submits that to the JobManager.
+This transformation phase also examines the data types (schema) of the data exchanged between operators and creates serializers and other type/schema specific code.
+DataSet programs additionally go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in \autoref{sec:batch}).
 
-The JobManager is Flink’s master node.
-It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers and receiving statistics, controls the tasks’ lifecycle, and coordinates checkpointing used for recovery.
-The actual data processing takes place in the TaskManagers that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM.
-TMs communicate directly with each other using a multiplexed TCP connection per TM pair.
+The JobManager coordinates the distributed execution of the dataflow.
+It tracks the state and progress of each operator and stream, schedules new operators, and coordinates checkpoints and recovery.
+In a high-availability setup, the JobManager persists a minimal set of metadata at each checkpoint to a fault-tolerant storage, such that a standby JobManager can reconstruct the checkpoint and recover the dataflow execution from there (see \autoref{sec:fault-tolerance}).
+
+The actual data processing takes place in the TaskManagers.
+A TaskManager executes one or more operators that produce streams, and reports on their status to the JobManager.
+The TaskManagers maintain the buffer pools to buffer or materialize the streams, and the network connections to exchange the data streams between operators.
 
@@
 \centering
-	\includegraphics[width=.75\textwidth]{figs/process_model_updated}
+\begin{minipage}{0.49\textwidth}
+\centering
+\includegraphics[width=.88\textwidth]{figs/stack.pdf}
+\label{fig:FlinkStack}
+\caption{Flink's Software Stack.}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\centering
+	\includegraphics[width=.95\textwidth]{figs/architecture.pdf}
     \caption{Flink's Process Model.}
+    \label{fig:process-model}
+\end{minipage}%
 \end{figure}
-
-\begin{figure}[ht]
-\centering     %%% not \center
-\label{fig:FlinkStack}
-\includegraphics[width=.5\textwidth]{figs/stack.pdf}
-% \subfigure[The JobGraph is the logical view of a Flink job.]{\label{fig:JobGraph}\begin{tikzpicture}
-% [shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
-% \tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20, text width=0.5cm]
-% \tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]
-
-% \node[op] (op_1) {~~~~~~~};
-% \node[ir] (ir_1) [right of = op_1] {~~~~~~~};
-% \node[op] (op_2) [right of = ir_1] {~~~~~~~};
-% \node[ir] (ir_2) [right of = op_2] {~~~~~~~};
-% \node[op] (op_3) [below of = op_2] {~~~~~~~};
-% \node[ir] (ir_3) [right of = op_3] {~~~~~~~};
-% \node[op] (op_4) [right of = ir_2] {~~~~~~~};
-% \node[op] (op_5) [right of = ir_3] {~~~~~~~};
-
-% \path[->] 	
-% 			(op_1) 	edge node [above] {} (ir_1)
-% 			(ir_1) 	edge node [above] {} (op_2)
-% 			(op_2) 	edge node [above] {} (ir_2)
-% 			(ir_2) 	edge node [above] {} (op_4)
-% 			(op_3) 	edge node [above] {} (ir_3)
-% 			(ir_3) 	edge node [above] {} (op_4)
-% 			(ir_3) 	edge node [above] {} (op_5)
-% 			;
-% \end{tikzpicture}}
-\caption{Flink's software stack.}
-\end{figure}

### 2015-11-25 exec-model.tex
Incorporated latest suggestions from doc
More copy-pasting.

@@
 
-\section{Execution model}
+\section{The Common Fabric: Streaming Dataflows}
 \label{sec:execution}
-\para{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
+
+Although users can write Flink programs using a multitude of APIs, all Flink programs eventually compile down to a common representation: the dataflow graph\footnote{Internally called JobGraph}.
+The dataflow graph is executed by Flink's runtime engine, the common layer underneath both the batch processing (DataSet) and stream processing (DataStream) APIs.
+
+\subsection{Dataflow Graphs}
+The dataflow graph, as depicted in \autoref{fig:dataflow}, is a directed acyclic graph (DAG) that consists of (i) stateful operators, and (ii) data streams that represent data produced by an operator and are available for consumption by operators.
+Since dataflow graphs are executed in a data-parallel fashion, operators are parallelized into one or more parallel instances called \emph{subtasks} and streams are split into one or more \emph{stream partitions} (one partition per producing subtask).
+
+\begin{figure}
+\centering
+\includegraphics[width=.7\textwidth]{figs/dataflow}
+\vspace{-3mm}
+\caption{A simple dataflow graph comprising stateful operators (sources, sinks and operators), and intermediate streams (transient and materialized).}
+\label{fig:dataflow}
+\end{figure}
+
+The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions, etc.
+Many of these operators are implementations of textbook versions of well known algorithms.
+\autoref{sec:streaming} provides details on the implementation of windowing operators.
+Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, or merge.
+
+\subsection{Data Exchange through Intermediate Data Streams}
+Flink's intermediate data streams are the core abstraction for data exchange between operators.
+The notion of intemediate data streams in Flink is broader than the traditional notion of a continuous transient stream.
+The particular behavior of a data stream is parameterized by the higher layers in Flink (e.g., the program optimizer used by the DataSet API).
+
+\para{Pipelined Data Exchange.} Pipelined streams exchange data between concurrently running producers and consumers resulting in pipelined execution.
+Synchronous streams propagate back pressure from consumer to producer (modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations).
+Flink uses these streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.
+
+\para{Blocking Data Exchange.} Blocking streams are applicable to bounded data streams.
+The decoupled stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages.
+Blocking streams naturally require more memory, frequently spill to secondary storage, and do not propagate backpressure.
+They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as joins, can cause distributed deadlocks.
+
+\para{Balancing Latency and Throughput.} Flink’s data exchange mechanisms are implemented around the exchange of buffers.
+When a data record is ready on the producer side, it is serialized and stored into one or more buffers that can be forwarded to consumers.
+A buffer in Flink is sent to a consumer either i) as soon as it is full, or ii) when a timeout condition is reached.
+This enables Flink to achieve, high throughput by setting the size of buffers to a high value, e.g., a few kilobytes, as well as low latency by setting the buffer timeout to a low value e.g., a few milliseconds.
+
+\para{Ordering Guarantees.} Data exchange of one stream partition preserves the order of records.
+As a consequence, chains of operators that consume one stream partition maintain order of records.
+Operators receiving more than one stream partition merge the streams as records arrive, to minimize back pressure.
+As a consequence, the streaming dataflow gives no ordering guarantees for records after any form of repartitioning or broadcasting.
+The responsibility to deal with out-of-order records is delegated to the operator implementation.
+We found that this gives the most efficient design, as most operators do not require deterministic order, and operators that need to compensate for out-of-order arrivals (such as event time windows) can do that efficiently.
+
+\para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events.
+These are special events injected in the data stream by producing operators and are delivered in-order along with all other data records and events within a stream partition.
+The receiving operators react to these events by performing certain actions upon their arrival.
+Flink uses control records in multiple mechanisms, such as:
+\begin{itemize}
+\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (Section XCY)
+\item To signal the progress of event time within a stream partition (Section XYZ). 
+\item Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows, to signal that a stream partition has reached the end of a superstep.
+\end{itemize}
+
+\subsection{Iterative Dataflows}
+{\color{red} Should we have a few sentences here?
+Would describe that cyclic flows enable iterative processing as BSP/SSP in batch [iterations paper, euranova SSP on Flink paper] and simply for streaming programs with feedback.}
+
+Cycles are formed by special feedback streams.
+Feedback streams do not act as a dependency (not treated by the scheduler, so Flink keeps simple DAG scheduling techniques) and their contents is treated as operator state (simplifies fault tolerance).
+
+\subsection{Fault Tolerance}
+Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution.
+The general assumption the system makes to effectively provide these guarantees is that the data sources are persistent and replayable.
+Examples of such sources are files and durable message queues (e.g.
+Apache Kafka).
+In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators.
+We provide more details about the approach we adopted to achieve fault tolerance for the DataSet and DataStream models respectively in sections 4.3 and 5.
+
+% \para{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
 There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs) as presented on Figure \autoref{fig:JobGraph}.
@@
 
-\para{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
+% \para{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
 Buffers contain one or more records, and a record can span multiple buffers.
@@
 
-We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
+% We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
 Internally, intermediate results are abstract classes with many implementations.
@@
 
-\para{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
+% \para{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
 Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
@@
 
-\para{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
+% \para{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
 To do that, the system has to materialize intermediate results (in memory or disk).

### 2015-11-25 streaming.tex
Figures and listings update
More copy-pasting.

@@
 
-Flink's DataStream API implements a full stream analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
+Flink’s DataStream API implements a full stream analytics framework on top of Flink’s runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
 The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
-Since the Flink runtime already has pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.The programming model of Flink's streaming API,  builds on two basic abstract data types, namely data streams and window streams.
-Each of the two supports specific operations since they exhibit different properties.
-For example, windows support transformations that are only possible on bounded collections such as joins, groupings and aggregations.
-Other transformations, such as (flat)map are  can be applied incrementally on data streams.
+Since Flink’s runtime already supports pipelined data transfers, continuous stateful operators, and a fault tolerance mechanism for consistent state updates, overlaying a stream processor on top of it, essentially boils down to implementing a windowing system and a state interface.
+As noted, these are invisible to the runtime, which sees windows as just an implementation of stateful operators.
 
-% Give an example here instead of simplicity
-The API is designed with simplicity in mind, while at the same time providing powerful tools to deal with time and uncertainty.
-Here is, for example, a word count program on a simple  time-based window:
+\subsection{The Notion of Time}
+Flink distinguishes between two notions of time: i) event time, that  stands for the time that an event was generated at its origin (e.g., the time that a sensor or a mobile device generated a signal) and ii) processing time, that is the wall-clock time of the machine that is processing the data.
 
-\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}
+In distributed systems there is an arbitrary skew between event-time and processing-time \cite{akidau2015dataflow}.
+This may mean arbitrary delays for getting an answer based on event-time semantics.
+To avoid arbitrary delays these systems regularly insert special events called ‘low watermarks’ that mark a global progress measure.
+In the case of time progress for example, that includes a time attribute ‘t’ indicating that all events lower than ‘t’ have already entered an operator.
+The watermarks aid the execution engine to process events in the correct event order and serialize operations (such as window computations) via a unified measure of progress.
 
-Flink distinguishes between two notions of time: 
-\begin{description}
-\item[Event time] is the time that an event actually happened (e.g., the time that a sensor emitted a signal, or the time that a person tapped on their smartphone).
-\item[Processing time] is the wall-clock time of the machine that is processing the data.
-\end{description}
+Watermarks originate at the sources of the topology, where we can determine the time inherent in future elements accurately.
+The watermarks propagate from the sources throughout the other operators of the data flow.
+Operators decide how they react to watermarks.
+Simple operations such as map/filter just forward the watermarks they receive, while more complex operators that do calculation based on the watermarks (e.g., event time windows) first compute results triggered by a watermark and then forward it.
+If an operation has more than one input, the system only forwards the minimum of the incoming watermarks to the operator, ensuring correct results.
 
-In distributed systems, there is an arbitrary lag between event-time and processing-time \cite{akidau2015dataflow}.
-This may mean arbitrary delays for getting an answer based on event-time semantics.
-To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system.
-The watermarks aid the execution engine to process events in the correct event order.
+Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit lower latency.
+Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
+Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter Flink.
+This provides  lower processing latency than event time, and more accurate results compared to processing time.
 
-Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit the best latency.
-Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
-Flink includes a third notion of time as a special case of event time called ingestion time, which is the time that events enter the system.
-This provides  a lower latency than event-time semantics, and avoids the arbitrary processing time semantics.
+\subsection{Stream Discretization}
+Many incremental computations over unbounded streams are often evaluated over finite logical views, called windows.
+Apache Flink creates windows via a highly flexible discretization process which is composed out of three core functions: a window assigner and optionally a trigger and an evictor.
+All three functions can be selected among a pool of common predefined implementations (e.g.
+sliding time window) or explicitly defined by the user (via UDFs).
 
-\para{Streaming Windows.} A window defines a logical group of records that are processed together.
-Flink's windowing system largely follows the Dataflow model~\cite{akidau2015dataflow} proposed by Google.
-A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor.
-The assigner predefines the logical groups to which each record belongs, e.g.
-count-based or time-based periodic windows.
-The trigger defines when the window operation is performed on each group.
-The evictor defines which records to keep in each group.
-The following example defines windows that operate in event time of 6 seconds that slide every 2 seconds (Assigner).
+More specifically, the assigner is responsible for assigning each record to logical windows.
+For example, this decision can be based on the timestamp of a record when it comes to event-time discretization.
+Note that in the case of sliding windows, an element can belong to multiple logical windows.
+An optional trigger defines when the operation associated with the window definition is performed.
+Finally, an optional evictor determines which records to retain within each window.
+The discretization operation of Apache Flink is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows.
+Note that windows incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsumes these discretization models.
+For example, below is a window definition of 6 seconds that slides every 2 seconds (Assigner).
 The window results are computed once the watermark passes the end of the window (Trigger).
-Flink also offer short hands for commonly used functionalities.
 
-\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}
+\begin{lstlisting}[language=Scala]
+stream
+  .window(SlidingTimeWindows.of(Time.of(6, SECONDS), Time.of(2, SECONDS))
+  .trigger(EventTimeTrigger.create())
+\end{lstlisting}
 
@@
 
-\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
+\begin{lstlisting}[language=Scala]
+stream
+  .window(GlobalWindow.create())
+  .trigger(Count.of(1000))
+  .evict(Count.of(100))
+\end{lstlisting}
 
 Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality~\cite{akidau2015dataflow}, and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).
+This mechanism can be used to implement a wide variety of windowing functionality [Dataflow].
 
-Simple programs in Flink's DataStream API looks like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-We incorporate mutable state in the API by providing interfaces to users to register any local variable within a transformation with the system's checkpointed mechanism and freely use this variable in their code.
+\subsection{Stateful Stream Processing}
+Simple programs in Flink’s DataStream API look like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
+However, to cover the need for stateful operations, we incorporate explicit mutable state in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partionable dynamic state and its operations.
+Flink’s checkpointing mechanism guarantees that any registered state is durably accessed exactly-once per operation via checkpointing.
+
+As an example consider a stream wordNums containing words and an associated number for each word.
+We can use the partitionable  OperatorState in a stateful map function to compute a rolling average number seen per word by marking a sum and count per word as explicit states along with their updates:
+
+\begin{lstlisting}[language=Scala]
+wordNums.keyBy(0).map(new StatefulAverage())
+
+public class StatefulAverage implements RichMapFunction<Long, Double> {
+    private OperatorState<Long> sum;
+    private OperatorState<Long> count;
+
+    public Double reduce(Tuple2<String,Long> num) {
+        count.update(count.value() + 1);
+        sum.update(sum.value() + num.f1)
+        return new Tuple2(num.f0, (double) sum.value()/count.value());
+    }
+...
+}
+\end{lstlisting}
+
+\subsection{Fault Tolerance via Asynchronous Snapshots}
+\label{sec:fault-tolerance}
+The basic fault tolerance mechanism of Apache Flink for unbounded streams builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees.
+The possibly unbounded nature of a data stream in the model of the DataStream API makes recomputation upon recovery impractical, as possibly months of computation will need to be replayed considering a long running job.
+To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.
+
+The core challenge lies in taking a consistent snapshot of all parallel operators without halting the execution of the topology.
+In essence, the snapshot of all operators should refer to the same logical time in the computation.
+The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
+Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
+
+\begin{figure}[ht]
+	\centering  	
+  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
+  	\vspace{-4mm}
+	\caption{Asynchronous Barrier Snapshotting.}
+	\vspace{-2mm}
+	\label{fig:FlinkStack}
+\end{figure}
+
+An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received.
+Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS).
+Once the state has been backed up, the operator forwards the barrier downstream.
+ABS bears resemblances to the seminal Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
+However, because of the DAG structure of a Flink program, ABS does not need to checkpoint in-flight records but solely rely on the aligning phase to apply all their effects to the operator states.
+This guarantees that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).
+
+Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier.
+The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers.
+Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.
+
+\vspace{1mm}
+\noindent ABS provides several benefits:\vspace{-3mm}
+\begin{itemize}
+\item It guarantees exactly-once state updates without ever pausing the computation \vspace{-3mm}
+\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval.
+\vspace{-3mm}
+\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
+\end{itemize}

### 2015-11-25 batch.tex
Reflects the google doc. Further changes will be done in latex.

@@
 However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and iii) programs that process bounded data sets are amenable to additional optimizations, more efficient book keeping for fault tolerance, and staged scheduling.
-.
 
 \vspace{2mm}
-\noindent Flink approaches batch processing as follows:
+\noindent Flink approaches batch processing as follows:\vspace{-3mm}
 \begin{itemize}
@@
 The runtime executable may be parameterized with blocked data streams to break up large computations in isolated stages that are scheduled successively.
-	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).
+\vspace{-3mm}
+	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).\vspace{-3mm}
 	\item Blocking operators (e.g., joins and sorts) are simply operator implementations that happen to block until they have consumed their entire input.
 The runtime is not aware of whether an operator is blocking or not.
-These operators use managed memory provided by Flink (either on- or off- the JVM heap), and can spill to disk if their inputs exceed their memory bounds.
-	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc)
-	\item A query optimization layer transforms a DataSet program to an efficient executable.
+These operators use managed memory provided by Flink (either on- or off- the JVM heap), and can spill to disk if their inputs exceed their memory bounds.\vspace{-3mm}
+	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc).\vspace{-3mm}
+	\item A query optimization layer transforms a DataSet program to an efficient executable.\vspace{-3mm}
 \end{itemize}
 
-\para{Query Optimization.} Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation.
+\subsection{Query Optimization}
+
+Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation.
 However, the arbitrary UDF-heavy DAGs that make up Flink’s dataflow programs do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer.
@@
 
-\para{Memory Management.} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into a  memory segments.
+\subsection{Memory Management} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into a  memory segments.
 Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed.
@@
 
-\para{Native Iterations on top of Dataflows.}  Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
+\subsection{Native Iterations on top of Dataflows}
+\label{sec:batch-iterations}
+Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
 The approach, implemented in Naiad \cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
@@
 Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
+
+\subsection{Fault Tolerance}
+\label{sec:batch-fault-tolerance}
+\TODO{We need to simply say that we use the blocking data exchange to materialize intermediate results and redeploy in a spark fashion.}

### 2015-11-25 exec-model.tex
Reflects the google doc. Further changes will be done in latex.

@@
 
-Although users can write Flink programs using a multitude of APIs, all Flink programs eventually compile down to a common representation: the dataflow graph\footnote{Internally called JobGraph}.
+Although users can write Flink programs using a multitude of APIs, all Flink programs eventually compile down to a common representation: the dataflow graph.
 The dataflow graph is executed by Flink's runtime engine, the common layer underneath both the batch processing (DataSet) and stream processing (DataStream) APIs.
@@
 Since dataflow graphs are executed in a data-parallel fashion, operators are parallelized into one or more parallel instances called \emph{subtasks} and streams are split into one or more \emph{stream partitions} (one partition per producing subtask).
+The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions, etc.
+Many of these operators are implementations of textbook versions of well known algorithms.
+\autoref{sec:streaming} provides details on the implementation of windowing operators.
+Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, or merge.
 
-\begin{figure}
+\begin{figure}[t!]
 \centering
@@
 
-The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions, etc.
-Many of these operators are implementations of textbook versions of well known algorithms.
-\autoref{sec:streaming} provides details on the implementation of windowing operators.
-Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, or merge.
+\subsection{Data Exchange through Intermediate Data Streams}
+\TODO{Asterios does not like the structure of this subsection. Minimal changes are needed but will be useful. for now it's a soup of ideas. Good news is, the content is already there.}
 
-\subsection{Data Exchange through Intermediate Data Streams}
 Flink's intermediate data streams are the core abstraction for data exchange between operators.
@@
 
-\para{Pipelined Data Exchange.} Pipelined streams exchange data between concurrently running producers and consumers resulting in pipelined execution.
+\para{Pipelined and Blocking Data Exchange.} \emph{Pipelined streams} exchange data between concurrently running producers and consumers resulting in pipelined execution.
 Synchronous streams propagate back pressure from consumer to producer (modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations).
 Flink uses these streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.
-
-\para{Blocking Data Exchange.} Blocking streams are applicable to bounded data streams.
+\emph{Blocking streams}, on the other hand, are applicable to bounded data streams.
 The decoupled stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages.
@@
 This enables Flink to achieve, high throughput by setting the size of buffers to a high value, e.g., a few kilobytes, as well as low latency by setting the buffer timeout to a low value e.g., a few milliseconds.
+
+\noindent \TODO{Do we need the one below? We will need more context if we decide to keep this here. Maybe move to streams?}
 
@@
 The receiving operators react to these events by performing certain actions upon their arrival.
-Flink uses control records in multiple mechanisms, such as:
+Flink uses control records in multiple mechanisms, such as: \vspace{-2mm}
 \begin{itemize}
-\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (Section XCY)
-\item To signal the progress of event time within a stream partition (Section XYZ). 
-\item Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows, to signal that a stream partition has reached the end of a superstep.
+\item Signaling the progress of event time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
+\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
+\item Signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}).
+\vspace{-3mm}
 \end{itemize}
@@
 \subsection{Iterative Dataflows}
-{\color{red} Should we have a few sentences here?
-Would describe that cyclic flows enable iterative processing as BSP/SSP in batch [iterations paper, euranova SSP on Flink paper] and simply for streaming programs with feedback.}
+\TODO{Should we have a few sentences here?
+Would describe that cyclic flows enable iterative processing as BSP/SSP in batch [iterations paper, euranova SSP on Flink paper] and simply for streaming programs with feedback.
+@Stephan, could you make a start here?
+Just to see how you would like to see it.}
 
@@
 In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators.
-We provide more details about the approach we adopted to achieve fault tolerance for the DataSet and DataStream models respectively in sections 4.3 and 5.
+We provide more details about the approach we adopted to achieve fault tolerance for DataStreams and DataSets in sections \ref{sec:fault-tolerance} and \ref{sec:batch-fault-tolerance} respectively.
 

### 2015-11-25 intro.tex
Reflects the google doc. Further changes will be done in latex.

@@
 The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning.
-Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing~\cite{CUSTOM:web/FlinkContrib}), and is used in production in several companies.
+Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing\footnote{For an update list, refer to \url{https://github.com/apache/flink/graphs/contributors}}), and is used in production in several companies.
 
-\noindent The contributions of this paper are the following:
+\vspace{2mm}
+\noindent The contributions of this paper are the following:\vspace{-2mm}
 \begin{itemize}
@@
 	\vspace{-3mm}
-	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault tolerant streaming dataflows (\autoref{sec:snapshots}).
+	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault tolerant streaming dataflows (\autoref{sec:execution}).
 	\vspace{-3mm}
-	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:windows}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
+	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:streaming}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
 \end{itemize}

### 2015-11-25 related.tex
Reflects the google doc. Further changes will be done in latex.

@@
 %!TEX root = paper.tex
-
+\vspace{-4mm}
 \section{Related work}
+\vspace{-2mm}
 \label{sec:related}
@@
 
-\para{Batch Processing.} Apache Hadoop\footnote{https://hadoop.apache.org/} is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
+\para{Batch Processing.} Apache Hadoop is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
 Dryad~\cite{isard2007dryad}, introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer} which added a language and an SQL optimizer on top of it.
-Apache Tez\footnote{https://tez.apache.org/} can be seen as an open source implementation of the ideas proposed in Dryad.
+Apache Tez~\cite{saha2015apache} can be seen as an open source implementation of the ideas proposed in Dryad.
 MPP databases \cite{dewitt1990gamma}, and recent open-source implementations~\cite{CUSTOM:web/Drill,kornacker2015impala} restrict their API to SQL variants.
-Similar to Flink, Apache Spark\footnote{spark.apache.org} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+Similar to Flink, Apache Spark \cite{DBLP:conf/hotcloud/ZahariaCFSS10} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
 In contrast, Flink is the only system that
@@
 \end{inparaenum}
-Newer open source streaming systems that scale horizontally, such as Apache Storm\footnote{http://storm.apache.org/} and Apache Samza\footnote{http://samza.apache.org/}, provide low level APIs and offer only at-least-once and at-most-once guarantees.
+Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza\footnote{Those can be found under \url{http://storm.apache.org/} and \url{http://samza.apache.org/} respectively.}, provide low level APIs and offer only at-least-once and at-most-once guarantees.
 MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics.

### 2015-11-25 streaming.tex
Reflects the google doc. Further changes will be done in latex.

@@
 \section{Stream Analytics on Top of Dataflows}
-\label{sec:windows}
+\label{sec:streaming}
 
@@
 \subsection{The Notion of Time}
+\label{sec:streaming-time}
 Flink distinguishes between two notions of time: i) event time, that  stands for the time that an event was generated at its origin (e.g., the time that a sensor or a mobile device generated a signal) and ii) processing time, that is the wall-clock time of the machine that is processing the data.
@@
 
-Watermarks originate at the sources of the topology, where we can determine the time inherent in future elements accurately.
+Watermarks originate at the sources of a topology, where we can determine the time inherent in future elements.
 The watermarks propagate from the sources throughout the other operators of the data flow.
@@
 
-\begin{lstlisting}[language=Scala]
+\begin{lstlisting}[language=Java]
 stream
@@
 Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
-This mechanism can be used to implement a wide variety of windowing functionality [Dataflow].
+This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
 
 \subsection{Stateful Stream Processing}
+\TODO{Asterios: This is very developer-heavy.
+We need to show the big picture.
+Long story short, the side-effect free functional view of the system changes with mutable state that is managed by the system.
+Why do we need state?
+What can be state?
+How is it implemented in Flink.
+@Paris, please answer these 3 questions in 2-3 paragraphs.
+The example below could also be missing.}
+
 Simple programs in Flink’s DataStream API look like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
@@
 
-\begin{lstlisting}[language=Scala]
+\begin{lstlisting}[language=Java]
 wordNums.keyBy(0).map(new StatefulAverage())
@@
 \label{sec:fault-tolerance}
-The basic fault tolerance mechanism of Apache Flink for unbounded streams builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees.
+The fault tolerance mechanism of Apache Flink for unbounded streams builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees.
 The possibly unbounded nature of a data stream in the model of the DataStream API makes recomputation upon recovery impractical, as possibly months of computation will need to be replayed considering a long running job.
@@
   	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
-  	\vspace{-4mm}
+  	\vspace{-6mm}
 	\caption{Asynchronous Barrier Snapshotting.}
 	\vspace{-2mm}
-	\label{fig:FlinkStack}
+	\label{fig:snapshots}
 \end{figure}

### 2015-11-25 src/genwin.scala
deleted useless source files

@@ -1,11 +1,0 @@
-//general window API
-stream.window(SlidingTimeWindows.of(Time.seconds(6), Time.seconds(2))
-  .trigger(TimeTrigger.create())
-
-//equivalent short hand for common use case
-stream.timeWindow(Time.seconds(6), Time.seconds(2))
-
-//global window
-stream.window(GlobalWindow.create)
-  .trigger(Count.of(1000))
-  .evict(Count.of(100))

### 2015-11-25 src/glowin.scala

@@ -1,4 +1,0 @@
-stream
-  .window(GlobalWindow.create)
-  .trigger(Count.of(1000))
-  .evict(Count.of(100))

### 2015-11-25 src/wwc.scala

@@ -1,19 +1,0 @@
-object WindowWordCount {
-  def main(args: Array[String]) {
-
-    val env = StreamExecutionEnvironment.getExecutionEnvironment
-    val text = env.socketTextStream("localhost", 9999)
-
-    val counts = text
-	.flatMap {line => line.toLowerCase.split("\\W+")
-				.filter { w => w.nonEmpty }}
-    .map {w =>  (w, 1) }
-    .keyBy(0)
-    .timeWindow(Time.of(5, TimeUnit.SECONDS))
-    .sum(1)
-
-    counts.print
-    env.execute("Window Stream WordCount")
-  }
-}
-

### 2015-11-26 streaming.tex
Updates from ShareLaTeX

--- 2015-11-25 streaming.tex
@@
 \subsection{Stateful Stream Processing}
-Simple programs in Flink’s DataStream API look like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-However, to cover the need for stateful operations, we incorporate explicit mutable state in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partionable dynamic state and its operations.
-Flink’s checkpointing mechanism guarantees that any registered state is durably accessed exactly-once per operation via checkpointing.
+While most operators in Flink’s DataStream API look like functional, side-effect free operators, they also support efficient stateful computations.
+Explicit mutable state is incorporated in the Flink DataStream API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
+Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system allowing very flexible custom state management in the streaming applications.
+Flink’s checkpointing mechanism guarantees that any registered state is durably accessed with exactly-once update semantics.
 
-As an example consider a stream wordNums containing words and an associated number for each word.
-We can use the partitionable  OperatorState in a stateful map function to compute a rolling average number seen per word by marking a sum and count per word as explicit states along with their updates:
+As a simple example consider a map operator that computes a continuous average for a given stocks' price.
+For this we need to keep a state with the current count and total sum for each stock in order to compute the average:
 
 \begin{lstlisting}[language=Scala]
-wordNums.keyBy(0).map(new StatefulAverage())
+val priceStream: DataStream[(String,Double)] = ...
  
-public class StatefulAverage implements RichMapFunction<Long, Double> {
-    private OperatorState<Long> sum;
-    private OperatorState<Long> count;
-
-    public Double reduce(Tuple2<String,Long> num) {
-        count.update(count.value() + 1);
-        sum.update(sum.value() + num.f1)
-        return new Tuple2(num.f0, (double) sum.value()/count.value());
-    }
-...
+val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
+    (in, state: Option[(Int, Double)]) => {
+		val (count, sum) = state.getOrElse((0,0.0))
+		val newState = (count + 1, sum + in._2)
+		((in._1, newState._2/newState._1), newState)
 	}

### 2015-11-26 streaming.tex
Merge with Paris' changes

--- 2015-11-25 streaming.tex
@@
 
-As an example consider a stream wordNums containing words and an associated number for each word.
-We can use the partitionable  OperatorState in a stateful map function to compute a rolling average number seen per word by marking a sum and count per word as explicit states along with their updates:
+As a simple example consider a map operator that computes a continuous average for a given stocks price.
+For this we need to keep a state with the current count and total sum for each stock in order to compute the average:
 
-\begin{lstlisting}[language=Java]
-wordNums.keyBy(0).map(new StatefulAverage())
+\begin{lstlisting}[language=Scala]
+val priceStream: DataStream[(String,Double)] = ...
  
-public class StatefulAverage implements RichMapFunction<Long, Double> {
-    private OperatorState<Long> sum;
-    private OperatorState<Long> count;
-
-    public Double reduce(Tuple2<String,Long> num) {
-        count.update(count.value() + 1);
-        sum.update(sum.value() + num.f1)
-        return new Tuple2(num.f0, (double) sum.value()/count.value());
-    }
-...
+val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
+    (in, state: Option[(Int, Double)]) => {
+		val (count, sum) = state.getOrElse((0,0.0))
+		val newState = (count + 1, sum + in._2)
+		((in._1, newState._2/newState._1), newState)
 	}

### 2015-11-26 exec-model.tex
general restructure proposed by stephan

@@
 \subsection{Fault Tolerance}
+
+\label{sec:fault-tolerance}
 Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution.
@@
 In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators.
-We provide more details about the approach we adopted to achieve fault tolerance for DataStreams and DataSets in sections \ref{sec:fault-tolerance} and \ref{sec:batch-fault-tolerance} respectively.
+
+The checkpointing mechanism of Apache Flink builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees.
+The possibly unbounded nature of a data stream makes re-computation upon recovery impractical, as possibly months of computation will need to be replayed considering a long running job.
+To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.
+
+\begin{figure}[t!]
+	\centering  	
+  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
+  	\vspace{-6mm}
+	\caption{Asynchronous Barrier Snapshotting.}
+	\vspace{-2mm}
+	\label{fig:snapshots}
+\end{figure}
+
+The core challenge lies in taking a consistent snapshot of all parallel operators without halting the execution of the topology.
+In essence, the snapshot of all operators should refer to the same logical time in the computation.
+The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
+Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
+
+An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received.
+Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS).
+Once the state has been backed up, the operator forwards the barrier downstream.
+ABS bears resemblances to the seminal Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
+However, because of the DAG structure of a Flink program, ABS does not need to checkpoint in-flight records but solely rely on the aligning phase to apply all their effects to the operator states.
+This guarantees that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).
+
+Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier.
+The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers.
+Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.
+\vspace{1mm}
+\noindent ABS provides several benefits:\vspace{-3mm}
+\begin{itemize}
+\item It guarantees exactly-once state updates without ever pausing the computation \vspace{-3mm}
+\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval.
+\vspace{-3mm}
+\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
+\end{itemize}
 

### 2015-11-26 streaming.tex
general restructure proposed by stephan

@@
 
-\subsection{Stream Discretization}
+\subsection{Stateful Stream Processing}
+\TODO{Asterios: This is very developer-heavy.
+We need to show the big picture.
+Long story short, the side-effect free functional view of the system changes with mutable state that is managed by the system.
+Why do we need state?
+What can be state?
+How is it implemented in Flink.
+@Paris, please answer these 3 questions in 2-3 paragraphs.
+The example below could also be missing.}
+
+While most operators in Flink’s DataStream API look like functional, side-effect free operators, they also support efficient stageful computations.
+Explicit mutable state is incorporated in the Flink DataStream API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
+Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system allowing very flexible custom state management in the streaming applications.
+Flink’s checkpointing mechanism guarantees that any registered state is durably accessed with exactly-once update semantics.
+
+As a simple example consider a map operator that computes a continuous average for a given stocks price.
+For this we need to keep a state with the current count and total sum for each stock in order to compute the average:
+
+\begin{lstlisting}[language=Scala]
+val priceStream: DataStream[(String,Double)] = ...
+ 
+val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
+    (in, state: Option[(Int, Double)]) => {
+		val (count, sum) = state.getOrElse((0,0.0))
+		val newState = (count + 1, sum + in._2)
+		((in._1, newState._2/newState._1), newState)
+	}
+\end{lstlisting}
+
+\subsection{Stream Windows}
 Many incremental computations over unbounded streams are often evaluated over finite logical views, called windows.
-Apache Flink creates windows via a highly flexible discretization process which is composed out of three core functions: a window assigner and optionally a trigger and an evictor.
+Apache Flink creates windows via a highly flexible process which is composed out of three core functions: a window assigner and optionally a trigger and an evictor.
 All three functions can be selected among a pool of common predefined implementations (e.g.
@@
 More specifically, the assigner is responsible for assigning each record to logical windows.
-For example, this decision can be based on the timestamp of a record when it comes to event-time discretization.
+For example, this decision can be based on the timestamp of a record when it comes to event-time windows.
 Note that in the case of sliding windows, an element can belong to multiple logical windows.
@@
 Finally, an optional evictor determines which records to retain within each window.
-The discretization operation of Apache Flink is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows.
-Note that windows incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsumes these discretization models.
+The window assignment process of Apache Flink is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows.
+Note that windows incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsumes these windowing models.
 For example, below is a window definition of 6 seconds that slides every 2 seconds (Assigner).
@@
 This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
-
-\subsection{Stateful Stream Processing}
-\TODO{Asterios: This is very developer-heavy.
-We need to show the big picture.
-Long story short, the side-effect free functional view of the system changes with mutable state that is managed by the system.
-Why do we need state?
-What can be state?
-How is it implemented in Flink.
-@Paris, please answer these 3 questions in 2-3 paragraphs.
-The example below could also be missing.}
-
-Simple programs in Flink’s DataStream API look like functional, side effect-free programs consisting of transformations on unbounded immutable collections.
-However, to cover the need for stateful operations, we incorporate explicit mutable state in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partionable dynamic state and its operations.
-Flink’s checkpointing mechanism guarantees that any registered state is durably accessed exactly-once per operation via checkpointing.
-
-As a simple example consider a map operator that computes a continuous average for a given stocks price.
-For this we need to keep a state with the current count and total sum for each stock in order to compute the average:
-
-\begin{lstlisting}[language=Scala]
-val priceStream: DataStream[(String,Double)] = ...
- 
-val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
-    (in, state: Option[(Int, Double)]) => {
-		val (count, sum) = state.getOrElse((0,0.0))
-		val newState = (count + 1, sum + in._2)
-		((in._1, newState._2/newState._1), newState)
-	}
-\end{lstlisting}
-
-\subsection{Fault Tolerance via Asynchronous Snapshots}
-\label{sec:fault-tolerance}
-The fault tolerance mechanism of Apache Flink for unbounded streams builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees.
-The possibly unbounded nature of a data stream in the model of the DataStream API makes recomputation upon recovery impractical, as possibly months of computation will need to be replayed considering a long running job.
-To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.
-
-The core challenge lies in taking a consistent snapshot of all parallel operators without halting the execution of the topology.
-In essence, the snapshot of all operators should refer to the same logical time in the computation.
-The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
-Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
-
-\begin{figure}[ht]
-	\centering  	
-  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
-  	\vspace{-6mm}
-	\caption{Asynchronous Barrier Snapshotting.}
-	\vspace{-2mm}
-	\label{fig:snapshots}
-\end{figure}
-
-An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received.
-Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS).
-Once the state has been backed up, the operator forwards the barrier downstream.
-ABS bears resemblances to the seminal Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
-However, because of the DAG structure of a Flink program, ABS does not need to checkpoint in-flight records but solely rely on the aligning phase to apply all their effects to the operator states.
-This guarantees that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).
-
-Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier.
-The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers.
-Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.
-
-\vspace{1mm}
-\noindent ABS provides several benefits:\vspace{-3mm}
-\begin{itemize}
-\item It guarantees exactly-once state updates without ever pausing the computation \vspace{-3mm}
-\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval.
-\vspace{-3mm}
-\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
-\end{itemize}

### 2015-11-26 streaming.tex
incorporate changes in stateful stream processing and windowing

@@
 \subsection{Stateful Stream Processing}
-\TODO{Asterios: This is very developer-heavy.
-We need to show the big picture.
-Long story short, the side-effect free functional view of the system changes with mutable state that is managed by the system.
-Why do we need state?
-What can be state?
-How is it implemented in Flink.
-@Paris, please answer these 3 questions in 2-3 paragraphs.
-The example below could also be missing.}
+While most operators in Flink’s DataStream API look like functional, side-effect free operators, they also support efficient stateful computations.
+State is critical to many applications such as machine learning model building, graph analysis, user session handling and window aggregations among others.
+There is a plethora of different types of states depending on the use case.
+For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine learning applications.
+Stream windows can also be modelled, in principle, by stateful operators which assign records to continuously updated buckets kept in memory as part of an operator state.
 
-While most operators in Flink’s DataStream API look like functional, side-effect free operators, they also support efficient stageful computations.
-Explicit mutable state is incorporated in the Flink DataStream API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
+In Apache Flink, state is made explicit, to cover all different use cases, and incorporated in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
 Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system allowing very flexible custom state management in the streaming applications.
@@
 Many incremental computations over unbounded streams are often evaluated over finite logical views, called windows.
-Apache Flink creates windows via a highly flexible process which is composed out of three core functions: a window assigner and optionally a trigger and an evictor.
+Apache Flink incorporates windowing within a stateful operator which is configured via a flexible declaration composed out of three core functions: a window assigner and optionally a trigger and an evictor.
 All three functions can be selected among a pool of common predefined implementations (e.g.

### 2015-11-26 streaming.tex
windows are evolving views

@@
 \subsection{Stream Windows}
-Many incremental computations over unbounded streams are often evaluated over finite logical views, called windows.
+Many incremental computations over unbounded streams are often evaluated over continously evolving logical views, called windows.
 Apache Flink incorporates windowing within a stateful operator which is configured via a flexible declaration composed out of three core functions: a window assigner and optionally a trigger and an evictor.

### 2015-11-26 architecture.tex
Reflects the google doc. Further changes will be done in latex.
small changes

@@
 \section{System Architecture}
+\label{sec:architecture}
 
@@
 
-\para{Flink's Runtime and APIS.} Flink's software stack can be seen in \autoref{fig:FlinkStack}.
+\para{Flink's Runtime and APIS.} Flink's software stack can be seen in \autoref{fig:stack}.
 The core of Flink is the distributed dataflow engine, which executes dataflow programs.
@@
 As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing.
-On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML, Gelly, and Table).
+On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML (machine learning), Gelly (graph processing), and Table (SQL-like operations)).
 
-As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the JobManager (JM - master), and one or more TaskManagers (TM - workers).
+As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the Job Manager, and one or more Task Managers.
 The client takes the program code, transforms it to a dataflow graph, and submits that to the JobManager.
@@
 In a high-availability setup, the JobManager persists a minimal set of metadata at each checkpoint to a fault-tolerant storage, such that a standby JobManager can reconstruct the checkpoint and recover the dataflow execution from there (see \autoref{sec:fault-tolerance}).
-
 The actual data processing takes place in the TaskManagers.
@@
 
-\begin{figure}
+\begin{figure}[t!]
+\begin{minipage}{1.1\linewidth}
       \centering
-\begin{minipage}{0.49\textwidth}
+      \hspace{-0.1\linewidth}
+      \begin{minipage}{0.48\linewidth}
+          \begin{figure}[H]
               \centering
-\includegraphics[width=.88\textwidth]{figs/stack.pdf}
-\label{fig:FlinkStack}
+			\includegraphics[width=.88\textwidth]{figs/stack}
 			\caption{Flink's Software Stack.}
+			\label{fig:stack}
+          \end{figure}
       \end{minipage}
-\begin{minipage}{.49\textwidth}
-\centering
+      % \hspace{0.05\linewidth}
+      \begin{minipage}{0.48\linewidth}
+          \begin{figure}[H]
 				\includegraphics[width=.95\textwidth]{figs/architecture.pdf}
@@
     			\label{fig:process-model}
-\end{minipage}%
           \end{figure}
+      \end{minipage}
+  \end{minipage}
+\end{figure}

### 2015-11-26 exec-model.tex
small changes

@@
 
-\noindent \TODO{Do we need the one below? We will need more context if we decide to keep this here. Maybe move to streams?}
-
-\para{Ordering Guarantees.} Data exchange of one stream partition preserves the order of records.
-As a consequence, chains of operators that consume one stream partition maintain order of records.
-Operators receiving more than one stream partition merge the streams as records arrive, to minimize back pressure.
-As a consequence, the streaming dataflow gives no ordering guarantees for records after any form of repartitioning or broadcasting.
-The responsibility to deal with out-of-order records is delegated to the operator implementation.
-We found that this gives the most efficient design, as most operators do not require deterministic order, and operators that need to compensate for out-of-order arrivals (such as event time windows) can do that efficiently.
-
 \para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events.
@@
 The receiving operators react to these events by performing certain actions upon their arrival.
-Flink uses control records in multiple mechanisms, such as: \vspace{-2mm}
+Flink uses special types of control events such as: \vspace{-2mm}
 \begin{itemize}
-\item Signaling the progress of event time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
-\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
-\item Signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}).
-\vspace{-3mm}
+\item \textit{Checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
+\item \textit{Watermarks} signaling the progress of event time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
+\item \textit{Iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}).
+\vspace{-1mm}
 \end{itemize}
+
+Control events assume that a stream partition preserves the order of records.
+To this end, unary operators in Flink consuming a single stream partition, \emph{guarantee a FIFO order of records}.
+However, operators receiving more than one stream partition merge the streams as records arrive, in order to keep up with the streams' rate and avoid back pressure.
+As a result, streaming dataflows in Flink do not provide ordering guarantees after any form of repartitioning or broadcasting, and the responsibility of dealing with out-of-order records is left to the operator implementation.
+We found that this gives the most efficient design, as most operators do not require deterministic order (e.g., hash-joins, maps, etc.), and operators that need to compensate for out-of-order arrivals (such as event time windows) can do that more efficiently as part of the operator logic.
 
@@
 
+The core challenge lies in taking a consistent snapshot of all parallel operators without halting the execution of the topology.
+In essence, the snapshot of all operators should refer to the same logical time in the computation.
+The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
+Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
+
 \begin{figure}[t!]
@@
 \end{figure}
-
-The core challenge lies in taking a consistent snapshot of all parallel operators without halting the execution of the topology.
-In essence, the snapshot of all operators should refer to the same logical time in the computation.
-The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
-Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
 
@@
 Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.
+
 \vspace{1mm}

### 2015-11-26 streaming.tex
small changes

--- 2015-11-26 streaming.tex
@@
 \subsection{Stateful Stream Processing}
-While most operators in Flink’s DataStream API look like functional, side-effect free operators, they also support efficient stateful computations.
-State is critical to many applications such as machine learning model building, graph analysis, user session handling and window aggregations among others.
+While most operators in Flink’s DataStream API look like functional, side-effect free operators, they provide support for efficient stateful computations.
+State is critical to many applications such as machine learning model building, graph analysis, user session handling and window aggregations.
 There is a plethora of different types of states depending on the use case.
 For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine learning applications.
-Stream windows can also be modelled, in principle, by stateful operators which assign records to continuously updated buckets kept in memory as part of an operator state.
+Stream windows can also be modelled, in principle, by stateful operators which assign records to continuously updated buckets kept in memory as part of the operator state.
 
-In Apache Flink, state is made explicit, to cover all different use cases, and incorporated in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
-Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system allowing very flexible custom state management in the streaming applications.
-Flink’s checkpointing mechanism guarantees that any registered state is durably accessed with exactly-once update semantics.
+To cover a lot of different use cases, in Flink state is made explicit and is incorporated in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
+Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in the streaming applications.
+Flink’s checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
 
@@
 \subsection{Stream Windows}
-Many incremental computations over unbounded streams are often evaluated over finite logical views, called windows.
-Apache Flink incorporates windowing within a stateful operator which is configured via a flexible declaration composed out of three core functions: a window assigner and optionally a trigger and an evictor.
+Incremental computations over unbounded streams are often evaluated over continously evolving logical views, called windows.
+Apache Flink incorporates windowing within a stateful operator which is configured via a flexible declaration composed out of three core functions: a window \textit{assigner} and optionally a \textit{trigger} and an \textit{evictor}.
 All three functions can be selected among a pool of common predefined implementations (e.g.
-sliding time window) or explicitly defined by the user (via UDFs).
+sliding time windows) or can be explicitly defined by the user as UDFs.
 
@@
 Finally, an optional evictor determines which records to retain within each window.
-The window assignment process of Apache Flink is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows.
-Note that windows incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsumes these windowing models.
+Flink's window assignment process is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows.
+Note that Flink's windowing capabilities incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsume these windowing models.
 For example, below is a window definition of 6 seconds that slides every 2 seconds (Assigner).
@@
 
-Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines.
+Note that streams are already partitioned on a key before windowing, so the above is a \textit{local operator} that does not require coordination between machines.
 This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.

### 2015-11-26 exec-model.tex
New figure
exec model
better latency figure

@@
 \begin{figure}[t!]
+\begin{minipage}{1.1\linewidth}
       \centering
-\includegraphics[width=.7\textwidth]{figs/dataflow}
+       \hspace{-0.1\linewidth}
+      \begin{minipage}{0.55\linewidth}
+        \begin{figure}[H]
+        \centering
+        \includegraphics[width=.999\textwidth]{figs/dataflow}
         \vspace{-3mm}
-\caption{A simple dataflow graph comprising stateful operators (sources, sinks and operators), and intermediate streams (transient and materialized).}
+        \caption{A simple dataflow graph.}
         \label{fig:dataflow}
+        \end{figure}
+      \end{minipage}
+      \hspace{0.03\linewidth}
+      % \vspace{-3mm}
+      \begin{minipage}{0.32\linewidth}
+          \begin{figure}[H]
+				\includegraphics[width=.99\textwidth]{figs/latency-throughput.pdf}
+				\vspace{-1mm}
+    			\caption{The effect of buffer-timeout in latency and throughput.}
+    			\label{fig:latency-throughput}
+          \end{figure}
+      \end{minipage}
+  \end{minipage}
 \end{figure}
@@
 This enables Flink to achieve, high throughput by setting the size of buffers to a high value, e.g., a few kilobytes, as well as low latency by setting the buffer timeout to a low value e.g., a few milliseconds.
+\autoref{fig:latency-throughput} shows the effect of buffer-timeouts to the throughput and latency of delivering records in a simple streaming grep job on 30 machines (120 cores).
+Flink can achieve an observable 99-th percentile latency of 20 milliseconds.
+The corresponding throughput is 24,500 events per second per core.
+As we increase the buffer timeout, we see an increase in latency with an increase in throughput, until full throughput is reached (where buffers fill up faster than the timeout expiration).
+At a buffer timeout of 50 milliseconds, the system reaches full throughput of 750,000 events per second per core with a 99-th percentile latency of 50 milliseconds.
 

### 2015-11-26 batch.tex
general restructure proposed by stephan
Iterations

@@
 \vspace{-3mm}
-	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).\vspace{-3mm}
+	\item Periodic snapshotting is turned off when its overhead is high.
+Instead, fault recovery can be achieved by replaying the lost stream partitions from the latest materialized intermediate stream (possibly the source).\vspace{-3mm}
 	\item Blocking operators (e.g., joins and sorts) are simply operator implementations that happen to block until they have consumed their entire input.
@@
 
-\subsection{Native Iterations on top of Dataflows}
+\subsection{Batch Iterations on top of Dataflows}
 \label{sec:batch-iterations}
-Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program.
-The approach, implemented in Naiad \cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations.
-A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration \cite{low2012distributed}.
-
-Flink follows an approach that retains the DAG-based runtime, but allows for special “head” and “tail” tasks to signify the beginning and end of iteration, that exchange data via shared memory.
-This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}, and outperforms the driver-based approach.
-Flink supports delta iterations, which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink’s Graph API.
-Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.
-
-\subsection{Fault Tolerance}
-\label{sec:batch-fault-tolerance}
-\TODO{We need to simply say that we use the blocking data exchange to materialize intermediate results and redeploy in a spark fashion.}
+Most batch dataflow approaches execute iterations by submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}.
+Flink keeps the DAG-based runtime and scheduler, and allows for special ``head'' and ``tail'' tasks to signify the beginning and end of each iteration, that are connected with feedback edges.
+Flink's iterations are competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}.
+This is achieved through the concept of  \emph{delta} iterations, which exploit sparse computational dependencies, and are used as the basis for Gelly, Flink's Graph API.
+Finally, Flink offers Bulk Synchronous Parallel (BSP) as well as Stale Synchronous Parallel (SSP) iterations in its DataSet API.

### 2015-11-26 exec-model.tex
Iterations

@@
         \includegraphics[width=.999\textwidth]{figs/dataflow}
-        \vspace{-3mm}
+        \vspace{-5mm}
         \caption{A simple dataflow graph.}
@@
 				\includegraphics[width=.99\textwidth]{figs/latency-throughput.pdf}
-				\vspace{-1mm}
+				\vspace{-7mm}
     			\caption{The effect of buffer-timeout in latency and throughput.}
@@
 \subsection{Data Exchange through Intermediate Data Streams}
-\TODO{Asterios does not like the structure of this subsection. Minimal changes are needed but will be useful. for now it's a soup of ideas. Good news is, the content is already there.}
-
 Flink's intermediate data streams are the core abstraction for data exchange between operators.
-The notion of intemediate data streams in Flink is broader than the traditional notion of a continuous transient stream.
+An intermediate data stream represents a logical handle to the data that is produced by an operator and can be consumed by one or more operators.
+Intermediate streams are logical in the sense that the data they point to may or may not be materialized on disk.
 The particular behavior of a data stream is parameterized by the higher layers in Flink (e.g., the program optimizer used by the DataSet API).
 
-\para{Pipelined and Blocking Data Exchange.} \emph{Pipelined streams} exchange data between concurrently running producers and consumers resulting in pipelined execution.
-Synchronous streams propagate back pressure from consumer to producer (modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations).
-Flink uses these streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.
+\para{Pipelined and Blocking Data Exchange.} \emph{Pipelined intermediate streams} exchange data between concurrently running producers and consumers resulting in pipelined execution.
+As a result, pipelined streams propagate back pressure from consumers to producers, modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations.
+Flink uses pipelined streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.
 \emph{Blocking streams}, on the other hand, are applicable to bounded data streams.
-The decoupled stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages.
+A blocking stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages.
 Blocking streams naturally require more memory, frequently spill to secondary storage, and do not propagate backpressure.
-They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as joins, can cause distributed deadlocks.
+They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as joins, may cause distributed deadlocks.
 
 \para{Balancing Latency and Throughput.} Flink’s data exchange mechanisms are implemented around the exchange of buffers.
-When a data record is ready on the producer side, it is serialized and stored into one or more buffers that can be forwarded to consumers.
-A buffer in Flink is sent to a consumer either i) as soon as it is full, or ii) when a timeout condition is reached.
-This enables Flink to achieve, high throughput by setting the size of buffers to a high value, e.g., a few kilobytes, as well as low latency by setting the buffer timeout to a low value e.g., a few milliseconds.
+When a data record is ready on the producer side, it is serialized and split into one or more buffers that can be forwarded to consumers.
+A buffer is sent to a consumer either i) as soon as it is full, or ii) when a timeout condition is reached.
+This enables Flink to achieve high throughput by setting the size of buffers to a high value (e.g., a few kilobytes), as well as low latency by setting the buffer timeout to a low value (e.g., a few milliseconds).
 \autoref{fig:latency-throughput} shows the effect of buffer-timeouts to the throughput and latency of delivering records in a simple streaming grep job on 30 machines (120 cores).
@@
 \para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events.
-These are special events injected in the data stream by producing operators and are delivered in-order along with all other data records and events within a stream partition.
+These are special events injected in the data stream by operators, and are delivered in-order along with all other data records and events within a stream partition.
 The receiving operators react to these events by performing certain actions upon their arrival.
-Flink uses special types of control events such as: \vspace{-2mm}
+Flink uses lots of special types of control events, including: \vspace{-2mm}
 \begin{itemize}
@@
 
-Control events assume that a stream partition preserves the order of records.
+As mentioned above, control events assume that a stream partition preserves the order of records.
 To this end, unary operators in Flink consuming a single stream partition, \emph{guarantee a FIFO order of records}.
@@
 
-\subsection{Iterative Dataflows}
-\TODO{Should we have a few sentences here?
-Would describe that cyclic flows enable iterative processing as BSP/SSP in batch [iterations paper, euranova SSP on Flink paper] and simply for streaming programs with feedback.
-@Stephan, could you make a start here?
-Just to see how you would like to see it.}
+\subsection{Fault Tolerance}
+\label{sec:fault-tolerance}
 
-Cycles are formed by special feedback streams.
-Feedback streams do not act as a dependency (not treated by the scheduler, so Flink keeps simple DAG scheduling techniques) and their contents is treated as operator state (simplifies fault tolerance).
-
-\subsection{Fault Tolerance}
-
-\label{sec:fault-tolerance}
 Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution.
@@
 \vspace{1mm}
-\noindent ABS provides several benefits:\vspace{-3mm}
+\noindent ABS provides several benefits:\vspace{-2mm}
 \begin{itemize}
@@
 \end{itemize}
-
-% \para{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges.
-There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs) as presented on Figure \autoref{fig:JobGraph}.
-For example, the aforementioned graph consists of five operators (circles), and three intermediate results (cylinders).
-Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g.
-reading data from a file system, a socket, a message queue, etc.), and data sinks.
-Operators produce intermediate results, as well as updates to state.
-An intermediate result is a logical handle (pointer) to the data that is produced by one operator.
-An intermediate result can be consumed by one or more operators.
-Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.
-When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph that consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs).
-
-% \para{Data Transfer.} The unit of data transfer in the Flink runtime is a buffer.
-Buffers contain one or more records, and a record can span multiple buffers.
-Buffers are requested and relinquished from local buffer pools, shared among operators that live in the same task manager.
-An IRP is simply a collection of buffers.
-Work in Flink progresses (i.e., records flow through the pipeline) as long as there are buffers available, essentially implementing distributed blocking queues (the logical streams) with bounded capacity (the amount of memory available to the buffer pools, which can be configured by the user).
-This mechanism, in addition to implementing record network transfers doubles down as a natural way to backpressure the flow in the case of slow operators (including external systems that consume data).
-
-% We mentioned that the intermediate results are logical handles to the data, rather than the data itself.
-Internally, intermediate results are abstract classes with many implementations.
-These implementations can perform pipelined data exchange, and blocking data exchange.
-
-% \para{Pipelined Data Exchange.} Pipelining (also called intra-operator parallelism), means that a producing and a consuming operator make progress at the same time, without the consumer waiting for the producer to finish.
-Pipelining is required in streaming systems, and is also used in batch systems to reduce latency.
-Flink implements pipelining by implementing intermediate results which activate network transfers between the producer task and consumer task,  as soon as their first buffer is available.
-Flink allows the configuration of its buffers' size and timeout.
-A buffer is sent to its consumer task as soon it is filled,  or as soon as a timeout is reached.
-Hence, by setting the buffer size and/or timeout to lower values, one can reduce latency, while achieving the opposite effect (i.e., high throughput) by using higher values.
-
-% \para{Blocking Data Exchange.} Sometimes it is desirable to break a job into stages, scheduling and executing each stage individually (e.g., to enable interactive processing, and better staging of resources in large batch jobs).
-To do that, the system has to materialize intermediate results (in memory or disk).
-Flink implements blocking data exchange via an intermediate result that signals its availability only when all the buffers from the producer have been materialized.
-The cached buffers can double down as a materialized reusable intermediate result.

### 2015-11-26 paper.tex
Added a couple of citations
Figures and listings update
Incorporated latest suggestions from doc
More copy-pasting.
Reflects the google doc. Further changes will be done in latex.
Iterations

@@
 \documentclass[11pt]{article}
-
-% A hack to make deauthor not complain which redefines
-% the titlepage command
-% \newcommand{\titlepage}{}
 
@@
 
-% JobGraph and Execution figs
+% JobGraph fig
 \usepackage{tikz}
-\usetikzlibrary{arrows,automata,calc,positioning,shapes.geometric}
+\usetikzlibrary{arrows,automata,positioning,shapes.geometric}
 
@@
 \usepackage{textcomp}
+
+\usepackage{float}
+
+\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
 
@@
 % there is no built in support or Scala yet, good enough
-\lstset{frame=tb,
+\lstset{frame=l,
   language=Java,
@@
   showstringspaces=false,
+  xleftmargin=5pt,
+%   framexleftmargin=-1pt,
   columns=flexible,
@@
   emph={%  
-    object, def, val, zip%
-    },emphstyle={\color{blue}}%
+    object, def, val, zip, window, trigger, evict%
+    },emphstyle={\color{blue}\textbf}%
 }%
@@
 
+\usepackage{subfigure}
+
+\def\definitionautorefname{Definition}
+\def\sectionautorefname{Section}
+\def\subsectionautorefname{Section}
+\def\subsubsectionautorefname{Section}
+\def\algorithmautorefname{Algorithm}
+\def\figureautorefname{Figure}
+
 \begin{document}
 
-\title{Apache Flink\texttrademark : Streaming Dataflows as a Basis \\ for Universal Data Analytics}
+\title{Apache Flink\texttrademark: Unified Stream and Batch Processing \\ in a Single Engine}
+
 \author{Authors}
@@
 \input{exec-model}
-\input{snapshots}
 \input{streaming}
@@
 While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
-
+{
+\small
 \bibliographystyle{abbrv}
 \bibliography{references}
-
+}
 \end{document}

### 2015-11-26 streaming.tex
Iterations

@@
 %!TEX root = paper.tex
-
 \section{Stream Analytics on Top of Dataflows}
@@
 To avoid arbitrary delays these systems regularly insert special events called ‘low watermarks’ that mark a global progress measure.
-In the case of time progress for example, that includes a time attribute ‘t’ indicating that all events lower than ‘t’ have already entered an operator.
+In the case of time progress for example, that includes a time attribute $t$ indicating that all events lower than $t$ have already entered an operator.
 The watermarks aid the execution engine to process events in the correct event order and serialize operations (such as window computations) via a unified measure of progress.
@@
 
-To cover a lot of different use cases, in Flink state is made explicit and is incorporated in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations.
+Ιn Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator state abstraction for marking partitioned key-value states and the associated operations.
 Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in the streaming applications.
 Flink’s checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
+\vspace{-8mm}
 
-As a simple example consider a map operator that computes a continuous average for a given stocks price.
+% As a simple example consider a map operator that computes a continuous average for a given stocks price.
 For this we need to keep a state with the current count and total sum for each stock in order to compute the average:
 
-\begin{lstlisting}[language=Scala]
-val priceStream: DataStream[(String,Double)] = ...
+% \begin{lstlisting}[language=Java]
+% val priceStream: DataStream[(String,Double)] = ...
  
-val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
-    (in, state: Option[(Int, Double)]) => {
-		val (count, sum) = state.getOrElse((0,0.0))
-		val newState = (count + 1, sum + in._2)
-		((in._1, newState._2/newState._1), newState)
-	}
-\end{lstlisting}
+% val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
+%     (in, state: Option[(Int, Double)]) => {
+% 		val (count, sum) = state.getOrElse((0,0.0))
+% 		val newState = (count + 1, sum + in._2)
+% 		((in._1, newState._2/newState._1), newState)
+% 	}
+% \end{lstlisting}
 
@@
 
-A global window creates a single logical group.
+\noindent A global window creates a single logical group.
 The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
@@
 This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
+
+\subsection{Asynchronous Streaming Iterations}
+Stream loops are crucial for several applications.
+Common examples are feedback loops for incrementally building and training machine learning models, reinforcement learning, and executing graph algorithms.
+Flink supports loops through explicit feedback edges in its dataflow graph, allowing to implement asynchronous iterative/incremental algorihtms.
+In Flink, feedback streams have special semantics and are treated in a way that is transparent to its scheduler: feedback streams do not act as dependencies in the dataflow graph, allowing Flink to employ simple DAG scheduling techniques while the contents of feedback streams are treated as operator state, in order to simplify fault tolerance \cite{carbone2015lightweight}.
+Flink allows constructing loops on top of its DataStream API, and can trivially subsume structured loops such as differential dataflows \cite{murray2013naiad}.
+\autoref{sec:batch-iterations} describes iterations support for batch analytics.

### 2015-11-27 exec-model.tex
small fixes

@@
 To this end, unary operators in Flink consuming a single stream partition, \emph{guarantee a FIFO order of records}.
-However, operators receiving more than one stream partition merge the streams as records arrive, in order to keep up with the streams' rate and avoid back pressure.
+However, operators receiving more than one stream partition merge the streams in arrival order, in order to keep up with the streams' rate and avoid back pressure.
 As a result, streaming dataflows in Flink do not provide ordering guarantees after any form of repartitioning or broadcasting, and the responsibility of dealing with out-of-order records is left to the operator implementation.

### 2015-11-27 streaming.tex
small fixes

@@
 \subsection{Stateful Stream Processing}
-While most operators in Flink’s DataStream API look like functional, side-effect free operators, they provide support for efficient stateful computations.
+While most operators in Flink's DataStream API look like functional, side-effect free operators, they provide support for efficient stateful computations.
 State is critical to many applications such as machine learning model building, graph analysis, user session handling and window aggregations.
@@
 
-Ιn Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator state abstraction for marking partitioned key-value states and the associated operations.
+In Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator state abstraction for marking partitioned key-value states and the associated operations.
 Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in the streaming applications.
-Flink’s checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
+Flink's checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
 \vspace{-8mm}
@@
 Note that Flink's windowing capabilities incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsume these windowing models.
-For example, below is a window definition of 6 seconds that slides every 2 seconds (Assigner).
+For example, below is a window definition with a range of 6 seconds that slides every 2 seconds (Assigner).
 The window results are computed once the watermark passes the end of the window (Trigger).

### 2015-11-27 exec-model.tex
small change in fault tolerance

@@
 The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers.
-Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.
+Furthermore, partial recovery of a failed subtask is possible by additionally replaying unprocessed records  buffered at the immediate upstream subtasks \cite{carbone2015lightweight}.
 

### 2015-11-27 paper.tex
small change in fault tolerance

@@
 \label{sec:conclusions}
-In this paper we presented Apache Flink, a platform that implements one universal dataflow engine designed to perform both stream and batch analytics.
-Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by the batch and a data stream APIs.
-The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state, and to partition, transform, discretize and aggregate a data stream.
+In this paper we presented Apache Flink, a platform that implements a universal dataflow engine designed to perform both stream and batch analytics.
+Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by both the batch and a data stream APIs, with different parameters.
+The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state and to partition, transform, and aggregate windows of a data stream.
 While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.

### 2015-11-27 streaming.tex
fix figure and add some example references

@@
 
-Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, but exhibit lower latency.
+Flink programs that are based on processing time, rely on the machine clocks, and hence a less reliable notion of time, which can lead to inconsistent replays upon recovery.
+However, they exhibit lower latency.
 Programs that are based on event time provide the most reliable semantics, but may exhibit latency due to event time-processing time lag.
@@
 For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine learning applications.
-Stream windows can also be modelled, in principle, by stateful operators which assign records to continuously updated buckets kept in memory as part of the operator state.
+Stream windows are  stateful operators which assign records to continuously updated buckets kept in memory as part of the operator state.
 
-In Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator state abstraction for marking partitioned key-value states and the associated operations.
+In Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator state abstraction for declaring partitioned key-value states and its associated operations.
 Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in the streaming applications.

### 2015-11-27 batch.tex
small fix

@@
 Most batch dataflow approaches execute iterations by submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}.
-Flink keeps the DAG-based runtime and scheduler, and allows for special ``head'' and ``tail'' tasks to signify the beginning and end of each iteration, that are connected with feedback edges.
+Flink keeps the DAG-based runtime and scheduler, and allows for special ``head'' and ``tail'' tasks that are connected with feedback edges, to signify the beginning and end of each iteration.
 Flink's iterations are competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}.

### 2015-11-27 intro.tex
small fix

@@
 Apache Flink follows a paradigm that embraces data stream processing as the unifying model for real-time analysis, continuous streams, and batch processing, both in the programming model and in the execution engine.
-In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka\footnote{http://kafka.apache.org/} or Amazon Kinesis\footnote{https://aws.amazon.com/kinesis/}), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
+In combination with durable message queues that allow quasi arbitrary replay of data streams (like Apache Kafka or Amazon Kinesis), stream processing programs make no difference between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data: these types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
 Through a highly flexible windowing mechanism, Flink programs can compute both early and approximate, as well as delayed and accurate results in the same operation, obviating the need to combine different systems for the two use cases.
@@
 The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning.
-Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing\footnote{For an update list, refer to \url{https://github.com/apache/flink/graphs/contributors}}), and is used in production in several companies.
+Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing, and is used in production in several companies.
 

### 2015-11-27 streaming.tex
small fix

@@
 
-Note that streams are already partitioned on a key before windowing, so the above is a \textit{local operator} that does not require coordination between machines.
+Note that if the stream above is partitioned on a key before  windowing, the window operation above is \textit{local} and thus, it does not require coordination between machines.
 This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
@@
 Common examples are feedback loops for incrementally building and training machine learning models, reinforcement learning, and executing graph algorithms.
-Flink supports loops through explicit feedback edges in its dataflow graph, allowing to implement asynchronous iterative/incremental algorihtms.
+Flink supports loops through explicit feedback edges in its dataflow graph, allowing the implementation of asynchronous iterative/incremental algorihtms.
 In Flink, feedback streams have special semantics and are treated in a way that is transparent to its scheduler: feedback streams do not act as dependencies in the dataflow graph, allowing Flink to employ simple DAG scheduling techniques while the contents of feedback streams are treated as operator state, in order to simplify fault tolerance \cite{carbone2015lightweight}.

### 2015-11-29 batch.tex
Small changes
remake iteration sections

@@
 \label{sec:batch-iterations}
-Most batch dataflow approaches execute iterations by submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}.
-Flink keeps the DAG-based runtime and scheduler, and allows for special ``head'' and ``tail'' tasks that are connected with feedback edges, to signify the beginning and end of each iteration.
-Flink's iterations are competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}.
-This is achieved through the concept of  \emph{delta} iterations, which exploit sparse computational dependencies, and are used as the basis for Gelly, Flink's Graph API.
-Finally, Flink offers Bulk Synchronous Parallel (BSP) as well as Stale Synchronous Parallel (SSP) iterations in its DataSet API.
+Structured iteration support is the basis of most distributed batch processing applications.
+Iterative graph analytics, parallel gradient descent and optimisation techniques have been implemented in the past on top of Bulk Synchronous Parallel (BSP) and Stale Synchronous Parallel (SSP) models, among others.
+The Flink execution model, allows for any type of structured iteration logic using iteration control events, as explained in \ref{sec:iterations}.
+For example in the case of a BSP execution, iteration control events mark the beginning and the end of supersteps in an iterative computation.
+Finally, Flink introduces further novel optimisation techniques such as the concept of \emph{delta} iterations \cite{DBLP:journals/pvldb/EwenTKM12}, which can exploit sparse computational dependencies (already part of Gelly, Flink's Graph API).

### 2015-11-29 exec-model.tex
fix figure and add some example references
remake iteration sections

@@
 Once the state has been backed up, the operator forwards the barrier downstream.
+Eventually, all operators will register a snapshot of their state and a global snapshot will be complete.
+For example, in \autoref{fig:snapshots} we show that snapshot $t2$ contains all operator states that are the result of consuming all records before \emph{t2 barrier}.
 ABS bears resemblances to the seminal Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
@@
 \end{itemize}
+
+\subsection{Iterative Dataflows}
+\label{sec:iterations}
+
+\begin{figure}[t!]
+  \centering
+    \includegraphics[width=.75\textwidth]{figs/iterations.pdf}
+    \vspace{-6mm}
+  \caption{The iteration model of Apache Flink}
+  \vspace{-2mm}
+  \label{fig:iterations}
+\end{figure}
+
+Loops and iterations are crucial for several applications.
+Support for iterations in data-intensive computing systems typically relies on submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}.
+As depicted in \autoref{fig:iterations}, Flink maintains the DAG-based runtime and scheduler, and allows for special iteration ``head'' and ``tail'' tasks that are implicitly connected with feedback edges.
+The role of these tasks is to establish an active feedback channel to the DAG and provide coordination for processing data records in transit within this feedback channel.
+Coordination is needed for implementing any type of structured parallel iteration model such as the Bulk Synchronous Parallel (BSP) model as we explain in section \ref{sec:batch-iterations}.

### 2015-11-29 streaming.tex
remake iteration sections

@@
 \subsection{Asynchronous Streaming Iterations}
-Stream loops are crucial for several applications.
-Common examples are feedback loops for incrementally building and training machine learning models, reinforcement learning, and executing graph algorithms.
-Flink supports loops through explicit feedback edges in its dataflow graph, allowing the implementation of asynchronous iterative/incremental algorihtms.
-In Flink, feedback streams have special semantics and are treated in a way that is transparent to its scheduler: feedback streams do not act as dependencies in the dataflow graph, allowing Flink to employ simple DAG scheduling techniques while the contents of feedback streams are treated as operator state, in order to simplify fault tolerance \cite{carbone2015lightweight}.
-Flink allows constructing loops on top of its DataStream API, and can trivially subsume structured loops such as differential dataflows \cite{murray2013naiad}.
-\autoref{sec:batch-iterations} describes iterations support for batch analytics.
+Stream loops are essential for several applications.
+Common examples are feedback loops for incrementally building and training machine learning models, reinforcement learning and graph algorithms in the semi-streaming model \cite{feigenbaum2005graph}.
+In most such cases, feedback loops need no coordination.
+Asynchronous iterations cover the communication needs for streaming applications and differ from parallel optimisation problems that are based on structured iterations on finite data.
+As presented, in \ref{sec:iterations} and \ref{fig:iterations} the execution model of Apache Flink already covers asynchronous iterations, when no iteration control mechanism is enabled.
+In addition, to comply with fault-tolerance guarantees feedback streams are treated as operator state within the implicit iteration head operator and are part of a global snapshot, as explained in detail in \cite{carbone2015lightweight}.
+The DataStream API allows for an explicit definition of feedback streams and can trivially subsume future support for structured loops over streams such as differential dataflows \cite{murray2013naiad}.

### 2015-11-29 streaming.tex
autoref not ref

@@
 Asynchronous iterations cover the communication needs for streaming applications and differ from parallel optimisation problems that are based on structured iterations on finite data.
-As presented, in \ref{sec:iterations} and \ref{fig:iterations} the execution model of Apache Flink already covers asynchronous iterations, when no iteration control mechanism is enabled.
+As presented, in \autoref{sec:iterations} and \autoref{fig:iterations} the execution model of Apache Flink already covers asynchronous iterations, when no iteration control mechanism is enabled.
 In addition, to comply with fault-tolerance guarantees feedback streams are treated as operator state within the implicit iteration head operator and are part of a global snapshot, as explained in detail in \cite{carbone2015lightweight}.

### 2015-11-29 exec-model.tex
small edit

@@
 Support for iterations in data-intensive computing systems typically relies on submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}.
+Iterations in Flink are implemented as specialized operators that can contain other operators (iteration step) and can be part of the dag.
 As depicted in \autoref{fig:iterations}, Flink maintains the DAG-based runtime and scheduler, and allows for special iteration ``head'' and ``tail'' tasks that are implicitly connected with feedback edges.
-The role of these tasks is to establish an active feedback channel to the DAG and provide coordination for processing data records in transit within this feedback channel.
+The role of these tasks is to establish an active feedback channel to the iteration step and provide coordination for processing data records in transit within this feedback channel.
 Coordination is needed for implementing any type of structured parallel iteration model such as the Bulk Synchronous Parallel (BSP) model as we explain in section \ref{sec:batch-iterations}.

### 2015-11-30 batch.tex
autoref not ref
Updates from ShareLaTeX

--- 2015-11-27 batch.tex
@@
 \subsection{Query Optimization} 
-
 Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation.

### 2015-11-30 batch.tex

--- 2015-11-29 batch.tex
@@
 Iterative graph analytics, parallel gradient descent and optimisation techniques have been implemented in the past on top of Bulk Synchronous Parallel (BSP) and Stale Synchronous Parallel (SSP) models, among others.
-The Flink execution model, allows for any type of structured iteration logic using iteration control events, as explained in \ref{sec:iterations}.
+The Flink execution model, allows for any type of structured iteration logic using iteration control events, as explained in \autoref{sec:iterations}.
 For example in the case of a BSP execution, iteration control events mark the beginning and the end of supersteps in an iterative computation.

### 2015-12-02 batch.tex
Sent to Stephan, Kostas.

@@
 \label{sec:batch-iterations}
-Structured iteration support is the basis of most distributed batch processing applications.
 Iterative graph analytics, parallel gradient descent and optimisation techniques have been implemented in the past on top of Bulk Synchronous Parallel (BSP) and Stale Synchronous Parallel (SSP) models, among others.
-The Flink execution model, allows for any type of structured iteration logic using iteration control events, as explained in \autoref{sec:iterations}.
-For example in the case of a BSP execution, iteration control events mark the beginning and the end of supersteps in an iterative computation.
+The Flink execution model, allows for any type of structured iteration logic to be implemented on top, by using iteration control events.
+For instance, in the case of a BSP execution, iteration control events mark the beginning and the end of supersteps in an iterative computation.
 Finally, Flink introduces further novel optimisation techniques such as the concept of \emph{delta} iterations \cite{DBLP:journals/pvldb/EwenTKM12}, which can exploit sparse computational dependencies (already part of Gelly, Flink's Graph API).

### 2015-12-02 exec-model.tex
Updates from ShareLaTeX
Sent to Stephan, Kostas.

--- 2015-11-27 exec-model.tex
@@
 \end{itemize}
+
+\subsection{Iterative Dataflows}
+\label{sec:iterations}
+
+\begin{figure}[t!]
+  \centering
+    \includegraphics[width=.75\textwidth]{figs/iterations.pdf}
+    \vspace{-6mm}
+  \caption{The iteration model of Apache Flink}
+  \vspace{-2mm}
+  \label{fig:iterations}
+\end{figure}
+
+Incremental processing and iterations are crucial for applications such as graph processing and machine learning.
+Support for iterations in data-parallel processing platforms typically relies on submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10},
+or feedback edges \cite{murray2013naiad}.
+
+Iterations in Flink are implemented as  \emph{iteration steps}, special operators that themselves can contain an execution graph (\autoref{fig:iterations}).
+To maintain the (acyclic) DAG-based runtime and scheduler, Flink allows for iteration ``head'' and ``tail'' tasks that are \emph{implicitly} connected with feedback edges.
+The role of these tasks is to establish an active feedback channel to the iteration step and provide coordination for processing data records in transit within this feedback channel.
+Coordination is needed for implementing any type of structured parallel iteration model such as the Bulk Synchronous Parallel (BSP) model and is implemented using control event.
+We explain how iterations are implemented in the DataStream and DataSet APIs in \autoref{sec:stream-iterations} and  \autoref{sec:batch-iterations} respectively.

### 2015-12-02 streaming.tex
stream not streaming
Sent to Stephan, Kostas.

@@
 
-\subsection{Asynchronous Streaming Iterations}
-Stream loops are essential for several applications.
-Common examples are feedback loops for incrementally building and training machine learning models, reinforcement learning and graph algorithms in the semi-streaming model \cite{feigenbaum2005graph}.
+\subsection{Asynchronous Stream Iterations}
+\label{sec:stream-iterations}
+Loops in streams are essential for several applications such as incrementally building and training machine learning models, reinforcement learning and graph algorithms in the semi-streaming model \cite{feigenbaum2005graph}.
 In most such cases, feedback loops need no coordination.

### 2016-05-25 paper.tex
Sent to Stephan, Kostas.
camera-ready v1

@@
 
-\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
-
 % code listings
@@
 
+\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
 \newcommand{\sectionautorefname}{Section}%
-
 \newcommand{\para}[1]{\vspace{2mm}\noindent\textbf{#1}}
@@
 % url references
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 
-\usepackage{subfigure}
+\usepackage{subfig}
 
@@
 
-\title{Apache Flink\texttrademark: Unified Stream and Batch Processing \\ in a Single Engine}
+\title{Apache Flink\texttrademark: Stream and Batch Processing in a Single Engine}
 
-\author{Authors}
+\author{Paris Carbone\textsuperscript{1} \\ Asterios Katsifodimos\textsuperscript{3} \\\\ \small{ \textsf{\small \textsuperscript{1}KTH \& SICS Sweden}} \\ \small \href{parisc,haridi@kth.se}{parisc,haridi@kth.se}
+\and Stefan Ewen\textsuperscript{2} \\ Volker Markl\textsuperscript{3} \\\\ \small \textsf{\textsuperscript{2}Data Artisans GmbH} \\ \small \href{first@data-artisans.de}{first@data-artisans.de}
+\and Seif Haridi\textsuperscript{1} \\ Kostas Tzoumas\textsuperscript{2} \\\\ \small \textsf{\textsuperscript{3}TU Berlin} \\ \small \href{fist.last@tu-berlin.de}{first.last@tu-berlin.de}
+%
+\vspace{-5mm}
+}
+
+%\and Stefan Ewen$^2$ \and Seif Haridi$^1$ \and Asterios Katsifodimos$^3$ \and Volker Markl$^3$ \and Kostas Tzoumas$^2$}
 
@@
 \begin{abstract}
-Apache Flink is an open source system for processing streaming and batch data.
-Flink is built on the philosophy that many classes of data processing applications, including real-time analytics, continuous data pipelines, historic data processing (batch), and iterative algorithms (machine learning, graph analysis) can be expressed and executed as pipelined fault tolerant dataflows.
-In this paper, we present Flink’s architecture and expand on how a (seemingly diverse) set of use cases can be unified under the same execution model.
+Apache Flink is an open-source system for processing streaming and batch data.
+Flink is built on the philosophy that many classes of data processing applications, including real-time analytics, continuous data pipelines, historic data processing (batch), and iterative algorithms (machine learning, graph analysis) can be expressed and executed as pipelined fault-tolerant dataflows.
+In this paper, we present Flink's architecture and expand on how a (seemingly diverse) set of use cases can be unified under the same execution model.
 \end{abstract}
 
-\input{intro}
-\input{architecture}
-\input{exec-model}
-\input{streaming}
-\input{batch}
-\input{related}
+%!TEX root = paper.tex
 
+\section{Introduction}
+\label{sec:intro}
+Data-stream processing (e.g, as exemplified by Complex Event Processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop ) were traditionally considered as two very different types of applications.
+They were programmed using different programming models and APIs, and were executed by different systems (e.g., dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft StreamInsight, or Streambase versus relational databases or execution engines for Hadoop, including Apache Spark, and Apache Drill).
+Traditionally, batch data analysis made up for the lion's share of the use cases, data sizes, and market, while streaming data analysis mostly served specialized applications.
+
+It is becoming more and more apparent, however, that a huge number of today's large-scale data processing use cases handle data that is, in reality, continuously produced over time.
+These continuous streams of data come for example from web logs, application logs, sensors, or as changes to application state in databases (transaction log records).
+Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of data production.
+Instead, data records are (often artificially) batched into static data sets (for example hourly, daily, or monthly chunks) and then processed in a time-agnostic fashion.
+Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline.
+Architectural patterns such as the "lambda architecture" \cite{marz2015big} combine batch and stream processing systems to implement multiple paths of computation: a streaming fast path for timely approximate results, and a batch offline path for late accurate results.
+All these approaches suffer from high latency (imposed by batches), high complexity (connecting and orchestrating several systems, and implementing business logic twice), as well as arbitrary inaccuracy, as the time dimension is not explicitly handled by the application code.
+
+Apache Flink follows a paradigm that embraces data-stream processing as the unifying model for real-time analysis, continuous streams, and batch processing both in the programming model and in the execution engine.
+In combination with durable message queues that allow quasi-arbitrary replay of data streams (like Apache Kafka or Amazon Kinesis), stream processing programs make no distinction between processing the latest events in real-time, continuously aggregating data periodically in large windows, or processing terabytes of historical data.
+Instead, these different types of computations simply start their processing at different points in the durable stream, and maintain different forms of state during the computation.
+Through a highly flexible windowing mechanism, Flink programs can compute both early and approximate, as well as delayed and accurate, results in the same operation, obviating the need to combine different systems for the two use cases.
+Flink supports different notions of time (event-time, ingestion-time, processing-time) in order to give programmers high flexibility in defining how events should be correlated.
+
+At the same time, Flink acknowledges that there is, and will be, a need for dedicated batch processing (dealing with static data sets).
+Complex queries over static data are still a good match for a batch processing abstraction.
+Furthermore, batch processing is still needed both for legacy implementations of streaming use cases, and for analysis applications where no efficient algorithms are yet known that perform this kind of processing on streaming data.
+Batch programs are special cases of streaming programs, where the stream is finite, and the order and time of records does not matter (all records implicitly belong to one all-encompassing window).
+However, to support batch use cases with competitive ease and performance, Flink has a specialized API for processing static data sets, uses specialized data structures and algorithms for the batch versions of operators like join or grouping, and uses dedicated scheduling strategies.
+The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning.
+Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 170 open-source contributors at the time of this writing), and is used in production in several companies.
+
+\vspace{2mm}
+\noindent The contributions of this paper are the following:\vspace{-2mm}
+\begin{itemize}
+	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
+	\vspace{-3mm}
+	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault-tolerant streaming dataflows (\autoref{sec:execution}).
+	\vspace{-3mm}
+	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:streaming}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
+\end{itemize}
+
+%!TEX root = paper.tex
+
+% \vspace{-mm}
+\section{System Architecture}
+\label{sec:architecture}
+
+In this section we lay out the architecture of Flink as a software stack, and as a distributed system.
+While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
+
+\para{Flink's Runtime and APIs.}  \autoref{fig:stack} shows Flink's software stack.
+The core of Flink is the distributed dataflow engine, which executes dataflow programs.
+A Flink runtime program is a DAG of stateful operators connected with data streams.
+There are two core APIs in Flink: the DataSet API for processing finite data sets (often referred to as \emph{batch processing}), and the DataStream API for processing potentially unbounded data streams (often referred to as \emph{stream processing}).
+Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create runtime programs executable by the engine.
+As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing.
+On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML (machine learning), Gelly (graph processing), and Table (SQL-like operations)).
+
+As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the Job Manager, and one or more Task Managers.
+The client takes the program code, transforms it to a dataflow graph, and submits that to the JobManager.
+This transformation phase also examines the data types (schema) of the data exchanged between operators and creates serializers and other type/schema specific code.
+DataSet programs additionally go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in \autoref{sec:batch}).
+
+The JobManager coordinates the distributed execution of the dataflow.
+It tracks the state and progress of each operator and stream, schedules new operators, and coordinates checkpoints and recovery.
+In a high-availability setup, the JobManager persists a minimal set of metadata at each checkpoint to a fault-tolerant storage, such that a standby JobManager can reconstruct the checkpoint and recover the dataflow execution from there.
+The actual data processing takes place in the TaskManagers.
+A TaskManager executes one or more operators that produce streams, and reports on their status to the JobManager.
+The TaskManagers maintain the buffer pools to buffer or materialize the streams, and the network connections to exchange the data streams between operators.
+
+\begin{figure}[t!]
+\begin{minipage}{1.1\linewidth}
+      \centering
+      \hspace{-0.1\linewidth}
+      \begin{minipage}{0.48\linewidth}
+          \begin{figure}[H]
+              \centering
+			\includegraphics[width=.88\textwidth]{figs/stack}
+			\vspace{-3mm}
+			\caption{Flink's software stack.}
+			\vspace{-3mm}
+			\label{fig:stack}
+          \end{figure}
+      \end{minipage}
+      % \hspace{0.05\linewidth}
+      \begin{minipage}{0.5\linewidth}
+          \begin{figure}[H]
+				\includegraphics[width=.95\textwidth]{figs/architecture.pdf}
+    			\vspace{-2.5mm}
+    			\caption{Flink's process model.}
+    			\label{fig:process-model}
+          \end{figure}
+      \end{minipage}
+  \end{minipage}
+\end{figure}
+
+\vspace{-3mm}
+\section{The Common Fabric: Streaming Dataflows}
+\label{sec:execution}
+
+Although users can write Flink programs using a multitude of APIs, all Flink programs eventually compile down to a common representation: the dataflow graph.
+The dataflow graph is executed by Flink's runtime engine, the common layer underneath both the batch processing (DataSet) and stream processing (DataStream) APIs.
+
+\vspace{-2mm}
+\subsection{Dataflow Graphs}
+The dataflow graph, as depicted in \autoref{fig:dataflow}, is a directed acyclic graph (DAG) that consists of (i) stateful operators, and (ii) data streams that represent data produced by an operator and are available for consumption by operators.
+Since dataflow graphs are executed in a data-parallel fashion, operators are parallelized into one or more parallel instances called \emph{subtasks} and streams are split into one or more \emph{stream partitions} (one partition per producing subtask).
+The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions.
+Many of these operators are implementations of textbook versions of well known algorithms.
+\autoref{sec:streaming} provides details on the implementation of windowing operators.
+Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, and merge.
+
+\begin{figure}[t!]
+\begin{minipage}{1.1\linewidth}
+      \centering
+       \hspace{-0.1\linewidth}
+      \begin{minipage}{0.55\linewidth}
+        \begin{figure}[H]
+        \centering
+        \includegraphics[width=.999\textwidth]{figs/dataflow}
+        \vspace{-5mm}
+        \caption{A simple dataflow graph.}
+        \label{fig:dataflow}
+        \end{figure}
+      \end{minipage}
+      \hspace{0.03\linewidth}
+      % \vspace{-3mm}
+      \begin{minipage}{0.32\linewidth}
+          \begin{figure}[H]
+				\includegraphics[width=.99\textwidth]{figs/latency-throughput.pdf}
+				\vspace{-7mm}
+    			\caption{The effect of buffer-timeout in latency and throughput.}
+    			\label{fig:latency-throughput}
+          \end{figure}
+      \end{minipage}
+  \end{minipage}
+  \vspace{-4mm}
+\end{figure}
+
+\subsection{Data Exchange through Intermediate Data Streams}
+Flink's intermediate data streams are the core abstraction for data-exchange between operators.
+An intermediate data stream represents a logical handle to the data that is produced by an operator and can be consumed by one or more operators.
+Intermediate streams are logical in the sense that the data they point to may or may not be materialized on disk.
+The particular behavior of a data stream is parameterized by the higher layers in Flink (e.g., the program optimizer used by the DataSet API).
+
+\para{Pipelined and Blocking Data Exchange.} \emph{Pipelined intermediate streams} exchange data between concurrently running producers and consumers resulting in pipelined execution.
+As a result, pipelined streams propagate back pressure from consumers to producers, modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations.
+Flink uses pipelined streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.
+\emph{Blocking streams}, on the other hand, are applicable to bounded data streams.
+A blocking stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages.
+Blocking streams naturally require more memory, frequently spill to secondary storage, and do not propagate backpressure.
+They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as sort-merge joins, may cause distributed deadlocks.
+
+\para{Balancing Latency and Throughput.} Flink's data-exchange mechanisms are implemented around the exchange of buffers.
+When a data record is ready on the producer side, it is serialized split into one or more buffers (a buffer can also fit multiple records) that can be forwarded to consumers.
+A buffer is sent to a consumer either i) as soon as it is full, or ii) when a timeout condition is reached.
+This enables Flink to achieve high throughput by setting the size of buffers to a high value (e.g., a few kilobytes), as well as low latency by setting the buffer timeout to a low value (e.g., a few milliseconds).
+\autoref{fig:latency-throughput} shows the effect of buffer-timeouts to the throughput and latency of delivering records in a simple streaming grep job on 30 machines (120 cores).
+Flink can achieve an observable 99$^{\mbox{\small{th}}}$-percentile latency of 20 milliseconds.
+The corresponding throughput is 1.5 million events per second.
+As we increase the buffer timeout, we see an increase in latency with an increase in throughput, until full throughput is reached (i.e., buffers fill up faster than the timeout expiration).
+At a buffer timeout of 50 milliseconds, the cluster reaches a throughput of more than 80 million events per second with a 99$^{\mbox{\small{th}}}$-percentile latency of 50 milliseconds.
+
+\para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events.
+These are special events injected in the data stream by operators, and are delivered in-order along with all other data records and events within a stream partition.
+The receiving operators react to these events by performing certain actions upon their arrival.
+Flink uses lots of special types of control events, including: \vspace{-2mm}
+\begin{itemize}
+\item \textit{Checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
+\item \textit{Watermarks} signaling the progress of event-time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
+\item \textit{Iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}).
+\vspace{-1mm}
+\end{itemize}
+
+As mentioned above, control events assume that a stream partition preserves the order of records.
+To this end, unary operators in Flink that consume a single stream partition, \emph{guarantee a FIFO order of records}.
+However, operators receiving more than one stream partition merge the streams in arrival order, in order to keep up with the streams' rates and avoid back pressure.
+As a result, streaming dataflows in Flink do not provide ordering guarantees after any form of repartitioning or broadcasting, and the responsibility of dealing with out-of-order records is left to the operator implementation.
+We found that this arrangement gives the most efficient design, as most operators do not require deterministic order (e.g., hash-joins, maps, etc.), and operators that need to compensate for out-of-order arrivals (such as event-time windows) can do that more efficiently as part of the operator logic.
+
+\vspace{-3mm}
+\subsection{Fault Tolerance}
+\label{sec:fault-tolerance}
+
+Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution.
+The general assumption the system makes to effectively provide these guarantees is that the data sources are persistent and replayable.
+Examples of such sources are files and durable message queues (e.g.
+Apache Kafka).
+In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators.
+
+The checkpointing mechanism of Apache Flink builds on the notion of distributed consistent snapshots to achieve exactly-once-processing guarantees.
+The possibly unbounded nature of a data stream makes re-computation upon recovery impractical, as possibly months of computation will need to be replayed for a long-running job.
+To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams, at regular intervals.
+
+The core challenge lies in taking a consistent snapshot of all parallel operators without halting the execution of the topology.
+In essence, the snapshot of all operators should refer to the same logical time in the computation.
+The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
+Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
+
+\begin{figure}[t!]
+	\centering
+  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
+  	\vspace{-6mm}
+	\caption{Asynchronous Barrier Snapshotting.}
+	\vspace{-2mm}
+	\label{fig:snapshots}
+\end{figure}
+
+An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received.
+Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS).
+Once the state has been backed up, the operator forwards the barrier downstream.
+Eventually, all operators will register a snapshot of their state and a global snapshot will be complete.
+For example, in \autoref{fig:snapshots} we show that \emph{snapshot t2} contains all operator states that are the result of consuming all records before \emph{t2 barrier}.
+ABS bears resemblances to the Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
+However, because of the DAG structure of a Flink program, ABS does not need to checkpoint in-flight records but solely relies on the aligning phase to apply all their effects to the operator states.
+This guarantees that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).
+
+Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier for which there is a snapshot.
+The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers.
+Furthermore, partial recovery of a failed subtask is possible by additionally replaying unprocessed records  buffered at the immediate upstream subtasks \cite{carbone2015lightweight}.
+
+\vspace{1mm}
+\noindent ABS provides several benefits:
+\begin{inparaenum}[i)]
+\item it guarantees exactly-once state updates without ever pausing the computation
+\item it is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval and
+\item it is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc., depending on the larger environment in which Flink is used.
+\end{inparaenum}
+
+\subsection{Iterative Dataflows}
+\label{sec:iterations}
+
+Incremental processing and iterations are crucial for applications such as graph processing and machine learning.
+Support for iterations in data-parallel processing platforms typically relies on submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, or feedback edges \cite{murray2013naiad}.
+Iterations in Flink are implemented as  \emph{iteration steps}, special operators that themselves can contain an execution graph (\autoref{fig:iterations}).
+To maintain the (acyclic) DAG-based runtime and scheduler, Flink allows for iteration ``head'' and ``tail'' tasks that are \emph{implicitly} connected with feedback edges.
+The role of these tasks is to establish an active feedback channel to the iteration step and provide coordination for processing data records in transit within this feedback channel.
+Coordination is needed for implementing any type of structured parallel iteration model such as the Bulk Synchronous Parallel (BSP) model and is implemented using control event.
+We explain how iterations are implemented in the DataStream and DataSet APIs in \autoref{sec:stream-iterations} and  \autoref{sec:batch-iterations} respectively.
+
+\vspace{-4mm}
+\section{Stream Analytics on Top of Dataflows}
+\label{sec:streaming}
+
+Flink's DataStream API implements a full stream-analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
+The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
+Since Flink's runtime already supports pipelined data transfers, continuous stateful operators, and a fault-tolerance mechanism for consistent state updates, overlaying a stream processor on top of it essentially boils down to implementing a windowing system and a state interface.
+As noted, these are invisible to the runtime, which sees windows as just an implementation of stateful operators.
+
+\vspace{-3mm}
+\subsection{The Notion of Time}
+\label{sec:streaming-time}
+Flink distinguishes between two notions of time: i) event-time, that  stands for the time that an event was generated at its origin (e.g., the time that a sensor or a mobile device generated a signal) and ii) processing-time, which is the wall-clock time of the machine that is processing the data.
+
+In distributed systems there is an arbitrary skew between event-time and processing-time \cite{akidau2015dataflow}.
+This skew may mean arbitrary delays for getting an answer based on event-time semantics.
+To avoid arbitrary delays, these systems regularly insert special events called \emph{low watermarks} that mark a global progress measure.
+In the case of time progress for example, a watermark includes a time attribute $t$ indicating that all events lower than $t$ have already entered an operator.
+The watermarks aid the execution engine in processing events in the correct event order and serialize operations (such as window computations) via a unified measure of progress.
+
+Watermarks originate at the sources of a topology, where we can determine the time inherent in future elements.
+The watermarks propagate from the sources throughout the other operators of the data flow.
+Operators decide how they react to watermarks.
+Simple operations such as map or filter just forward the watermarks they receive, while more complex operators that do calculation based on the watermarks (e.g., event-time windows) first compute results triggered by a watermark and then forward it.
+If an operation has more than one input, the system only forwards the minimum of the incoming watermarks to the operator, ensuring correct results.
+
+Flink programs that are based on processing-time rely on the machine clocks, and hence a less reliable notion of time, which can lead to inconsistent replays upon recovery.
+However, they exhibit lower latency.
+Programs that are based on event-time provide the most reliable semantics, but may exhibit latency due to event-time-processing-time lag.
+Flink includes a third notion of time as a special case of event-time called \emph{ingestion-time}, which is the time that events enter Flink.
+This provides option lower processing latency than event-time, and more accurate results compared to processing-time.
+
+\begin{figure}[t!]
+  \centering
+    \includegraphics[width=.75\textwidth]{figs/iterations.pdf}
+    \vspace{-6mm}
+  \caption{The iteration model of Apache Flink}
+  \vspace{-2mm}
+  \label{fig:iterations}
+\end{figure}
+
+\vspace{-3mm}
+\subsection{Stateful Stream Processing}
+While most operators in Flink's DataStream API look like functional, side-effect-free operators, they provide support for efficient stateful computations.
+State is critical to many applications, such as machine-learning model building, graph analysis, user session handling and window aggregations.
+There is a plethora of different types of states depending on the use case.
+For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine-learning applications.
+Stream windows are  stateful operators that assign records to continuously updated buckets kept in memory as part of the operator state.
+
+In Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator-state abstraction for declaring partitioned key-value states and the state's associated operations.
+Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in streaming applications.
+Flink's checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
+\vspace{-8mm}
+
+% As a simple example consider a map operator that computes a continuous average for a given stocks price.
+For this we need to keep a state with the current count and total sum for each stock in order to compute the average:
+
+% \begin{lstlisting}[language=Java]
+% val priceStream: DataStream[(String,Double)] = ...
+ 
+% val avgPerStock : DataStream[(String, Double)] = source.keyBy(0).mapWithState(
+%     (in, state: Option[(Int, Double)]) => {
+% 		val (count, sum) = state.getOrElse((0,0.0))
+% 		val newState = (count + 1, sum + in._2)
+% 		((in._1, newState._2/newState._1), newState)
+% 	}
+% \end{lstlisting}
+
+\subsection{Stream Windows}
+Incremental computations over unbounded streams are often evaluated over continously evolving logical views, called windows.
+Apache Flink incorporates windowing within a stateful operator that is configured via a flexible declaration composed out of three core functions: a window \textit{assigner} and optionally a \textit{trigger} and an \textit{evictor}.
+All three functions can be selected among a pool of common predefined implementations (e.g.
+sliding time windows) or can be explicitly defined by the user as UDFs.
+
+More specifically, the assigner is responsible for assigning each record to logical windows.
+For example, this decision can be based on the timestamp of a record when it comes to event-time windows.
+Note that in the case of sliding windows, an element can belong to multiple logical windows.
+An optional trigger defines when the operation associated with the window definition is performed.
+Finally, an optional evictor determines which records to retain within each window.
+Flink's window assignment process is uniquely capable of covering all known window types such as periodic time- and count-windows, punctuation, landmark, session and delta windows.
+Note that Flink's windowing capabilities incorporate out-of-order processing seamlessly, similarly to Google Cloud Dataflow \cite{akidau2015dataflow} and, in principle, subsume these windowing models.
+For example, below is a window definition with a range of 6 seconds that slides every 2 seconds (Assigner).
+The window results are computed once the watermark passes the end of the window (Trigger).
+
+\begin{lstlisting}[language=Java]
+stream
+  .window(SlidingTimeWindows.of(Time.of(6, SECONDS), Time.of(2, SECONDS))
+  .trigger(EventTimeTrigger.create())
+\end{lstlisting}
+
+\noindent A global window creates a single logical group.
+The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
+
+\begin{lstlisting}[language=Java]
+stream
+  .window(GlobalWindow.create())
+  .trigger(Count.of(1000))
+  .evict(Count.of(100))
+\end{lstlisting}
+
+Note that if the stream above is partitioned on a key before  windowing, the window operation above is \textit{local} and thus, it does not require coordination between machines.
+This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
+
+\subsection{Asynchronous Stream Iterations}
+\label{sec:stream-iterations}
+Loops in streams are essential for several applications such as incrementally building and training machine learning models, reinforcement learning and graph approximations \cite{feigenbaum2005graph,chandramouli2009fly}.
+In most such cases, feedback loops need no coordination.
+Asynchronous iterations cover the communication needs for streaming applications and differ from parallel optimisation problems that are based on structured iterations on finite data.
+As presented in \autoref{sec:iterations} and \autoref{fig:iterations}, the execution model of Apache Flink already covers asynchronous iterations, when no iteration control mechanism is enabled.
+In addition, to comply with fault-tolerance guarantees, feedback streams are treated as operator state within the implicit-iteration head operator and are part of a global snapshot \cite{carbone2015lightweight}.
+The DataStream API allows for an explicit definition of feedback streams and can trivially subsume support for structured loops over streams\cite{murray2013naiad} as well as progress tracking\cite{chandramouli2009fly}.
+
+%!TEX root = paper.tex
+
+\vspace{-3mm}
+\section{Batch Analytics on Top of Dataflows}
+\label{sec:batch}
+
+A bounded data set is a special case of an unbounded data stream.
+Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by FlinkFlink'ss features that we presented above.
+However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and ii) programs that process bounded data sets are amenable to additional optimizations, more efficient book-keeping for fault-tolerance, and staged scheduling.
+
+\vspace{2mm}
+\noindent Flink approaches batch processing as follows:\vspace{-3mm}
+\begin{itemize}
+	\item Batch computations are executed by the same runtime as streaming computations.
+The runtime executable may be parameterized with blocked data streams to break up large computations into isolated stages that are scheduled successively.
+\vspace{-3mm}
+	\item Periodic snapshotting is turned off when its overhead is high.
+Instead, fault recovery can be achieved by replaying the lost stream partitions from the latest materialized intermediate stream (possibly the source).\vspace{-3mm}
+	\item Blocking operators (e.g., sorts) are simply operator implementations that happen to block until they have consumed their entire input.
+The runtime is not aware of whether an operator is blocking or not.
+These operators use managed memory provided by Flink (either on or off the JVM heap), and can spill to disk if their inputs exceed their memory bounds.\vspace{-3mm}
+	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc).\vspace{-3mm}
+	\item A query optimization layer transforms a DataSet program to an efficient executable.\vspace{-3mm}
+\end{itemize}
+
+\noindent We provide more details on these aspects below.
+
+\subsection{Query Optimization} 
+Flink's optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting-property propagation.
+However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer.
+For the same reason, cardinality and cost-estimation methods are equally difficult to employ.
+Flink's runtime supports various execution strategies including repartition and broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations.
+Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation \cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans.
+The cost includes network and disk I/O as well as CPU cost.
+To overcome the cardinality estimation issues in the presence of UDFs, Flink's optimizer can use hints that are provided by the programmer.
+
+\subsection{Memory Management} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into memory segments.
+Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the serialization and deserialization overhead at a minimum and partially spilling data to disk when needed.
+To handle arbitrary objects, Flink uses type inference and  custom serialization mechanisms.
+By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and use cache-efficient and robust algorithms that scale gracefully under memory pressure.
+
+\subsection{Batch Iterations}
+\label{sec:batch-iterations}
+Iterative graph analytics, parallel gradient descent and optimisation techniques have been implemented in the past on top of Bulk Synchronous Parallel (BSP) and Stale Synchronous Parallel (SSP) models, among others.
+Flink's execution model allows for any type of structured iteration logic to be implemented on top, by using iteration-control events.
+For instance, in the case of a BSP execution, iteration-control events mark the beginning and the end of supersteps in an iterative computation.
+Finally, Flink introduces further novel optimisation techniques such as the concept of \emph{delta} iterations \cite{DBLP:journals/pvldb/EwenTKM12}, which can exploit sparse computational dependencies (already part of Gelly, Flink's Graph API).
+
+%!TEX root = paper.tex
+\vspace{-4mm}
+\section{Related work}
+\vspace{-2mm}
+\label{sec:related}
+There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
+
+\para{Batch Processing.} Apache Hadoop is the most popular open-source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
+Dryad~\cite{isard2007dryad} introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer}, which added a language and an SQL optimizer on top of it.
+Apache Tez~\cite{saha2015apache} can be seen as an open source implementation of the ideas proposed in Dryad.
+MPP databases \cite{dewitt1990gamma}, and recent open-source implementations like Apache Drill and Impala \cite{kornacker2015impala}, restrict their API to SQL variants.
+Similar to Flink, Apache Spark \cite{DBLP:conf/hotcloud/ZahariaCFSS10} is a data-processing framework that implements a DAG-based execution engine, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+In contrast, Flink is the only system that
+\begin{inparaenum}[i)]
+  \item supports an optimizer that can optimize DAG programs beyond SQL queries,
+  \item is able to perform very efficient iterative processing natively,
+  \item includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
+\end{inparaenum}
+
+\para{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems, such as StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
+Many of these systems are based on research in the database community in projects such as PSoup/Telegraph \cite{chandrasekaran2003psoup}, Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{gedik2008spade}.
+Most of the above systems are either
+\begin{inparaenum}[i)]
+  \item academic prototypes,
+  \item closed-source commercial products, or 
+  \item do not scale the computation horizontally on clusters of commodity servers.
+\end{inparaenum}
+Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide low level APIs and offer only at-least-once and at-most-once guarantees.
+MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics.
+To the best of our knowledge, Flink is the only open-source project that
+\begin{inparaenum}[i)]
+  \item offers high level APIs and processing libraries,
+  \item provides state management with exactly-once guarantees and
+  \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.
+\end{inparaenum}
+
+\vspace{-3mm}
 \section{Conclusion}
@@
 In this paper we presented Apache Flink, a platform that implements a universal dataflow engine designed to perform both stream and batch analytics.
-Flink's dataflow engine treats operator state and logical intermediate results as first class citizens, and is used by both the batch and a data stream APIs, with different parameters.
-The streaming API that is built on top of  Flink's streaming dataflow engine, provides the means to keep recoverable state and to partition, transform, and aggregate windows of a data stream.
+Flink's dataflow engine treats operator state and logical intermediate results as first-class citizens, and is used by both the batch and a data stream APIs, with different parameters.
+The streaming API that is built on top of  Flink's streaming dataflow engine provides the means to keep recoverable state and to partition, transform, and aggregate windows of a data stream.
 While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
+
 {
@@
 }
+
 \end{document}

### 2016-05-25 paper.tex
fixes in related work mainly, and other corrections

@@
 Apache Kafka).
-In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators.
+In practice, non-persistent sources can also be incorporated by keeping a write-ahead log within the state of the source operators.
 
@@
 Similar to Flink, Apache Spark \cite{DBLP:conf/hotcloud/ZahariaCFSS10} is a data-processing framework that implements a DAG-based execution engine, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
-In contrast, Flink is the only system that
+In contrast, Flink is the only system that incorporates
 \begin{inparaenum}[i)]
-  \item supports an optimizer that can optimize DAG programs beyond SQL queries,
-  \item is able to perform very efficient iterative processing natively,
-  \item includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
+  \item a distributed dataflow runtime that exploits pipelined streaming execution for batch and stream workloads,
+  \item exactly-once state consistency through lightweight checkpointing,
+  \item efficient, native iterative processing,
+  \item sophisticated window semantics, supporting out-of-order processing.
 \end{inparaenum}
 
-\para{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems, such as StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community in projects such as PSoup/Telegraph \cite{chandrasekaran2003psoup}, Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{gedik2008spade}.
+\para{Stream Processing.} There is a wealth of prior work on academic and commercial stream processing systems, such as SEEP, StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
+Many of these systems are based on research in the database community \cite{chandrasekaran2003psoup,abadi2005design,arasu2004stream,chandramouli2014trill,gedik2008spade,migliavacca2010seep}.
 Most of the above systems are either
@@
 \end{inparaenum}
-Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza, provide low level APIs and offer only at-least-once and at-most-once guarantees.
-MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics.
-To the best of our knowledge, Flink is the only open-source project that
+More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with some state consistency guarantees (e.g.
+at-least-once processing in Apache Storm and Samza).
+Notably, concepts such as ``out of order processing'' (OOP \cite{li2008out}) gained significant attraction and were adopted by Google's MillWheel~\cite{akidau2013millwheel} a preliminary version of the current cloud executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
+Millwheel served as a proof of concept for exactly-once low latency stream processing and OOP, thus, being very influential to the evolution of Flink.
+To the best of our knowledge, Flink is the only open-source project that:
 \begin{inparaenum}[i)]
   \item offers high level APIs and processing libraries,
-  \item provides state management with exactly-once guarantees and
+  \item provides consistent managed state with exactly-once guarantees and
   \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.

### 2016-05-25 paper.tex
add acknowledgements

@@
 \begin{abstract}
-Apache Flink is an open-source system for processing streaming and batch data.
+Apache Flink \footnote{The authors of this paper make no claim in being the sole inventors or implementers of the ideas behind Apache Flink that are presented, but rather a group of people that attempt to accurately document these concepts and their significance in writing.
+For a complete list see \autoref{sec:acknowledgements}} is an open-source system for processing streaming and batch data.
 Flink is built on the philosophy that many classes of data processing applications, including real-time analytics, continuous data pipelines, historic data processing (batch), and iterative algorithms (machine learning, graph analysis) can be expressed and executed as pipelined fault-tolerant dataflows.
@@
 
+\vspace{-4mm}
+\section{Acknowledgements}
+\vspace{-2mm}
+\label{sec:acknowledgements}
+The development of the Apache Flink project is overseen by a self-selected team of active contributors to the project.
+A Project Management Committee (PMC) guides the project's ongoing operations, including community development and product releases.
+At the current time of writing this, the list of Flink committers (excluding any affiliated authors) are : M\'arton Balassi, Ufuk Celebi, Gyula F\'ora, Alan Gates, Fabian Hueske, Vasia Kalavri, Aljoscha Krettek, Andra Lungu, Robert Metzger, Maximilian Michels, Chiwan Park, Till Rohrmann, Henry Saputra, Matthias J.
+Sax, Sebastian Schelter, Timo Walther, Daniel Warneke, ChengXiang Li, and Greg Hogan.
+In addition to these individuals, we want to acknowledge the broader Flink community of more than 180 contributors.
+
 %!TEX root = paper.tex

### 2016-05-25 paper.tex
Updates from ShareLaTeX

--- 2016-05-25 paper.tex
@@
 at-least-once processing in Apache Storm and Samza).
-Notably, concepts such as ``out of order processing'' (OOP \cite{li2008out}) gained significant attraction and were adopted by Google's MillWheel~\cite{akidau2013millwheel} a preliminary version of the current cloud executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
+Notably, concepts such as ``out of order processing'' (OOP \cite{li2008out}) gained significant attraction and were adopted by Google's MillWheel~\cite{akidau2013millwheel} a preliminary version of the commercial cloud executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
 Millwheel served as a proof of concept for exactly-once low latency stream processing and OOP, thus, being very influential to the evolution of Flink.

### 2016-05-25 paper.tex
symbols for authors

@@
 
-\author{Paris Carbone\textsuperscript{1} \\ Asterios Katsifodimos\textsuperscript{3} \\\\ \small{ \textsf{\small \textsuperscript{1}KTH \& SICS Sweden}} \\ \small \href{parisc,haridi@kth.se}{parisc,haridi@kth.se}
-\and Stefan Ewen\textsuperscript{2} \\ Volker Markl\textsuperscript{3} \\\\ \small \textsf{\textsuperscript{2}Data Artisans GmbH} \\ \small \href{first@data-artisans.de}{first@data-artisans.de}
-\and Seif Haridi\textsuperscript{1} \\ Kostas Tzoumas\textsuperscript{2} \\\\ \small \textsf{\textsuperscript{3}TU Berlin} \\ \small \href{fist.last@tu-berlin.de}{first.last@tu-berlin.de}
+\author{Paris Carbone\textsuperscript{$\dagger$} \\ Asterios Katsifodimos\textsuperscript{*} \\\\ \small{ \textsf{\small \textsuperscript{$\dagger$}KTH \& SICS Sweden}} \\ \small \href{parisc,haridi@kth.se}{parisc,haridi@kth.se}
+\and Stephan Ewen\textsuperscript{$\ddagger$} \\ Volker Markl\textsuperscript{*} \\\\ \small \textsf{\textsuperscript{$\ddagger$}Data Artisans GmbH} \\ \small \href{first@data-artisans.de}{first@data-artisans.de}
+\and Seif Haridi\textsuperscript{$\dagger$} \\ Kostas Tzoumas\textsuperscript{$\ddagger$} \\\\ \small \textsf{\textsuperscript{*}TU Berlin} \\ \small \href{fist.last@tu-berlin.de}{first.last@tu-berlin.de}
 %
@@
 \begin{abstract}
-Apache Flink \footnote{The authors of this paper make no claim in being the sole inventors or implementers of the ideas behind Apache Flink that are presented, but rather a group of people that attempt to accurately document these concepts and their significance in writing.
-For a complete list see \autoref{sec:acknowledgements}} is an open-source system for processing streaming and batch data.
+Apache Flink \footnote{The authors of this paper make no claim in being the sole inventors or implementers of the ideas behind Apache Flink, but rather a group of people that attempt to accurately document these concepts and their significance.
+Consult \autoref{sec:acknowledgements} for acknowledgements.} is an open-source system for processing streaming and batch data.
 Flink is built on the philosophy that many classes of data processing applications, including real-time analytics, continuous data pipelines, historic data processing (batch), and iterative algorithms (machine learning, graph analysis) can be expressed and executed as pipelined fault-tolerant dataflows.
@@
 \label{sec:intro}
-Data-stream processing (e.g, as exemplified by Complex Event Processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop ) were traditionally considered as two very different types of applications.
+Data-stream processing (e.g, as exemplified by Complex Event Processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop) were traditionally considered as two very different types of applications.
 They were programmed using different programming models and APIs, and were executed by different systems (e.g., dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft StreamInsight, or Streambase versus relational databases or execution engines for Hadoop, including Apache Spark, and Apache Drill).

### 2016-05-25 paper.tex
date and changes

@@
 \label{sec:related}
-There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.
+There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise the main systems below.
 
-\para{Batch Processing.} Apache Hadoop is the most popular open-source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
-Dryad~\cite{isard2007dryad} introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer}, which added a language and an SQL optimizer on top of it.
+\para{Batch Processing.} Apache Hadoop is one of the most popular open-source systems for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}.
+Dryad~\cite{isard2007dryad} introduced embedded user-defined functions in general DAG-based dataflows and was enriched by SCOPE~\cite{scopeOptimizer}, which a language and an SQL optimizer on top of it.
 Apache Tez~\cite{saha2015apache} can be seen as an open source implementation of the ideas proposed in Dryad.
@@
 
-\para{Stream Processing.} There is a wealth of prior work on academic and commercial stream processing systems, such as SEEP, StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams.
-Many of these systems are based on research in the database community \cite{chandrasekaran2003psoup,abadi2005design,arasu2004stream,chandramouli2014trill,gedik2008spade,migliavacca2010seep}.
+\para{Stream Processing.} There is a wealth of prior work on academic and commercial stream processing systems, such as SEEP, Naiad, Microsoft StreamInsight, and IBM Streams.
+Many of these systems are based on research in the database community \cite{chandrasekaran2003psoup,abadi2005design,arasu2004stream,chandramouli2014trill,gedik2008spade,migliavacca2010seep,murray2013naiad}.
 Most of the above systems are either
@@
 at-least-once processing in Apache Storm and Samza).
-Notably, concepts such as ``out of order processing'' (OOP \cite{li2008out}) gained significant attraction and were adopted by Google's MillWheel~\cite{akidau2013millwheel} a preliminary version of the commercial cloud executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
-Millwheel served as a proof of concept for exactly-once low latency stream processing and OOP, thus, being very influential to the evolution of Flink.
+Notably, concepts such as ``out of order processing'' (OOOP)~\cite{li2008out} gained significant attraction and were adopted by Google's MillWheel~\cite{akidau2013millwheel} a preliminary version of the commercial cloud executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
+Millwheel served as a proof of concept for exactly-once low latency stream processing and OOOP, thus, being very influential to the evolution of Flink.
 To the best of our knowledge, Flink is the only open-source project that:

### 2016-05-25 paper.tex
limited consistency

@@
 \end{inparaenum}
-More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with some state consistency guarantees (e.g.
+More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with limited state consistency guarantees (e.g.
 at-least-once processing in Apache Storm and Samza).

### 2016-05-25 paper.tex
weak guarantees

--- 2016-05-25 paper.tex
@@
 \end{inparaenum}
-More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with some state consistency guarantees (e.g.
+More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with weaker state consistency guarantees (e.g.
 at-least-once processing in Apache Storm and Samza).

### 2016-05-26 paper.tex
shrink to 11 pages
foobar
Updates from ShareLaTeX

@@
 
+\usepackage[
+backend=biber
+% style=alphabetic,
+% sorting=abbrv
+]{biblatex}
+
+\renewbibmacro{in:}{}
+
+\addbibresource{references.bib}
+\renewcommand{\bibfont}{\small} % or any other  appropriate font command
+
 % url references
@@
 \author{Paris Carbone\textsuperscript{$\dagger$} \\ Asterios Katsifodimos\textsuperscript{*} \\\\ \small{ \textsf{\small \textsuperscript{$\dagger$}KTH \& SICS Sweden}} \\ \small \href{parisc,haridi@kth.se}{parisc,haridi@kth.se}
-\and Stephan Ewen\textsuperscript{$\ddagger$} \\ Volker Markl\textsuperscript{*} \\\\ \small \textsf{\textsuperscript{$\ddagger$}Data Artisans GmbH} \\ \small \href{first@data-artisans.de}{first@data-artisans.de}
+\and Stephan Ewen\textsuperscript{$\ddagger$} \\ Volker Markl\textsuperscript{*} \\\\ \small \textsf{\textsuperscript{$\ddagger$}data Artisans} \\ \small \href{first@data-artisans.com}{first@data-artisans.com}
 \and Seif Haridi\textsuperscript{$\dagger$} \\ Kostas Tzoumas\textsuperscript{$\ddagger$} \\\\ \small \textsf{\textsuperscript{*}TU Berlin} \\ \small \href{fist.last@tu-berlin.de}{first.last@tu-berlin.de}
@@
 
-\vspace{-4mm}
-\section{Acknowledgements}
-\vspace{-2mm}
-\label{sec:acknowledgements}
-The development of the Apache Flink project is overseen by a self-selected team of active contributors to the project.
-A Project Management Committee (PMC) guides the project's ongoing operations, including community development and product releases.
-At the current time of writing this, the list of Flink committers (excluding any affiliated authors) are : M\'arton Balassi, Ufuk Celebi, Gyula F\'ora, Alan Gates, Fabian Hueske, Vasia Kalavri, Aljoscha Krettek, Andra Lungu, Robert Metzger, Maximilian Michels, Chiwan Park, Till Rohrmann, Henry Saputra, Matthias J.
-Sax, Sebastian Schelter, Timo Walther, Daniel Warneke, ChengXiang Li, and Greg Hogan.
-In addition to these individuals, we want to acknowledge the broader Flink community of more than 180 contributors.
-
 %!TEX root = paper.tex
@@
   \item exactly-once state consistency through lightweight checkpointing,
-  \item efficient, native iterative processing,
+  \item native iterative processing,
   \item sophisticated window semantics, supporting out-of-order processing.
@@
 
-\vspace{-3mm}
+\vspace{-4mm}
+\section{Acknowledgements}
+\vspace{-2mm}
+\label{sec:acknowledgements}
+The development of the Apache Flink project is overseen by a self-selected team of active contributors to the project.
+A Project Management Committee (PMC) guides the project's ongoing operations, including community development and product releases.
+At the current time of writing this, the list of Flink committers (excluding any affiliated authors) are : M\'arton Balassi, Ufuk Celebi, Gyula F\'ora, Alan Gates, Fabian Hueske, Vasia Kalavri, Aljoscha Krettek, Andra Lungu, Robert Metzger, Maximilian Michels, Chiwan Park, Till Rohrmann, Henry Saputra, Matthias J.
+Sax, Sebastian Schelter, Timo Walther, Daniel Warneke, ChengXiang Li, and Greg Hogan.
+In addition to these individuals, we want to acknowledge the broader Flink community of more than 180 contributors.
+
+\vspace{-4mm}
 \section{Conclusion}
+\vspace{-2mm}
 \label{sec:conclusions}
@@
 
-{
-\small
-\bibliographystyle{abbrv}
-\bibliography{references}
-}
+% {
+% \small
+% \bibliographystyle{abbrv}
+% \bibliography{references}
+% }
+\printbibliography
 

### 2016-05-26 paper.tex
Updates from ShareLaTeX

@@
 The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning.
-Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 170 open-source contributors at the time of this writing), and is used in production in several companies.
+Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 180 open-source contributors at the time of this writing), and is used in production in several companies.
 
@@
 MPP databases \cite{dewitt1990gamma}, and recent open-source implementations like Apache Drill and Impala \cite{kornacker2015impala}, restrict their API to SQL variants.
-Similar to Flink, Apache Spark \cite{DBLP:conf/hotcloud/ZahariaCFSS10} is a data-processing framework that implements a DAG-based execution engine, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches.
+Similar to Flink, Apache Spark \cite{DBLP:conf/hotcloud/ZahariaCFSS10} is a data-processing framework that implements a DAG-based execution engine, provides an SQL optimizer, performs driver-based iterations, and treats unbounded computation as micro-batches.
 In contrast, Flink is the only system that incorporates
@@
 at-least-once processing in Apache Storm and Samza).
-Notably, concepts such as ``out of order processing'' (OOOP)~\cite{li2008out} gained significant attraction and were adopted by Google's MillWheel~\cite{akidau2013millwheel} a preliminary version of the commercial cloud executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
-Millwheel served as a proof of concept for exactly-once low latency stream processing and OOOP, thus, being very influential to the evolution of Flink.
+Notably, concepts such as ``out of order processing'' (OOP)~\cite{li2008out} gained significant attraction and were adopted by  MillWheel~\cite{akidau2013millwheel}, Google's internal version of the later offered commercial executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
+Millwheel served as a proof of concept for exactly-once low latency stream processing and OOP, thus, being very influential to the evolution of Flink.
 To the best of our knowledge, Flink is the only open-source project that:
 \begin{inparaenum}[i)]
-  \item offers high level APIs and processing libraries,
-  \item provides consistent managed state with exactly-once guarantees and
-  \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.
+  \item supports event time and out-of-order event processing
+  \item provides consistent managed state with exactly-once guarantees
+  \item achieves high throughput and low latency, serving both batch and streaming
 \end{inparaenum}
@@
 A Project Management Committee (PMC) guides the project's ongoing operations, including community development and product releases.
-At the current time of writing this, the list of Flink committers (excluding any affiliated authors) are : M\'arton Balassi, Ufuk Celebi, Gyula F\'ora, Alan Gates, Fabian Hueske, Vasia Kalavri, Aljoscha Krettek, Andra Lungu, Robert Metzger, Maximilian Michels, Chiwan Park, Till Rohrmann, Henry Saputra, Matthias J.
-Sax, Sebastian Schelter, Timo Walther, Daniel Warneke, ChengXiang Li, and Greg Hogan.
+At the current time of writing this, the list of Flink committers are : M\'arton Balassi, Paris Carbone, Ufuk Celebi, Stephan Ewen, Gyula F\'ora, Alan Gates, Greg Hogan, Fabian Hueske, Vasia Kalavri, Aljoscha Krettek, ChengXiang Li, Andra Lungu, Robert Metzger, Maximilian Michels, Chiwan Park, Till Rohrmann, Henry Saputra, Matthias J.
+Sax, Sebastian Schelter, Kostas Tzoumas, Timo Walther and Daniel Warneke.
 In addition to these individuals, we want to acknowledge the broader Flink community of more than 180 contributors.
@@
 The streaming API that is built on top of  Flink's streaming dataflow engine provides the means to keep recoverable state and to partition, transform, and aggregate windows of a data stream.
-While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution with a novel query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
+While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution using a query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
 

### 2016-05-26 paper.tex
Updates from ShareLaTeX

@@
 Flink is built on the philosophy that many classes of data processing applications, including real-time analytics, continuous data pipelines, historic data processing (batch), and iterative algorithms (machine learning, graph analysis) can be expressed and executed as pipelined fault-tolerant dataflows.
-In this paper, we present Flink's architecture and expand on how a (seemingly diverse) set of use cases can be unified under the same execution model.
+In this paper, we present Flink's architecture and expand on how a (seemingly diverse) set of use cases can be unified under a single execution model.
 \end{abstract}
@@
 \label{sec:intro}
-Data-stream processing (e.g, as exemplified by Complex Event Processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop) were traditionally considered as two very different types of applications.
-They were programmed using different programming models and APIs, and were executed by different systems (e.g., dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft StreamInsight, or Streambase versus relational databases or execution engines for Hadoop, including Apache Spark, and Apache Drill).
+Data-stream processing (e.g., as exemplified by complex event processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop) were traditionally considered as two very different types of applications.
+They were programmed using different programming models and APIs, and were executed by different systems (e.g., dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft StreamInsight, or Streambase versus relational databases or execution engines for Hadoop, including Apache Spark and Apache Drill).
 Traditionally, batch data analysis made up for the lion's share of the use cases, data sizes, and market, while streaming data analysis mostly served specialized applications.
 
-It is becoming more and more apparent, however, that a huge number of today's large-scale data processing use cases handle data that is, in reality, continuously produced over time.
+It is becoming more and more apparent, however, that a huge number of today's large-scale data processing use cases handle data that is, in reality, produced continuously over time.
 These continuous streams of data come for example from web logs, application logs, sensors, or as changes to application state in databases (transaction log records).
 Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of data production.
-Instead, data records are (often artificially) batched into static data sets (for example hourly, daily, or monthly chunks) and then processed in a time-agnostic fashion.
+Instead, data records are (often artificially) batched into static data sets (e.g., hourly, daily, or monthly chunks) and then processed in a time-agnostic fashion.
 Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline.
@@
 The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning.
-Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 180 open-source contributors at the time of this writing), and is used in production in several companies.
+Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (consisting of over 180 open-source contributors as of the time of this writing), and is used in production in several companies.
 
 \vspace{2mm}
-\noindent The contributions of this paper are the following:\vspace{-2mm}
+\noindent The contributions of this paper are as follows:\vspace{-2mm}
 \begin{itemize}
@@
 	\vspace{-3mm}
-	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault-tolerant streaming dataflows (\autoref{sec:execution}).
+	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault-tolerant streaming dataflows (in \autoref{sec:execution}).
 	\vspace{-3mm}
-	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:streaming}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
+	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (in \autoref{sec:streaming}), as well as a full-fledged batch processor (in \autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
 \end{itemize}
@@
 
-In this section we lay out the architecture of Flink as a software stack, and as a distributed system.
+In this section we lay out the architecture of Flink as a software stack and as a distributed system.
 While Flink's stack of APIs continues to grow, we can distinguish four main layers: deployment, core, APIs, and libraries.
@@
 As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing.
-On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML (machine learning), Gelly (graph processing), and Table (SQL-like operations)).
+On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs, currently, FlinkML for machine learning, Gelly  for graph processing and Table for SQL-like operations.
 
-As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the Job Manager, and one or more Task Managers.
+As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the Job Manager, and at least one Task Manager.
 The client takes the program code, transforms it to a dataflow graph, and submits that to the JobManager.
 This transformation phase also examines the data types (schema) of the data exchanged between operators and creates serializers and other type/schema specific code.
-DataSet programs additionally go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in \autoref{sec:batch}).
+DataSet programs additionally go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (for more details see \autoref{sec:batch}).
 
@@
 			\vspace{-3mm}
-			\caption{Flink's software stack.}
+			\caption{The Flink software stack.}
 			\vspace{-3mm}
@@
     			\vspace{-2.5mm}
-    			\caption{Flink's process model.}
+    			\caption{The Flink process model.}
     			\label{fig:process-model}
@@
 \subsection{Dataflow Graphs}
-The dataflow graph, as depicted in \autoref{fig:dataflow}, is a directed acyclic graph (DAG) that consists of (i) stateful operators, and (ii) data streams that represent data produced by an operator and are available for consumption by operators.
+The dataflow graph as depicted in \autoref{fig:dataflow} is a directed acyclic graph (DAG) that consists of: (i) stateful operators  and (ii) data streams that represent data produced by an operator and are available for consumption by operators.
 Since dataflow graphs are executed in a data-parallel fashion, operators are parallelized into one or more parallel instances called \emph{subtasks} and streams are split into one or more \emph{stream partitions} (one partition per producing subtask).
-The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions.
+The stateful operators, which may be stateless as a special case implement all of the processing logic  (e.g., filters, hash joins and stream window functions).
 Many of these operators are implementations of textbook versions of well known algorithms.
-\autoref{sec:streaming} provides details on the implementation of windowing operators.
+In \autoref{sec:streaming}, we provide details on the implementation of windowing operators.
 Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, and merge.
@@
 Flink uses pipelined streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.
-\emph{Blocking streams}, on the other hand, are applicable to bounded data streams.
+\emph{Blocking streams} on the other hand are applicable to bounded data streams.
 A blocking stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages.
 Blocking streams naturally require more memory, frequently spill to secondary storage, and do not propagate backpressure.
-They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as sort-merge joins, may cause distributed deadlocks.
+They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators, such as sort-merge joins may cause distributed deadlocks.
 
 \para{Balancing Latency and Throughput.} Flink's data-exchange mechanisms are implemented around the exchange of buffers.
-When a data record is ready on the producer side, it is serialized split into one or more buffers (a buffer can also fit multiple records) that can be forwarded to consumers.
-A buffer is sent to a consumer either i) as soon as it is full, or ii) when a timeout condition is reached.
+When a data record is ready on the producer side, it is serialized and split into one or more buffers (a buffer can also fit multiple records) that can be forwarded to consumers.
+A buffer is sent to a consumer either i) as soon as it is full or ii) when a timeout condition is reached.
 This enables Flink to achieve high throughput by setting the size of buffers to a high value (e.g., a few kilobytes), as well as low latency by setting the buffer timeout to a low value (e.g., a few milliseconds).
-\autoref{fig:latency-throughput} shows the effect of buffer-timeouts to the throughput and latency of delivering records in a simple streaming grep job on 30 machines (120 cores).
-Flink can achieve an observable 99$^{\mbox{\small{th}}}$-percentile latency of 20 milliseconds.
+\autoref{fig:latency-throughput} shows the effect of buffer-timeouts on the throughput and latency of delivering records in a simple streaming grep job on 30 machines (120 cores).
+Flink can achieve an observable 99$^{\mbox{\small{th}}}$-percentile latency of 20ms.
 The corresponding throughput is 1.5 million events per second.
 As we increase the buffer timeout, we see an increase in latency with an increase in throughput, until full throughput is reached (i.e., buffers fill up faster than the timeout expiration).
-At a buffer timeout of 50 milliseconds, the cluster reaches a throughput of more than 80 million events per second with a 99$^{\mbox{\small{th}}}$-percentile latency of 50 milliseconds.
+At a buffer timeout of 50ms, the cluster reaches a throughput of more than 80 million events per second with a 99$^{\mbox{\small{th}}}$-percentile latency of 50ms.
 
@@
 \begin{itemize}
-\item \textit{Checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
-\item \textit{Watermarks} signaling the progress of event-time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
-\item \textit{Iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}).
+\item \textit{Checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (discussed in \autoref{sec:fault-tolerance}). \vspace{-3mm}
+\item \textit{Watermarks} signaling the progress of event-time within a stream partition (discussed in \autoref{sec:streaming-time}). \vspace{-3mm}
+\item \textit{Iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (discussed in \autoref{sec:batch-iterations}).
 \vspace{-1mm}
@@
 However, operators receiving more than one stream partition merge the streams in arrival order, in order to keep up with the streams' rates and avoid back pressure.
-As a result, streaming dataflows in Flink do not provide ordering guarantees after any form of repartitioning or broadcasting, and the responsibility of dealing with out-of-order records is left to the operator implementation.
-We found that this arrangement gives the most efficient design, as most operators do not require deterministic order (e.g., hash-joins, maps, etc.), and operators that need to compensate for out-of-order arrivals (such as event-time windows) can do that more efficiently as part of the operator logic.
+As a result, streaming dataflows in Flink do not provide ordering guarantees after any form of repartitioning or broadcasting and the responsibility of dealing with out-of-order records is left to the operator implementation.
+We found that this arrangement gives the most efficient design, as most operators do not require deterministic order (e.g., hash-joins, maps), and operators that need to compensate for out-of-order arrivals, such as event-time windows can do that more efficiently as part of the operator logic.
 
@@
 The general assumption the system makes to effectively provide these guarantees is that the data sources are persistent and replayable.
-Examples of such sources are files and durable message queues (e.g.
-Apache Kafka).
+Examples of such sources are files and durable message queues (e.g., Apache Kafka).
 In practice, non-persistent sources can also be incorporated by keeping a write-ahead log within the state of the source operators.
@@
 The possibly unbounded nature of a data stream makes re-computation upon recovery impractical, as possibly months of computation will need to be replayed for a long-running job.
-To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams, at regular intervals.
+To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.
 
@@
 In essence, the snapshot of all operators should refer to the same logical time in the computation.
-The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}.
-Barriers are control records injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.
+The mechanism used in Flink is called Asynchronous Barrier Snapshotting (ABS~\cite{carbone2015lightweight}).
+Barriers are control records injected into the input streams that correspond to a logical time and logically separate the stream to the part whose effects will be included in the current snapshot and the part that will be snapshotted later.
 
@@
 An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received.
-Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS).
+Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (e.g., the storage backend can be an external system such as HDFS).
 Once the state has been backed up, the operator forwards the barrier downstream.
@@
 ABS bears resemblances to the Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}.
-However, because of the DAG structure of a Flink program, ABS does not need to checkpoint in-flight records but solely relies on the aligning phase to apply all their effects to the operator states.
+However, because of the DAG structure of a Flink program, ABS does not need to checkpoint in-flight records, but solely relies on the aligning phase to apply all their effects to the operator states.
 This guarantees that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).
 
-Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier for which there is a snapshot.
+Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot and restarts the input streams starting from the latest barrier for which there is a snapshot.
 The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers.
@@
 \item it guarantees exactly-once state updates without ever pausing the computation
-\item it is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval and
+\item it is completely decoupled from other forms of control messages, (e.g., by events that trigger the computation of windows and thereby do not restrict the windowing mechanism to multiples of the checkpoint interval) and
 \item it is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc., depending on the larger environment in which Flink is used.
@@
 
-Incremental processing and iterations are crucial for applications such as graph processing and machine learning.
-Support for iterations in data-parallel processing platforms typically relies on submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, or feedback edges \cite{murray2013naiad}.
+Incremental processing and iterations are crucial for applications, such as graph processing and machine learning.
+Support for iterations in data-parallel processing platforms typically relies on submitting a new job for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10} or feedback edges \cite{murray2013naiad}.
 Iterations in Flink are implemented as  \emph{iteration steps}, special operators that themselves can contain an execution graph (\autoref{fig:iterations}).
-To maintain the (acyclic) DAG-based runtime and scheduler, Flink allows for iteration ``head'' and ``tail'' tasks that are \emph{implicitly} connected with feedback edges.
+To maintain the DAG-based runtime and scheduler, Flink allows for iteration ``head'' and ``tail'' tasks that are \emph{implicitly} connected with feedback edges.
 The role of these tasks is to establish an active feedback channel to the iteration step and provide coordination for processing data records in transit within this feedback channel.
-Coordination is needed for implementing any type of structured parallel iteration model such as the Bulk Synchronous Parallel (BSP) model and is implemented using control event.
-We explain how iterations are implemented in the DataStream and DataSet APIs in \autoref{sec:stream-iterations} and  \autoref{sec:batch-iterations} respectively.
+Coordination is needed for implementing any type of structured parallel iteration model, such as the Bulk Synchronous Parallel (BSP) model and is implemented using control event.
+We explain how iterations are implemented in the DataStream and DataSet APIs in \autoref{sec:stream-iterations} and  \autoref{sec:batch-iterations}, respectively.
 
@@
 
-Flink's DataStream API implements a full stream-analytics framework on top of Flink's runtime, including the mechanisms to manage time (including out-of-order event processing), defining windows, and maintaining and updating user-defined state.
+Flink's DataStream API implements a full stream-analytics framework on top of Flink's runtime, including the mechanisms to manage time such as out-of-order event processing, defining windows, and maintaining and updating user-defined state.
 The streaming API is based on the notion of a DataStream, a (possibly unbounded) immutable collection of elements of a given type.
@@
 \label{sec:streaming-time}
-Flink distinguishes between two notions of time: i) event-time, that  stands for the time that an event was generated at its origin (e.g., the time that a sensor or a mobile device generated a signal) and ii) processing-time, which is the wall-clock time of the machine that is processing the data.
+Flink distinguishes between two notions of time: i) event-time, which denotes the time when an event originates (e.g., the timestamp associated with a  signal arising from a sensor, such as a mobile device) and ii) processing-time, which is the wall-clock time of the machine that is processing the data.
 
@@
 In the case of time progress for example, a watermark includes a time attribute $t$ indicating that all events lower than $t$ have already entered an operator.
-The watermarks aid the execution engine in processing events in the correct event order and serialize operations (such as window computations) via a unified measure of progress.
+The watermarks aid the execution engine in processing events in the correct event order and serialize operations, such as window computations via a unified measure of progress.
 
@@
 Operators decide how they react to watermarks.
-Simple operations such as map or filter just forward the watermarks they receive, while more complex operators that do calculation based on the watermarks (e.g., event-time windows) first compute results triggered by a watermark and then forward it.
-If an operation has more than one input, the system only forwards the minimum of the incoming watermarks to the operator, ensuring correct results.
+Simple operations, such as map or filter just forward the watermarks they receive, while more complex operators that do calculations based on watermarks (e.g., event-time windows) first compute results triggered by a watermark and then forward it.
+If an operation has more than one input, the system only forwards the minimum of the incoming watermarks to the operator thereby ensuring correct results.
 
-Flink programs that are based on processing-time rely on the machine clocks, and hence a less reliable notion of time, which can lead to inconsistent replays upon recovery.
+Flink programs that are based on processing-time rely on local machine clocks, and hence possess a less reliable notion of time, which can lead to inconsistent replays upon recovery.
 However, they exhibit lower latency.
@@
 Flink includes a third notion of time as a special case of event-time called \emph{ingestion-time}, which is the time that events enter Flink.
-This provides option lower processing latency than event-time, and more accurate results compared to processing-time.
+That achieves a lower processing latency than event-time and leads to more accurate results in comparison to processing-time.
 
@@
     \vspace{-6mm}
-  \caption{The iteration model of Apache Flink}
+  \caption{The iteration model of Apache Flink.}
   \vspace{-2mm}
@@
 While most operators in Flink's DataStream API look like functional, side-effect-free operators, they provide support for efficient stateful computations.
-State is critical to many applications, such as machine-learning model building, graph analysis, user session handling and window aggregations.
+State is critical to many applications, such as machine-learning model building, graph analysis, user session handling, and window aggregations.
 There is a plethora of different types of states depending on the use case.
-For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine-learning applications.
+For example, the state can be something as simple as a counter or a sum or more complex, such as a classification tree or a large sparse matrix often used in machine-learning applications.
 Stream windows are  stateful operators that assign records to continuously updated buckets kept in memory as part of the operator state.
 
-In Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  ii)~an operator-state abstraction for declaring partitioned key-value states and the state's associated operations.
-Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in streaming applications.
-Flink's checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
+In Flink state is made explicit and is incorporated in the API by providing: i)~operator interfaces or annotations to statically register explicit local variables within the scope of an operator and  ii)~an operator-state abstraction for declaring partitioned key-value states and their associated operations.
+Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, thereby allowing highly flexible custom state management in streaming applications.
+Flink's checkpointing mechanism (discussed in \autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.
 \vspace{-8mm}
@@
 \subsection{Stream Windows}
-Incremental computations over unbounded streams are often evaluated over continously evolving logical views, called windows.
+Incremental computations over unbounded streams are often evaluated over continuously evolving logical views, called windows.
 Apache Flink incorporates windowing within a stateful operator that is configured via a flexible declaration composed out of three core functions: a window \textit{assigner} and optionally a \textit{trigger} and an \textit{evictor}.
-All three functions can be selected among a pool of common predefined implementations (e.g.
-sliding time windows) or can be explicitly defined by the user as UDFs.
+All three functions can be selected among a pool of common predefined implementations (e.g., sliding time windows) or can be explicitly defined by the user (i.e., user-defined functions).
 
@@
 Note that Flink's windowing capabilities incorporate out-of-order processing seamlessly, similarly to Google Cloud Dataflow \cite{akidau2015dataflow} and, in principle, subsume these windowing models.
-For example, below is a window definition with a range of 6 seconds that slides every 2 seconds (Assigner).
-The window results are computed once the watermark passes the end of the window (Trigger).
+For example, below is a window definition with a range of 6 seconds that slides every 2 seconds (the assigner).
+The window results are computed once the watermark passes the end of the window (the trigger).
 
@@
 \noindent A global window creates a single logical group.
-The following example defines a global window (assigner) that invokes the operation on every 1000 events (trigger) while keeping the last 100 elements (evictor).
+The following example defines a global window (i.e., the assigner) that invokes the operation on every 1000 events (i.e., the trigger) while keeping the last 100 elements (i.e., the evictor).
 
@@
 
-Note that if the stream above is partitioned on a key before  windowing, the window operation above is \textit{local} and thus, it does not require coordination between machines.
+Note that if the stream above is partitioned on a key before  windowing, the window operation above is \textit{local} and thus does not require coordination between workers.
 This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
@@
 \label{sec:stream-iterations}
-Loops in streams are essential for several applications such as incrementally building and training machine learning models, reinforcement learning and graph approximations \cite{feigenbaum2005graph,chandramouli2009fly}.
+Loops in streams are essential for several applications, such as incrementally building and training machine learning models, reinforcement learning and graph approximations \cite{feigenbaum2005graph,chandramouli2009fly}.
 In most such cases, feedback loops need no coordination.
@@
 A bounded data set is a special case of an unbounded data stream.
-Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by FlinkFlink'ss features that we presented above.
-However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and ii) programs that process bounded data sets are amenable to additional optimizations, more efficient book-keeping for fault-tolerance, and staged scheduling.
+Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by Flink's features that were presented above.
+However, i) the syntax (i.e., the API for batch computation) can be simplified (e.g., there is no need for artificial global window definitions) and ii) programs that process bounded data sets are amenable to additional optimizations, more efficient book-keeping for fault-tolerance, and staged scheduling.
 
@@
 The runtime is not aware of whether an operator is blocking or not.
-These operators use managed memory provided by Flink (either on or off the JVM heap), and can spill to disk if their inputs exceed their memory bounds.\vspace{-3mm}
-	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc).\vspace{-3mm}
-	\item A query optimization layer transforms a DataSet program to an efficient executable.\vspace{-3mm}
+These operators use managed memory provided by Flink (either on or off the JVM heap) and can spill to disk if their inputs exceed their memory bounds.\vspace{-3mm}
+	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure and transformations on DataSets (e.g., joins, aggregations, iterations).\vspace{-3mm}
+	\item A query optimization layer transforms a DataSet program into an efficient executable.\vspace{-3mm}
 \end{itemize}
 
-\noindent We provide more details on these aspects below.
+\noindent Below we describe these aspects in greater detail.
 
@@
 
-\subsection{Memory Management} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into memory segments.
+\subsection{Memory Management} Building on database technology, Flink serializes data into memory segments, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records.
 Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the serialization and deserialization overhead at a minimum and partially spilling data to disk when needed.
@@
 For instance, in the case of a BSP execution, iteration-control events mark the beginning and the end of supersteps in an iterative computation.
-Finally, Flink introduces further novel optimisation techniques such as the concept of \emph{delta} iterations \cite{DBLP:journals/pvldb/EwenTKM12}, which can exploit sparse computational dependencies (already part of Gelly, Flink's Graph API).
+Finally, Flink introduces further novel optimisation techniques such as the concept of \emph{delta} iterations \cite{DBLP:journals/pvldb/EwenTKM12}, which can exploit sparse computational dependencies Delta iterations are already exploited by Gelly, Flink's Graph API.
 
@@
 \label{sec:related}
-There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise the main systems below.
+Today, there is a wealth of engines for distributed batch and stream analytical processing. We categorise the main systems below.
 
@@
 \end{inparaenum}
-More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with weaker state consistency guarantees (e.g.
-at-least-once processing in Apache Storm and Samza).
+More recent approaches in data streaming enable horizontal scalability and compositional dataflow operators with weaker state consistency guarantees (e.g., at-least-once processing in Apache Storm and Samza).
 Notably, concepts such as ``out of order processing'' (OOP)~\cite{li2008out} gained significant attraction and were adopted by  MillWheel~\cite{akidau2013millwheel}, Google's internal version of the later offered commercial executor of Apache Beam/Google Dataflow~\cite{akidau2015dataflow}.
@@
 \label{sec:conclusions}
-In this paper we presented Apache Flink, a platform that implements a universal dataflow engine designed to perform both stream and batch analytics.
-Flink's dataflow engine treats operator state and logical intermediate results as first-class citizens, and is used by both the batch and a data stream APIs, with different parameters.
-The streaming API that is built on top of  Flink's streaming dataflow engine provides the means to keep recoverable state and to partition, transform, and aggregate windows of a data stream.
-While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution using a query optimizer, and by implementing blocking operators that gracefully spill to disk in the absence of memory.
+In this paper, we presented Apache Flink, a platform that implements a universal dataflow engine designed to perform both stream and batch analytics.
+Flink's dataflow engine treats operator state and logical intermediate results as first-class citizens and is used by both the batch and a data stream APIs with different parameters.
+The streaming API that is built on top of  Flink's streaming dataflow engine provides the means to keep recoverable state and to partition, transform, and aggregate data stream windows.
+While batch computations are, in theory, a special case of a streaming computations, Flink treats them specially, by optimizing their execution using a query optimizer and by implementing blocking operators that gracefully spill to disk in the absence of memory.
 

### 2016-05-30 paper.tex
Updates from ShareLaTeX

@@
 \and Stephan Ewen\textsuperscript{$\ddagger$} \\ Volker Markl\textsuperscript{*} \\\\ \small \textsf{\textsuperscript{$\ddagger$}data Artisans} \\ \small \href{first@data-artisans.com}{first@data-artisans.com}
-\and Seif Haridi\textsuperscript{$\dagger$} \\ Kostas Tzoumas\textsuperscript{$\ddagger$} \\\\ \small \textsf{\textsuperscript{*}TU Berlin} \\ \small \href{fist.last@tu-berlin.de}{first.last@tu-berlin.de}
+\and Seif Haridi\textsuperscript{$\dagger$} \\ Kostas Tzoumas\textsuperscript{$\ddagger$} \\\\ \small \textsf{\textsuperscript{*}TU Berlin \& DFKI} \\ \small \href{fist.last@tu-berlin.de}{first.last@tu-berlin.de}
 %

### 2016-06-02 paper.tex
11 pages, before last check

@@
 \usepackage{listings}
+
+\graphicspath{{figs/}}
 
@@
 
-\usepackage[
-backend=biber
-% style=alphabetic,
-% sorting=abbrv
-]{biblatex}
+% \usepackage[
+% % backend=biber
+% % style=alphabetic,
+% % sorting=abbrv
+% ]{biblatex}
 
-\renewbibmacro{in:}{}
+% \renewbibmacro{in:}{}
 
-\addbibresource{references.bib}
-\renewcommand{\bibfont}{\small} % or any other  appropriate font command
+% \addbibresource{references.bib}
+% \renewcommand{\bibfont}{\small} % or any other  appropriate font command
 
@@
 \label{sec:intro}
+\vspace{-3mm}
 Data-stream processing (e.g., as exemplified by complex event processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop) were traditionally considered as two very different types of applications.
@@
 
-%!TEX root = paper.tex
-
-% \vspace{-mm}
+% \vspace{-3mm}
 \section{System Architecture}
 \label{sec:architecture}
-
+\vspace{-2mm}
 In this section we lay out the architecture of Flink as a software stack and as a distributed system.
@@
               \centering
-			\includegraphics[width=.88\textwidth]{figs/stack}
+			\includegraphics[width=.88\textwidth,natheight=590,natwidth=500]{figs/stack.pdf}
 			\vspace{-3mm}
@@
           \begin{figure}[H]
-				\includegraphics[width=.95\textwidth]{figs/architecture.pdf}
+				\includegraphics[width=.95\textwidth,natheight=317,natwidth=204]{figs/architecture.pdf}
     			\vspace{-2.5mm}
@@
 
-\vspace{-3mm}
+% \vspace{-3mm}
 \section{The Common Fabric: Streaming Dataflows}
 \label{sec:execution}
+\vspace{-3mm}
 
@@
 \subsection{Dataflow Graphs}
+\vspace{-2mm}
 The dataflow graph as depicted in \autoref{fig:dataflow} is a directed acyclic graph (DAG) that consists of: (i) stateful operators  and (ii) data streams that represent data produced by an operator and are available for consumption by operators.
@@
         \centering
-        \includegraphics[width=.999\textwidth]{figs/dataflow}
+        \includegraphics[width=.999\textwidth,natheight=960,natwidth=540]{figs/dataflow.pdf}
         \vspace{-5mm}
@@
           \begin{figure}[H]
-				\includegraphics[width=.99\textwidth]{figs/latency-throughput.pdf}
+				\includegraphics[width=.99\textwidth,natheight=590,natwidth=500]{figs/latency-throughput.pdf}
 				\vspace{-7mm}
@@
 \label{sec:fault-tolerance}
-
+\vspace{-2mm}
 Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution.
@@
 	\centering
-  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
+  	\includegraphics[width=.75\textwidth,natheight=419,natwidth=159]{figs/snaps.pdf}
   	\vspace{-6mm}
@@
 
+\vspace{-3mm}
 \subsection{Iterative Dataflows}
 \label{sec:iterations}
-
+\vspace{-2mm}
 Incremental processing and iterations are crucial for applications, such as graph processing and machine learning.
@@
 
-\vspace{-4mm}
+% \vspace{-4mm}
 \section{Stream Analytics on Top of Dataflows}
 \label{sec:streaming}
+\vspace{-3mm}
 
@@
 \label{sec:streaming-time}
+\vspace{-2mm}
 Flink distinguishes between two notions of time: i) event-time, which denotes the time when an event originates (e.g., the timestamp associated with a  signal arising from a sensor, such as a mobile device) and ii) processing-time, which is the wall-clock time of the machine that is processing the data.
@@
   \centering
-    \includegraphics[width=.75\textwidth]{figs/iterations.pdf}
+    \includegraphics[width=.75\textwidth,natwidth=498,natheight=145]{figs/iterations.pdf}
     \vspace{-6mm}
@@
 \subsection{Stateful Stream Processing}
+\vspace{-2mm}
 While most operators in Flink's DataStream API look like functional, side-effect-free operators, they provide support for efficient stateful computations.
@@
 \subsection{Stream Windows}
+\vspace{-2mm}
 Incremental computations over unbounded streams are often evaluated over continuously evolving logical views, called windows.
@@
 
+\vspace{-3mm}
 \subsection{Asynchronous Stream Iterations}
 \label{sec:stream-iterations}
+\vspace{-2mm}
 Loops in streams are essential for several applications, such as incrementally building and training machine learning models, reinforcement learning and graph approximations \cite{feigenbaum2005graph,chandramouli2009fly}.
@@
 
-\vspace{-3mm}
+% \vspace{-4mm}
 \section{Batch Analytics on Top of Dataflows}
 \label{sec:batch}
-
+\vspace{-3mm}
 A bounded data set is a special case of an unbounded data stream.
@@
 
+\vspace{-3mm}
 \subsection{Query Optimization} 
+\vspace{-2mm}
 Flink's optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting-property propagation.
@@
 
-\subsection{Memory Management} Building on database technology, Flink serializes data into memory segments, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records.
+\vspace{-3mm}
+\subsection{Memory Management} 
+\vspace{-2mm}
+Building on database technology, Flink serializes data into memory segments, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records.
 Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the serialization and deserialization overhead at a minimum and partially spilling data to disk when needed.
@@
 
+\vspace{-3mm}
 \subsection{Batch Iterations}
 \label{sec:batch-iterations}
+\vspace{-2mm}
 Iterative graph analytics, parallel gradient descent and optimisation techniques have been implemented in the past on top of Bulk Synchronous Parallel (BSP) and Stale Synchronous Parallel (SSP) models, among others.
@@
 %!TEX root = paper.tex
-\vspace{-4mm}
+% \vspace{-4mm}
 \section{Related work}
@@
 
-\vspace{-4mm}
+\vspace{-5mm}
 \section{Conclusion}
@@
 
+% \begin{thebibliography}{10}
+% \itemsep=1pt
+% \begin{small}
+
+\vspace{-3mm}
+\begin{thebibliography}{10}
+\itemsep=1pt
+\begin{small}
+
+\bibitem{abadi2005design}
+D.~J. Abadi, Y.~Ahmad, M.~Balazinska, U.~Cetintemel, M.~Cherniack, J.-H. Hwang,
+  W.~Lindner, A.~Maskey, A.~Rasin, E.~Ryvkina, et~al.
+\newblock The design of the Borealis stream processing engine.
+\newblock {\em CIDR}, 2005.
+
+\bibitem{akidau2013millwheel}
+T.~Akidau, A.~Balikov, K.~Bekiro{\u{g}}lu, S.~Chernyak, J.~Haberman, R.~Lax,
+  S.~McVeety, D.~Mills, P.~Nordstrom, and S.~Whittle.
+\newblock Millwheel: fault-tolerant stream processing at Internet scale.
+\newblock {\em PVLDB}, 2013.
+
+\bibitem{akidau2015dataflow}
+T.~Akidau, R.~Bradshaw, C.~Chambers, S.~Chernyak, R.~J.
+  Fern{\'a}ndez-Moctezuma, R.~Lax, S.~McVeety, D.~Mills, F.~Perry, E.~Schmidt,
+  et~al.
+\newblock The dataflow model: a practical approach to balancing correctness,
+  latency, and cost in massive-scale, unbounded, out-of-order data processing.
+\newblock {\em PVLDB}, 2015.
+
+\bibitem{stratosphere}
+A.~Alexandrov, R.~Bergmann, S.~Ewen, J.-C. Freytag, F.~Hueske, A.~Heise,
+  O.~Kao, M.~Leich, U.~Leser, V.~Markl, F.~Naumann, M.~Peters, A.~Rheinlaender,
+  M.~J. Sax, S.~Schelter, M.~Hoeger, K.~Tzoumas, and D.~Warneke.
+\newblock The stratosphere platform for big data analytics.
+\newblock {\em VLDB Journal}, 2014.
+
+\bibitem{arasu2004stream}
+A.~Arasu, B.~Babcock, S.~Babu, J.~Cieslewicz, M.~Datar, K.~Ito, R.~Motwani,
+  U.~Srivastava, and J.~Widom.
+\newblock Stream: The stanford data stream management system.
+\newblock {\em Technical Report}, 2004.
+
+\bibitem{DBLP:journals/pvldb/BuHBE10}
+Y.~Bu, B.~Howe, M.~Balazinska, and M.~D. Ernst.
+\newblock {HaLoop: Efficient Iterative Data Processing on Large Clusters}.
+\newblock {\em PVLDB}, 2010.
+
+\bibitem{carbone2015lightweight}
+P.~Carbone, G.~F{\'o}ra, S.~Ewen, S.~Haridi, and K.~Tzoumas.
+\newblock Lightweight asynchronous snapshots for distributed dataflows.
+\newblock {\em arXiv:1506.08603}, 2015.
+
+\bibitem{chandramouli2014trill}
+B.~Chandramouli, J.~Goldstein, M.~Barnett, R.~DeLine, D.~Fisher, J.~C. Platt,
+  J.~F. Terwilliger, and J.~Wernsing.
+\newblock Trill: a high-performance incremental query processor for diverse
+  analytics.
+\newblock {\em PVLDB}, 2014.
+
+\bibitem{chandramouli2009fly}
+B.~Chandramouli, J.~Goldstein, and D.~Maier.
+\newblock On-the-fly progress detection in iterative stream queries.
+\newblock {\em PVLDB}, 2009.
+
+\bibitem{chandrasekaran2003psoup}
+S.~Chandrasekaran and M.~J. Franklin.
+\newblock Psoup: a system for streaming queries over streaming data.
+\newblock {\em VLDB Journal}, 2003.
+
+\bibitem{chandy1985distributed}
+K.~M. Chandy and L.~Lamport.
+\newblock Distributed snapshots: determining global states of distributed
+  systems.
+\newblock {\em ACM TOCS}, 1985.
+
+\bibitem{DBLP:journals/cacm/DeanG08}
+J.~Dean~et al.
+\newblock {MapReduce: simplified data processing on large clusters}.
+\newblock {\em Communications of the {ACM}}, 2008.
+
+\bibitem{dewitt1990gamma}
+D.~J. DeWitt, S.~Ghandeharizadeh, D.~Schneider, A.~Bricker, H.-I. Hsiao,
+  R.~Rasmussen, et~al.
+\newblock The gamma database machine project.
+\newblock {\em IEEE TKDE}, 1990.
+
+\bibitem{DBLP:journals/pvldb/EwenTKM12}
+S.~Ewen, K.~Tzoumas, M.~Kaufmann, and V.~Markl.
+\newblock {Spinning Fast Iterative Data Flows}.
+\newblock {\em {PVLDB}}, 2012.
+
+\bibitem{feigenbaum2005graph}
+J.~Feigenbaum, S.~Kannan, A.~McGregor, S.~Suri, and J.~Zhang.
+\newblock On graph problems in a semi-streaming model.
+\newblock {\em Theoretical Computer Science}, 2005.
+
+\bibitem{gedik2008spade}
+B.~Gedik, H.~Andrade, K.-L. Wu, P.~S. Yu, and M.~Doo.
+\newblock Spade: the system s declarative stream processing engine.
+\newblock {\em ACM SIGMOD}, 2008.
+
+\bibitem{blackBoxes}
+F.~Hueske, M.~Peters, M.~J. Sax, A.~Rheinl{\"a}nder, R.~Bergmann, A.~Krettek,
+  and K.~Tzoumas.
+\newblock {Opening the Black Boxes in Data Flow Optimization}.
+\newblock {\em PVLDB}, 2012.
+
+\bibitem{isard2007dryad}
+M.~Isard, M.~Budiu, Y.~Yu, A.~Birrell, and D.~Fetterly.
+\newblock Dryad: distributed data-parallel programs from sequential building
+  blocks.
+\newblock {\em ACM SIGOPS}, 2007.
+
+\bibitem{kornacker2015impala}
+M.~Kornacker, A.~Behm, V.~Bittorf, T.~Bobrovytsky, C.~Ching, A.~Choi,
+  J.~Erickson, M.~Grund, D.~Hecht, M.~Jacobs, et~al.
+\newblock Impala: A modern, open-source sql engine for hadoop.
+\newblock {\em CIDR}, 2015.
+
+\bibitem{li2008out}
+J.~Li, K.~Tufte, V.~Shkapenyuk, V.~Papadimos, T.~Johnson, and D.~Maier.
+\newblock Out-of-order processing: a new architecture for high-performance
+  stream systems.
+\newblock {\em PVLDB}, 2008.
+
+\bibitem{marz2015big}
+N.~Marz and J.~Warren.
+\newblock {\em Big Data: Principles and best practices of scalable realtime data systems}.
+\newblock Manning Publications Co., 2015.
+
+\bibitem{migliavacca2010seep}
+M.~Migliavacca, D.~Eyers, J.~Bacon, Y.~Papagiannis, B.~Shand, and P.~Pietzuch.
+\newblock Seep: scalable and elastic event processing.
+\newblock {\em Middleware'10 Posters and Demos Track}, page~4. ACM, 2010.
+
+\bibitem{murray2013naiad}
+D.~G. Murray, F.~McSherry, R.~Isaacs, M.~Isard, P.~Barham, and M.~Abadi.
+\newblock Naiad: a timely dataflow system.
+\newblock {\em ACM SOSP}, 2013.
+
+\bibitem{saha2015apache}
+B.~Saha, H.~Shah, S.~Seth, G.~Vijayaraghavan, A.~Murthy, and C.~Curino.
+\newblock Apache tez: A unifying framework for modeling and building data
+  processing applications.
+\newblock {\em ACM SIGMOD}, 2015.
+
+\bibitem{DBLP:conf/hotcloud/ZahariaCFSS10}
+M.~Zaharia, M.~Chowdhury, M.~J. Franklin, S.~Shenker, and I.~Stoica.
+\newblock {Spark: Cluster Computing with Working Sets}.
+\newblock {\em {USENIX HotCloud}}, 2010.
+
+\bibitem{scopeOptimizer}
+J.~Zhou, P.-A. Larson, and R.~Chaiken.
+\newblock Incorporating partitioning and parallel plans into the scope
+  optimizer.
+\newblock {\em IEEE ICDE}, 2010.
+
+\end{small}
+\end{thebibliography}
+
 % {
@@
 % }
-\printbibliography
+% \printbibliography
 

### 2016-06-02 paper.tex
Final

@@
 \begin{abstract}
-Apache Flink \footnote{The authors of this paper make no claim in being the sole inventors or implementers of the ideas behind Apache Flink, but rather a group of people that attempt to accurately document these concepts and their significance.
+Apache Flink\footnote{The authors of this paper make no claim in being the sole inventors or implementers of the ideas behind Apache Flink, but rather a group of people that attempt to accurately document Flink's concepts and their significance.
 Consult \autoref{sec:acknowledgements} for acknowledgements.} is an open-source system for processing streaming and batch data.
@@
 \begin{itemize}
-	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
+	\item we make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets,
 	\vspace{-3mm}
-	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault-tolerant streaming dataflows (in \autoref{sec:execution}).
+	\item we show how streaming, batch, iterative, and interactive analytics can be represented as fault-tolerant streaming dataflows (in \autoref{sec:execution}),
 	\vspace{-3mm}
-	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (in \autoref{sec:streaming}), as well as a full-fledged batch processor (in \autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
+	\item we discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (in \autoref{sec:streaming}), as well as a full-fledged batch processor (in \autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
 \end{itemize}
@@
 \begin{itemize}
-\item \textit{Checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (discussed in \autoref{sec:fault-tolerance}). \vspace{-3mm}
-\item \textit{Watermarks} signaling the progress of event-time within a stream partition (discussed in \autoref{sec:streaming-time}). \vspace{-3mm}
-\item \textit{Iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (discussed in \autoref{sec:batch-iterations}).
+\item \textit{checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (discussed in \autoref{sec:fault-tolerance}), \vspace{-3mm}
+\item \textit{watermarks} signaling the progress of event-time within a stream partition (discussed in \autoref{sec:streaming-time}), \vspace{-3mm}
+\item \textit{iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (discussed in \autoref{sec:batch-iterations}).
 \vspace{-1mm}
@@
 In addition, to comply with fault-tolerance guarantees, feedback streams are treated as operator state within the implicit-iteration head operator and are part of a global snapshot \cite{carbone2015lightweight}.
-The DataStream API allows for an explicit definition of feedback streams and can trivially subsume support for structured loops over streams\cite{murray2013naiad} as well as progress tracking\cite{chandramouli2009fly}.
+The DataStream API allows for an explicit definition of feedback streams and can trivially subsume support for structured loops over streams \cite{murray2013naiad} as well as progress tracking \cite{chandramouli2009fly}.
 
@@
 \newblock Seep: scalable and elastic event processing.
-\newblock {\em Middleware'10 Posters and Demos Track}, page~4. ACM, 2010.
+\newblock {\em ACM Middleware'10 Posters and Demos Track}, 2010.
 
